Cell autonomous role of leucine-rich repeat kinase in protection of dopaminergic neuron survival

Version 1

Abstract

Abstract

Mutations in leucine-rich repeat kinase 2 (LRRK2) are the most common genetic cause of Parkinson’s disease (PD), which is the leading neurodegenerative movement disorder characterized by the progressive loss of dopaminergic (DA) neurons in the substantia nigra pars compacta (SNpc). However, whether LRRK2 mutations cause PD and degeneration of DA neurons via a toxic gain-of-function or a loss-of-function mechanism is unresolved and has pivotal implications in LRRK2 based PD therapy. In this study, we investigate whether LRRK2 and its functional homologue LRRK1 play an essential, intrinsic role in DA neuron survival through the development of DA neuron-specific LRRK conditional double knockout (cDKO) mice. We first generated and characterized floxed LRRK1 and LRRK2 mice and then confirmed that germline deletions of the floxed LRRK1 and LRRK2 alleles result in null alleles, as evidenced by the absence of LRRK1 and LRRK2 mRNA and protein in the respective homozygous deleted mutant mice. We further examined the specificity of Cre-mediated recombination driven by the dopamine transporter-Cre (DAT-Cre) knockin (KI) allele using a GFP reporter line and confirmed that DAT-Cre-mediated recombination is restricted to DA neurons in the SNpc. Crossing these validated floxed LRRK1 and LRRK2 mice with DAT-Cre KI mice, we then generated DA neuron-restricted LRRK cDKO mice and further showed reduced levels of LRRK1 and LRRK2 in dissected ventral midbrains of LRRK cDKO mice. While DA neuron-restricted LRRK cDKO mice of both sexes exhibit normal mortality and body weight, they develop age-dependent loss of DA neurons in the SNpc, as demonstrated by the progressive reduction of DA neurons in the SNpc of cDKO mice at 20 and 24 months of age. Moreover, DA neurodegeneration is accompanied with increases of apoptosis and elevated microgliosis in the SNpc of LRRK cDKO mice. These findings provide unequivocal evidence for the importance of LRRK in DA neurons and raise the possibility that LRRK2 mutations may impair its protection of DA neurons, leading to the loss of DA neurons in Parkinson’s disease.

Introduction

Parkinson’s disease (PD) is the most common movement disorder and is characterized by the progressive loss of dopaminergic (DA) neurons in the substantia nigra pars compacta (SNpc). Dominantly inherited missense mutations in the leucine-rich repeat kinase 2 (LRRK2) gene are the most common cause of both familial and sporadic PD, highlighting the importance of LRRK2 in PD pathogenesis (1–16). LRRK2 is a large protein of 2,527 amino acid residues containing several functional domains including several leucine-rich repeats (LRRs), a GTPase-like domain of Ras-of-complex (Roc), a C-terminal of Roc (COR) domain, and a serine/threonine MAPKKK-like kinase domain. LRRK1, a homologue of LRRK2, belongs to the evolutionarily conserved Roco protein family and contains similar LRRs, Roc, COR, and kinase domains (17–19). LRRK proteins are broadly expressed with LRRK2 being most abundant in the kidney (20). While most PD mutations are found in LRRK2, rare variants in the Roc, COR, and kinase domains of LRRK1 have been reported and may be associated with PD (21). Furthermore, the Roc-COR domain of LRRK2 forms dimers and exhibits conventional Ras-like GTPase properties, and the R1441/C/G/H and I1371V mutations destabilize dimer formation and decrease GTPase activity (22, 23). Recent high-resolution cryoEM structural studies of full-length LRRK2 demonstrated its existence as dimers and pathogenic mutations such as R1441/C/G/H and Y1699I at the Roc-COR interface, whereas the G2910S mutant and wild-type LRRK2 share similar structure (24).

Previous genetic studies demonstrated that LRRK2 plays essential roles in the autophagy-lysosomal pathway (25–28). Consistent with abundant expression of LRRK2 in kidneys, LRRK2-/-mice develop age-dependent phenotypes in the kidney, including autophagy-lysosomal impairments and increases of α-synuclein, apoptosis, and inflammatory responses (25, 26, 29). It has also been reported that LRRK2 is important for maintaining lung homeostasis, and LRRK2 deficiency results in impaired autophagy in alveolar type II epithelial cells (28). It was proposed that the lack of brain phenotypes in LRRK2-/-mice might be due to the presence of LRRK1 and potential functional compensation mechanisms (25, 26). Indeed, LRRK double knockout (DKO) mice develop age-dependent, progressive loss of DA neurons in the SNpc, beginning at 14 months of age (30). However, LRRK DKO mice also exhibit lower body weight and increased mortality, raising the possibility that DA neurodegeneration in LRRK DKO mice may be secondary to poor health.

In this study, we investigate whether LRRK2 and its functional homologue LRRK1 play an essential, intrinsic role in DA neuron survival through the development of DA neuron-specific LRRK conditional DKO (cDKO) mice. We first generated and confirmed floxed LRRK1 and LRRK2 mice, and then crossed them with the deleter CMV-Cre to create germline deletions of the floxed LRRK1 and LRRK2 alleles, followed by Northern and Western analyses to confirm the absence of LRRK1 and LRRK2 mRNA and protein in the respective homozygous deleted mutant mice. We also crossed a GFP reporter mouse with DAT-Cre KI mice and found that Cre-mediated recombination occurs in most, if not all, DA neurons and is restricted to DA neurons in the SNpc. We then crossed these thoroughly validated floxed LRRK1 and LRRK2 mice with DAT-Cre KI mice to generate DA neuron-restricted LRRK cDKO mice and further confirmed the reduction of LRRK1 and LRRK2 in dissected ventral midbrains of LRRK cDKO mice. While DA neuron-restricted LRRK cDKO mice of both sexes exhibit normal mortality and body weight, they develop age-dependent loss of DA neurons in the SNpc, as demonstrated by the progressive reduction of TH+ DA neurons or NeuN+ neurons in the SNpc of cDKO mice at 20 and 24 months of age. Interestingly, quantitative electron microscopy analysis showed a similar number of electron-dense vacuoles in SNpc neurons of LRRK cDKO mice relative to controls, in contrast to age-dependent, dramatic increases of vacuoles in SNpc neurons of germline LRRK DKO mice (30, 31). Moreover, DA neurodegeneration is accompanied with increases of apoptosis and elevated microgliosis in the SNpc of LRRK cDKO mice. These findings provide unequivocal evidence for the importance of LRRK in DA neurons and raise the possibility that LRRK2 mutations may impair this crucial physiological function, leading to the loss of DA neurons in Parkinson’s disease.

Results

Generation and molecular characterization of the floxed and deleted LRRK1 and LRRK2 alleles

LRRK2 and its homologue LRRK1 share several functional (LRRs, GTPase Roc, COR, and kinase) domains (Figure 1A). To investigate the intrinsic role of LRRK in DA neurons, we generated floxed LRRK1 (fLRRK1) and floxed LRRK2 (fLRRK2) mice, which permit deletions of LRRK1 and LRRK2 selectively in DA neurons by DAT-Cre, which expresses Cre recombinase under the control of the endogenous DAT promoter (32). We introduced two loxP sites in introns 26 and 29 of LRRK1 through homologous recombination and site-specific recombination by FLP recombinase to remove the positive selection PGK-neo cassette, which is flanked by two FRT sites (Figure 1B, C; Figure 1-figure supplement 1, 2). The embryonic stem (ES) cells carrying the targeted allele or the floxed allele were identified and validated by Southern analysis using the 5’ and 3’ external probes as well as the neo probe (Figure 1D; Figure 1-figure supplement 3). The validated ES cells carrying the floxed LRRK1 allele were injected into mouse blastocysts to develop floxed LRRK1 mice, which were further confirmed by Southern using the 5’ and 3’ external probes (Figure 1D). In the presence of Cre recombinase, the floxed LRRK1 region containing part of intron 26 (1,288 bp), exons 27-29, which encode the kinase domain, and part of intron 29 (1,023 bp) is deleted (Figure 1C), and the removal of exons 27-29 (625 bp) results in a frameshift of downstream codons.

Generation and characterization of floxed and deleted LRRK1 and LRRK2 alleles

(A) Schematic illustrations of human LRRK1 and LRRK2 proteins showing similar functional domains. LRRK12015 protein is derived from exons 2-34 (Ensembl Genome Database: ENSG00000154237). LRRK22527 protein is derived from exons 1-51 (ENSG00000188906). LRR: Leucine rich repeats; Roc: Ras-of-complex; COR: C-terminal of Roc; KIN: Kinase domain.

(B) Schematic illustrations of the gene structures of mouse LRRK1 and LRRK2. The boxes in blue are exons that encode the LRRK1 and LRRK2 proteins, and the gray boxes represent the 5’ and 3’ UTRs. The exons are not drawn in scale. The start codon ATG is in exon 2 of LRRK1 and exon 1 of LRRK2. The exons 27-29 of LRRK1 and the promoter/exons 1-2 of LRRK2 are flanked with loxP sites (green arrowheads).

(C) Targeting strategy for the generation of the targeted, floxed, and deleted LRRK1 alleles. The red boxes represent the targeted exons 27-29, and the blue boxes represent the untargeted exons. The locations and sizes of the 5’ and 3’ external probes are shown. The targeting vector contains the 5’ and 3’ homologous regions (marked by dashed lines) and the middle-targeted region, which includes a loxP site (green arrowhead) in intron 26 (1,288 bp upstream of exon 27) and the PGK-neo selection cassette flanked by two FRT (FLP recognition target) sequences (gray circles) followed by another loxP site in intron 29 (1,023 bp downstream of exon 29). A negative selection cassette encoding diphtheria toxin fragment A (DTA) is also included in the targeting vector to reduce ES cells bearing randomly inserted targeting vectors. ES cells carrying the correctly targeted LRRK1 allele were transfected with pCAG-FLP to remove the PGK-neo cassette and generate the floxed LRRK1 allele. Floxed LRRK1 mice were bred with CMV-Cre transgenic mice to generate germline deleted LRRK1 mice. Detailed strategy for generating targeting vector and DNA sequence of floxed LRRK1 allele can be found in Figure 1-figure supplement 1 and 2, respectively.

(D) Southern analysis of the targeted and floxed LRRK1 alleles. Genomic DNA from ES cells or mouse tails was digested with HindIII and hybridized with the 5’ or 3’ external probe. For the 5’ probe, the resulting 17.0 kb and 4.8 kb bands represent the wild-type (WT) and the targeted (T) or floxed (F) alleles, respectively. For the 3’ probe, the resulting 17.0 kb and 12.2 kb bands represent the WT and the floxed (F) alleles, respectively. Detailed Southern strategy can be found in Figure 1-figure supplement 3.

(E) Targeting strategy for the generation of the targeted, floxed, and deleted LRRK2 alleles. The red boxes represent LRRK2 exons 1 and 2, and the start codon ATG resides in exon 1. The locations and sizes of the 5’ and 3’ external probes are shown. The targeting vector contains the 5’ and 3’ homologous regions (marked by dashed lines) and the middle targeted region (from the promoter region to intron 2), which includes a loxP site (green arrowhead) upstream (1,768 bp) of the transcription initiation site and the PGK-neo selection cassette flanked by two FRT sequences (gray circles) and two loxP sites (green arrowheads) in intron 2 (878 bp downstream of exon 2). A negative selection cassette encoding DTA is also included in the targeting vector. Mice carrying the correctly targeted LRRK2 allele were crossed with Actin-FLP deleter mice to generate floxed LRRK2 mice, which were bred with CMV-Cre transgenic mice to generate germline deleted LRRK2 mice. Detailed strategy for generating targeting vector and DNA sequence of floxed LRRK2 allele can be found in Figure 1-figure supplement 4 and 5, respectively.

(F) Southern analysis of the targeted and floxed LRRK2 alleles. Genomic DNA from ES cells or mouse tails was digested with NheI and hybridized with the 5’ or 3’ external probe. For the 5’ probe, the resulting 11.5 kb and 3.6 kb bands represent the wild-type (WT) and the targeted (T) or floxed (F) alleles, respectively. For the 3’ probe, the resulting 11.5 kb and 5.2 kb bands represent the WT and the floxed (F) alleles, respectively. Detailed Southern strategy can be found in Figure 1-figure supplement 6.

(G) Northern analysis of poly(A)+ RNA prepared from the lung of mice carrying homozygous germline deleted (Δ/Δ) LRRK1 alleles derived from the floxed LRRK1 alleles using the cDNA probe of exons 2-3 (left) and exons 27-29 (right). Using the upstream exons 2-3 probe, the LRRK1 transcripts in wild-type mice are the expected size of ∼7.4 kb, whereas the detected LRRK1 transcripts in LRRK1 Δ/Δ mice are truncated, consistent with the deletion of exons 27-29 (625 bp), which results in a frameshift and likely degradation of the truncated LRRK1 mRNA. Using a probe specific for exons 27-29, there is no LRRK1 transcript in LRRK1 Δ/Δ mice, as expected. Both blots were hybridized with a GAPDH probe as loading controls. Detailed Northern strategy and full-size blots are included in Figure 1-figure supplement 7 and 8, respectively. See also Figure 1-figure supplement 9.

(H) Northern analysis of total RNA prepared from the neocortex of mice carrying homozygous germline LRRK2 deleted (Δ/Δ) alleles using the cDNA probe of exons 1-5 shows the absence of LRRK2 transcripts. The blot was hybridized with a GAPDH probe as a loading control. The full-size blot is included in Figure 1-figure supplement 10. See also Figure 1-figure supplement 11

(I) Left: Western analysis of wild-type (+/+) and homozygous LRRK1 Δ/Δ brains show the absence of LRRK1 protein. Right: Western analysis of the neocortex of wild-type (+/+) and homozygous LRRK2 Δ/Δ mice shows the absence of LRRK2 protein. Vinculin was used as a loading control. Full-size blots are included in Figure 1-Source Data 1.

The LRRK2 targeting vector contains the 5’ homologous region, a loxP site 1,768 bp upstream of the transcription initiation site, exons 1-2, the PGK-neo cassette flanked by two loxP sites and two FRT sequences, and the 3’ homologous region (Figure 1B, E; Figure 1-figure supplement 4, 5). The ES cell clones carrying the targeted allele were identified and validated by Southern analysis using the 5’ external probe (Figure 1F; Figure 1-figure supplement 6) and genomic PCR followed by sequencing. Mice carrying the LRRK2 targeted allele were further verified by Southern analysis using the 5’ and 3’ external probes (Figure 1-figure supplement 6) as well as the neo probe and were then crossed with the Actin-FLP mice (33) to remove the PGK-neo cassette flanked by two FRT sites to generate floxed LRRK2 mice (Figure 1E). The resulting floxed LRRK2 mice were confirmed by Southern analysis using the 5’ and 3’ external probes (Figure 1F). In the presence of Cre recombinase, the floxed LRRK2 region containing the promoter and exons 1-2 are deleted, likely resulting in a null allele, as we previously targeted a very similar region (∼2.5 kb upstream of the transcription initiation site and exons 1-2) to generate germline deletion of LRRK2 (25).

Dopaminergic neurons in the SNpc are a small neuronal population embedded in the ventral midbrain, making it difficult to confirm whether DA neuron-specific deletions of the floxed LRRK1 and LRRK2 regions result in null alleles. We previously generated three independent LRRK1 knockout (KO) mice, and only one KO line (Line 2) represents a LRRK1 null allele (30), whereas deletion of LRRK2 promoter region and exons 1-2 resulted in a LRRK2 null allele (25). We therefore generated germline deleted LRRK1 and LRRK2 mice from floxed LRRK1 and LRRK2 mice, respectively, by crossing them to germline deleter CMV-Cre transgenic mice (34).

We then performed Northern analysis of LRRK1 using both an upstream probe specific for exons 2-3 and a downstream probe specific for exons 27-29 (Figure 1G; Figure 1-figure supplement 7, 8). Because of the low expression level of LRRK1 and the relative abundance of LRRK1 in the lung (20), we enriched polyA+ RNA from the lung of the mice carrying homozygous germline deleted (Δ/Δ) LRRK1 alleles derived from the floxed LRRK1 alleles. Using a probe specific for exons 2-3, the LRRK1 transcripts in wild-type mice are the expected size of ∼7.4 kb, whereas the detected LRRK1 transcripts in LRRK1 Δ/Δ mice are smaller, consistent with the deletion of exons 27-29 (625 bp), resulting in a frameshift of downstream codons and the likely degradation of the truncated LRRK1 mRNA (Figure. 1G; Figure 1-figure supplement 8). Using a probe specific for exons 27-29, there is no LRRK1 transcript in LRRK1 Δ/Δ mice (Figure. 1G; Figure 1-figure supplement 8). Extensive RT-PCR analysis of total RNA isolated from the kidney, brain, and lung of LRRK1 Δ/Δ mice using an exon 32-specific primer for RT and exon-specific primer sets for PCR (e.g. exons 4-8, 11-17, 20-25, and 25-31), followed by sequencing confirmation of the PCR products, indicated normal LRRK1 splicing in fLRRK1/fLRRK1 mice and the lack of LRRK1 exons 27-29 in LRRK1 Δ/Δ mice (Figure 1-figure supplement 9).

Similarly, Northern analysis of LRRK2 using a probe specific for exons 1-5 and RT-PCR followed by sequencing confirmed the absence of LRRK2 mRNA in LRRK2 Δ/Δ brains and normal LRRK2 transcripts in fLRRK2/fLRRK2 brains (Figure 1H; Figure 1-figure supplement 10, 11). Furthermore, Western analysis confirmed the absence of LRRK1 and LRRK2 proteins in the brain of LRRK1 Δ/Δ and LRRK2 Δ/Δ mice, respectively (Fig. 1I). Taken together, our Northern, RT-PCR followed by sequencing, and Western analyses demonstrated that deletion of the floxed LRRK1 and LRRK2 alleles results in null mutations. Thus, floxed LRRK1 and LRRK2 alleles can be used to generate DA neuron-specific LRRK cDKO mice.

Generation and molecular characterization of DA neuron-specific LRRK cDKO mice

To generate DA neuron-specific LRRK cDKO mice, we used DAT-Cre KI mice, which express Cre recombinase under the control of the endogenous DAT promoter (32). To confirm if Cre-mediated recombination occurs broadly and specifically in DA neurons of the SNpc, we crossed DAT-Cre mice with the Rosa26-CAG-LSL-ZsGreen1 reporter mouse (35). Upon Cre expression, Cre recombinase removes the floxed “stop” cassette, resulting in the expression of EGFP. We found that Cre-mediated recombination (GFP+) occurs in TH+ cells in the SNpc (Figure 2A). Quantification of GFP+ and/or TH+ cells in the SNpc showed that 99% of TH+ dopaminergic neurons are also GFP+, demonstrating that Cre-mediated recombination takes place in essentially all DA neurons in the SNpc (Figure 2B).

Generation and characterization of DA neuron-specific LRRK cDKO mice

(A) Immunostaining of GFP and/or TH in the SNpc of DAT-Cre/+; Rosa26-CAG-LSL-ZsGreen1/+ mice at 2 months of age. Cre recombinase is expressed under the control of the DAT endogenous promoter and removes the floxed “stop” cassette, resulting in the expression of EGFP under the control of the ubiquitous CAG promoter.

(B) Quantification of GFP+/TH+ and TH+ cells shows that 99% of TH+ DA neurons (722 ± 46 TH+ cells per brain) in the SNpc are also GFP+ (713 ± 46 cells per brain), indicating that DAT-Cre mediated recombination occurs in essentially all TH+ DA neurons. N = 3 mice, 3 comparable sections per hemisphere, 320 μm apart. Raw quantification data are included in Figure 2-Source Data 1.

(C) Similar body weight between LRRK cDKO mice and littermate controls at all ages examined (F1,73 = 0.0210, p = 0.8853; 2M, 20M, 25M: p > 0.9999; 15M: p = 0.7468, two-way ANOVA with Bonferroni’s post hoc multiple comparisons). Raw quantification data are included in Figure 2-Source Data 2.

(D) Similar brain weight between LRRK cDKO and control mice (F1,61 = 0.0440, p = 0.8345; 2M: p = 0.3893; 15M: p > 0.9999; 20M: p > 0.9999; 25M: p = 0.3223, two-way ANOVA with Bonferroni’s post hoc multiple comparisons). Raw quantification data are included in Figure 2-Source Data 2.

(E) Western analysis of LRRK1 and LRRK2 proteins in the dissected cerebral cortex (CTX) and ventral midbrain (vMB) of LRRK cDKO and littermate controls at 2 months of age. Full-size blots are included in Figure 2-Source Data 3.

(F) Quantification shows significant decreases of LRRK1 and LRRK2 in the dissected ventral midbrain of LRRK cDKO mice (LRRK1, p = 0.0432; LRRK2, p = 0.0162, unpaired two-tailed Student’s t-test), compared to controls, but not in the dissected cortex of cDKO mice (LRRK1, p = 0.7648; LRRK2, p = 0.2325). Raw quantification data are included in Figure 2-Source Data 4.

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. *p < 0.05. Scale bar: 100 μm.

Having confirmed fLRRK1 and fLRRK2 mice as well as DAT-Cre mice, we then bred them together to generate LRRK cDKO mice (fLRRK1/fLRRK1; fLRRK2/fLRRK2; DAT-Cre/+), which were further bred with fLRRK1/fLRRK1; fLRRK2/fLRRK2 mice to generate cDKO and littermate controls (fLRRK1/fLRRK1; fLRRK2/fLRRK2). It was previously reported that germline LRRK DKO mice failed to gain weight as they aged (30). However, DA neuron-specific LRRK cDKO and littermate control mice have similar body and brain weights at the ages of 2-24 months (Figure 2C, D). Western analysis showed a significant reduction of LRRK1 and LRRK2 proteins in the dissected ventral midbrain but not in the cerebral cortex of DA neuron-specific LRRK cDKO, relative to littermate controls at 2-3 months of age (Figure 2E, F).

Age-dependent loss of DA neurons in the SNpc of LRRK cDKO mice

To determine whether the inactivation of LRRK selectively in DA neurons of the SNpc impacts their survival, we performed TH immunostaining and quantified TH+ DA neurons in the SNpc of LRRK cDKO mice and littermate controls. The morphology of TH+ DA neurons in LRRK cDKO mice at the ages of 15, 20, and 24 months appears normal (Figure 3A). TH+ neuron counting revealed that the number of DA neurons in the SNpc at the age of 15 months is similar between cDKO mice (10,000 ± 141) and littermate controls (10,077 ± 310, F1,46 = 16.59, p = 0.0002; p > 0.9999, two-way ANOVA with Bonferroni’s post hoc multiple comparisons; Figure 3B). However, at the age of 20 months, the number of TH+ neurons in the SNpc of cDKO mice (8,948 ± 273) is significantly reduced, compared to controls (10,244 ± 220; p = 0.0041), and is further decreased at the age of 24 months (Control: 9,675 ± 232, cDKO: 8,188 ± 452; p = 0.0010).

Age-dependent loss of DA neurons in the SNpc of LRRK cDKO mice

(A) TH immunostaining shows TH+ DA neurons in the SNpc of LRRK cDKO and littermate controls at the age of 15, 20, and 24 months. Higher power views of the boxed areas show grossly normal DA neuron morphology in LRRK cDKO mice.

(B) Quantification of all TH+ DA neurons in the SNpc reveals similar numbers of DA neurons in LRRK cDKO mice (10,000 ± 141) and littermate controls (10,077 ± 310, p > 0.9999) at 15 months of age. At 20 months of age, the number of DA neurons in the SNpc of LRRK cDKO mice (8,948 ± 273) is significantly reduced, compared to control mice (10,244 ± 220, F1,46 = 16.59, p = 0.0002; p = 0.0041, two-way ANOVA with Bonferroni’s post hoc multiple comparisons). By 24 months of age, the reduction of DA neurons in the SNpc of LRRK cDKO mice (8,188 ± 452) relative to controls (9,675 ± 232, p = 0.0010) is greater, compared to that at 20 months of age. Raw quantification data are included in Figure 3-Source Data 1.

(C) Immunohistological analysis of TH and NeuN shows TH+ dopaminergic neurons (green) and NeuN+ neurons (red) in the SNpc of LRRK cDKO mice and controls at 24 months of age.

(D) Quantification of NeuN+ cells in the SNpc shows that the number of NeuN+ neurons in LRRK cDKO mice (17,923 ± 813) is significantly lower than that in control mice (21,907 ± 469, p = 0.0006, Student’s t-test), indicating loss of neurons in the SNpc of LRRK cDKO mice. All TH+ cells are NeuN+. The number of TH+/NeuN+ cells in the SNpc of LRRK cDKO mice (10,500 ± 644) is also lower, compared to control mice (14,102 ± 310, p = 0.0001). There is no significant difference in the number of NeuN+/TH-neurons between littermate controls (7,804 ± 249) and cDKO mice (7,423 ± 344, p = 0.3747). Raw quantification data are included in Figure 3-Source Data 2.

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. **p < 0.01, ***p < 0.001. Scale bar: 100 μm.

We further performed TH and NeuN double immunostaining of LRRK cDKO and control mice at 24 months of age (Figure 3C), and found that the number of NeuN+ neurons is also significantly reduced in the SNpc of cDKO mice (17,923 ± 813), compared to controls (21,907 ± 469, p = 0.0006, Student’s t-test; Figure 3D). The number of TH+/NeuN+ cells in the SNpc of LRRK cDKO mice (10,500 ± 644) is also lower, compared to control mice (14,102 ± 310, p = 0.0001; Figure 3D). These data indicate that the decreases in the number of TH+ cells are not due to decreases in TH expression in DA neurons, but rather a result of the loss of DA neuron cell bodies in the SNpc of LRRK cDKO mice.

We further evaluated apoptosis in the SNpc of LRRK cDKO and littermate controls at the age of 24 months using an antibody specific for active Caspase-3 to label apoptotic cells. We observed increases in apoptotic DA neurons, labeled by active Caspase-3+ and TH+ immunoreactivity, in the SNpc of LRRK cDKO mice (Figure 4A). Quantification of active Caspase-3+/TH+ apoptotic DA neurons shows a significant increase in the SNpc of LRRK cDKO mice (323 ± 38), compared to controls (157 ± 8, p = 0.0004, Student’s t-test; Figure 4B). These results further support that LRRK plays an intrinsic role in the survival of DA neurons in the SNpc during aging.

Increases in apoptotic DA neurons in the SNpc of LRRK cDKO mice

(A) Representative images of TH and active Caspase-3 immunostaining show TH+ DA neurons (green) and active Caspase-3+ apoptotic cells (red) in the SNpc of LRRK cDKO and control mice at the age of 24 months.

(B) Quantification of active Caspase-3+/TH+ cells shows significant increases of apoptotic DA neurons in the SNpc of LRRK cDKO mice (323 ± 38) at 24 months of age, relative to controls (157 ± 8, p = 0.0004, unpaired Student’s t-test). Raw quantification data are included in Figure 4-Source Data 1.

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. ***p < 0.001. Scale bar: 100 μm.

Previously, we reported the reduction of noradrenergic neurons in the locus coeruleus (LC) of LRRK DKO mice (30). We therefore quantified TH+ noradrenergic neurons in the LC of LRRK cDKO mice at the age of 24 months, and found a similar number of TH+ cells in the LC of LRRK cDKO and littermate controls (Control: 3,419 ± 86, cDKO: 3,350 ± 99, p = 0.6057, Student’s t-test; Figure 5A, B). Further examination of the GFP reporter line crossed with DAT-Cre showed the lack of Cre-mediated recombination in the LC (Figure 5C).

Normal number of TH+ noradrenergic neurons in the LC of LRRK cDKO mice

(A) Representative images of TH+ noradrenergic neurons in the locus coeruleus (LC) of LRRK cDKO mice and littermate controls at 24 months of age.

(B) Quantification of TH+ neurons shows similar numbers of TH+ noradrenergic neurons in the LC of LRRK cDKO mice (3,350 ± 99) and controls (3,419 ± 86, p = 0.6057, unpaired Student’s t-test). Raw quantification data are included in Figure 5-Source Data 1.

(C) Top: Immunostaining of TH and GFP in the LC of DAT-Cre/+; Rosa26-CAG-LSL-ZsGreen1/+ mice at 2 months of age. There is no GFP+ (green) cell in the LC, indicating that DAT-Cre is not expressed in the LC. Bottom: Higher power views of the boxed areas.

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. Scale bar: 100 μm.

Quantitative EM analysis of the SNpc in LRRK cDKO mice

We previously reported striking age-dependent increases of electron-dense autophagic and autolysosomal vacuoles as well as the presence of large lipofuscin granules in the surviving SNpc neurons of LRRK DKO mice beginning at 10 months of age (30, 31). To determine whether selective inactivation of LRRK in DA neurons similarly results in the accumulation of electron-dense autophagic and autolysosomal vacuoles in the SNpc, we performed EM analysis in the SNpc of LRRK cDKO mice and littermate controls at the ages of 25 months (Figure 6). We observed various electron-dense double membrane autophagosomes and single membrane autolysosomes as well as lipofuscin granules composed of lipid-containing residues of lysosomal digestion in the SNpc of LRRK cDKO and littermate control mice (Figure 6C-F). Interestingly, there are similar numbers of electron-dense vacuoles in the SNpc of LRRK cDKO mice and littermate controls at the age of 25 months (Control: 6.72 ± 0.43, cDKO: 6.99 ± 0.52, p = 0.6839, Student’s t-test; Figure 6G). We also found no significant difference in the area of electron-dense vacuoles in the SNpc between LRRK cDKO and littermate controls at the age of 25 months (Control: 4.43 ± 0.44 μm2; cDKO: 4.60 ± 0.49 μm2, p = 0.8048; Figure 6G). The difference in age-dependent accumulation of electron-dense vacuoles in the SNpc between germline DKO mice and DA neuron-restricted cDKO suggests that LRRK in non-DA neurons such as microglia, which is lacking in germline DKO but is expressed normally in cDKO mice, may play a more prominent role.

Unchanged number of electron-dense vacuoles in the SNpc of LRRK cDKO mice

(A, B) Representative EM images showing electron-dense vacuoles (arrowheads) in SNpc neurons of cDKO mice and littermate controls at the age of 25 months.

(C-F) Higher power views showing various electron-dense vacuoles, autolysosomes (single arrows), autophagosomes (double arrows), and lipid-containing vacuoles (asterisks) in SNpc neurons of littermate control (C, D) and cDKO (E, F) mice.

(G) Left: The average number of electron-dense vacuoles (>0.5 μm in diameter) in the SNpc neuronal profiles per mouse is not significantly different between LRRK cDKO mice and littermate controls at the age of 25 months (Control: 6.72 ± 0.43; cDKO: 6.99 ± 0.52, p = 0.6839, unpaired two-tailed Student’s t-test). Right: The total area of electron-dense vacuoles (>0.5 μm in diameter) in the SNpc neuronal profiles per mouse is similar between LRRK cDKO and littermate controls (Control: 4.43 ± 0.44 μm2; cDKO: 4.60 ± 0.49 μm2, p = 0.8048). Raw quantification data are included in Figure 6-Source Data 1.

The value in parentheses indicates the number of mice (left) and neuron profiles (right) used in the quantification. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. Scale bar: 1 μm.

Enhanced microgliosis in the SNpc of LRRK cDKO mice

We further evaluated whether elevated microgliosis accompanies DA neurodegeneration in the SNpc of LRRK cDKO mice. We performed immunohistochemical analysis of Iba1, which labels microglia, and TH, which marks DA neurons and processes, thus showing the boundary of the SNpc (Figure 7A-C). We found that the number of Iba1+ immunoreactive microglia is significantly increased in the SNpc of LRRK cDKO mice at 15 months of age (2541.3 ± 193.0), compared to controls (1736.7 ± 82.9; p = 0.0017, two-way ANOVA with Bonferroni’s post hoc multiple comparisons; Figure 7A, D). The number of Iba1+ immunoreactive microglia in the SNpc of LRRK cDKO is further increased compared to controls at the age of 20 (Control: 2426.0 ± 67.7, cDKO: 3639.0 ± 127.1, p < 0.0001; Figure 7B, D) and 25 months (Control: 2640.0 ± 187.1, cDKO: 4088.8 ± 100.3, p < 0.0001; Figure 7C, D). These results show that despite the selective inactivation of LRRK in DA neurons of LRRK cDKO mice, microgliosis accompanies DA neuronal loss in the SNpc.

Elevated microgliosis in the SNpc of LRRK cDKO mice

(A-C) Representative images of Iba1+ microglia (red, marked by yellow arrowheads) and TH+ dopaminergic neurons (green) in the SNpc of LRRK DKO mice and controls at 15 (A), 20 (B), and 24 (C) months of age.

(D) Quantification of Iba1+ microglia shows significant increases in the number of Iba1+ microglia in the SNpc of LRRK cDKO mice compared to control mice at 15 (control: 1736.7 ± 82.9, cDKO: 2541.3 ± 193.0, p = 0.0017, two-way ANOVA with Bonferroni’s post hoc multiple comparisons), 20 (control: 2426.0 ± 67.7, cDKO: 3639.0 ± 127.1, p < 0.0001), and 24 months of age (control: 2640.0 ± 187.1, cDKO: 4088.8 ± 100.3, p < 0.0001). Raw quantification data are included in Figure 7-Source Data 1.

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. ****p < 0.0001. Scale bar: 100 μm.

Discussion

LRRK2 mutations represent ∼5% familial and 1-2% of sporadic PD cases, and the G2019S mutation is the most common mutation but with a higher age of onset and incomplete penetrance (74% at age 79) (36). Despite the importance of LRRK2 in PD pathogenesis, whether LRRK2 mutations cause PD via a loss-of-function or a gain-of-function mechanism remains highly debated and unresolved. Neither LRRK2-deficient mice nor LRRK2 R1441C and G2019S KI mice develop dopaminergic neurodegeneration (25, 37, 38). It was proposed that the lack of brain phenotypes in LRRK2-/-mice might be due to the presence of LRRK1 (25, 26), which is broadly expressed in the brain including the midbrain (https://www.proteinatlas.org/ENSG00000154237-LRRK1/brain). Indeed, germline deletions of LRRK2 and LRRK1 result in age-dependent loss of DA neurons and increases of apoptosis and microgliosis in the SNpc of LRRK DKO mice (30, 31). However, the pleiotropic roles of LRRK1 and LRRK2 in the kidney (25, 26, 29), lung (28), and bone (39, 40), and the observed earlier mortality and lower body weight in LRRK DKO mice raised the possibility that DA neurodegeneration in LRRK DKO mice may be due to poor health.

In this study, we address this important question through the development of floxed LRRK1 and LRRK2 mice (Figure 1; Figure 1-figure supplement 1-11), which permit the dissection of cell intrinsic function of LRRK in DA neurons. We generated DA neuron-specific LRRK cDKO mice using the DAT-

Cre KI allele (32) to delete floxed LRRK1 and LRRK2 regions selectively in DA neurons. Indeed, using an EGFP reporter line, we found that DAT-Cre drives Cre-mediated recombination in almost all DA neurons in the SNpc (Figure 2). Unlike the LRRK germline DKO mice (30), DA neuron-specific LRRK cDKO mice exhibit normal body weight and mortality during mouse lifespan (Figure 2). Importantly, LRRK cDKO mice develop age-dependent DA neurodegeneration, as evidenced by progressive reductions of DA neurons in the SNpc at the age of 20-24 months (Figure 3) and increases of apoptotic DA neurons (Figure 4), whereas the number of noradrenergic neurons in the LC is uncaged, consistent with the lack of Cre-mediated recombination in the LC by DAT-Cre (Figure 5). Interestingly, quantitative EM analysis showed similar numbers of electron-dense vacuoles in the SNpc of LRRK cDKO and control mice even at 25 months of age (Figure 6), in contrast to dramatic, age-dependent increases of vacuoles in the SNpc of LRRK DKO mice (30, 31). Lastly, DA neurodegeneration in the SNpc of LRRK cDKO mice is accompanied with elevated microgliosis (Figure 7).

The toxic gain-of-function pathogenic mechanism was supported by earlier biochemical reports showing that LRRK2 mutations (e.g. R1441C/G, G2019S) enhance kinase activity, leading to elevated levels of pSer1292-LRRK2 (41), pSer935-LRRK2 (42, 43), pT73-Rab10, and pS106-Rab12 (44). However, most transgenic mice overexpressing mutant LRRK2 did not produce neurodegeneration or DA neuron loss (45–53), suggesting that enhanced kinase activity does not correlate with compromised DA neuronal survival, though two reports showed that overexpression of LRRK2 G2019S but not R1441C resulted in DA neurodegeneration (54, 55). Thus, further studies are needed to elucidate the differences of these seemingly conflicting transgenic findings and to understand the differences between the R1441C and G2019S mutations.

The loss-of-function pathogenic mechanism was supported by mouse knockout findings showing that germline inactivation of LRRK results in age-dependent loss of DA neurons in the SNpc and DA terminals in the striatum (30, 31). The current study demonstrated a cell intrinsic role of LRRK in support of DA neuron survival in the SNpc of the aging brain, providing further genetic evidence of an essential, cell autonomous requirement of LRRK in DA neuron survival. However, highly penetrant PD-linked mutations are dominantly-inherited missense mutations (6) in LRRK2, and there have been no reported deletion mutations that are linked to or causal of familial PD, despite various variants were reported in LRRK2 and LRRK1 that are associated with sporadic PD (but not linked to familial PD). One plausible explanation is that LRRK2 mutant may act in a dominant negative manner, inhibiting the activity of the wild-type LRRK2. This notion is supported by structural studies showing that the GTPase domain forms a dimer, and that the R1441/C/G/H and I1371V mutations destabilize dimer formation and decrease GTPase activity (22). LRRK2 may function as a dimer, which is further supported by the crystal structure of LRRK2’s WD40 domain dimer at 2.6 Å resolution (56). Additional studies also suggested that LRRK2 functions as a dimer (57–59), and that LRRK2 and LRRK1 can form a heterodimer (60). Furthermore, LRRK2 G2385R variant, a risk factor for PD (61–64), has been reported as a partial loss-of-function mutation (65, 66). Future studies are needed to differentiate the possibility between a toxic gain-of-function pathogenic mechanism vs. loss-of-function mutations exerting a dominant negative effect further impairing LRRK2 activity of the wild-type protein encoded by the LRRK2 wild-type allele, as the gain-of-function or loss-of-function pathogenic mechanism underlying LRRK2 mutations pivots opposing LRRK2 based therapeutic development of PD therapies.

Finally, it was surprising to find no accumulation of autophagic and autolysosomal vacuoles in SNpc neurons of LRRK cDKO mice, even at the age of 25 months. This suggests that non-dopaminergic cells (e.g. microglia), in which LRRK expression is unaffected in cDKO mice but is absent in germline DKO mice, may contribute to the dramatic phenotypic differences. Development of cell type-specific cDKO mice, such as microglia-specific LRRK cDKO mice, would permit further dissection of cell autonomous and non-cell autonomous role of LRRK in dopaminergic neuronal survival and whether glial LRRK plays a more important role in the regulation of the autophagy-lysosomal pathway.

Materials and methods

Key Resource Table

Mice

All animal use was approved by the IACUC committees of Harvard Medical School and Brigham and Women’s Hospital, and conformed to the USDA Animal Welfare Act, PHS Policy on Humane Care and Use of Laboratory Animals, the “ILAR Guide for the Care and Use of Laboratory Animals” and other applicable laws and regulations. Mice were housed in constant humidity- and temperature-controlled rooms and maintained on a 12 hr light/dark cycle and were given standard rodent chow and water. Mice of both sexes at multiple ages, from 2 months to 25 months, were used. Floxed LRRK1 (fLRRK1) and the resulting germline deleted LRRK1Δ alleles, and floxed LRRK2 (fLRRK2) and the resulting germline deleted LRRK2Δ alleles were generated and thoroughly validated at the genomic DNA, mRNA, and protein levels. DAT-IRES-Cre (The Jackson Laboratory, IMSR_JAX:006660), ACT-FLPe (IMSR_JAX:005703), CMV-Cre transgenic (IMSR_JAX:006054), and Rosa26-CAG-LSL-ZsGreen1 mice (IMSR_JAX:007906) used in the current study were previously characterized and reported (32–35). LRRK1/LRRK2 cDKO mice (fLRRK1/fLRRK1; fLRRK2/fLRRK2; DAT-IRES-Cre/+) and littermate controls (fLRRK1/fLRRK1; fLRRK2/fLRRK2) were obtained from multiple breeding cages of fLRRK1/fLRRK1; fLRRK2/fLRRK2 and cDKO mice. All mice were maintained on the C57BL6 and 129 hybrid genetic background (F1: IMSR_JAX:101043). All phenotypic analyses were performed in a genotype blind manner, as previously described (67).

Generation of targeted and floxed LRRK1 and LRRK2 alleles

The generation and validation of the LRRK1 and LRRK2 targeting vectors, the targeted, floxed, and deleted LRRK1 and LRRK2 alleles by Southern, Northern, RT-PCR, and the sequences of the floxed LRRK1 and LRRK2 alleles are included in Figure 1-figure supplement 1-11.

LRRK1

To generate the LRRK1 targeting vector, we first PCR amplified the left middle homologous region (2,079 bp) containing partial intron 26, exon 27, and partial intron 27 of LRRK1 from mouse BAC DNA (clone RP23-213J23, BACPAC Resources Center) using primers, P3 and P4. The PCR fragment was subcloned into the pGEM-T vector (A1360, Promega) to generate pLM1 (for details, see Figure 1-figure supplement 1). The right middle homologous region (3,403) containing LRRK1 genomic region from partial intron 27 to partial intron 29 was amplified by PCR using primers, P5 and P6, and then subcloned into the pGEM-T vector (pLM2). The left middle homologous region containing pGEM-T plasmid was then digested with NotI and KpnI (endogenous KpnI site in the intron 27) and subcloned into the NotI and KpnI sites of the right middle homologous region containing pGEM-T plasmid (pLM3).

After that, the middle homologous region (5.5 kb), from partial intron 26 to partial intron 29, was released by NotI and SacII digestions and was blunted by Klenow, and then subcloned into the SmaI site of PGKneoF2L2DTA vector (#13445, addgene) to generate the middle homologous region-PGKneoF2L2DTA plasmid (pLM4).

The left homologous region (2,016 bp) containing partial intron 25, exon 26, and partial intron 27 of LRRK1 was PCR amplified from mouse BAC DNA (clone RP23-213J23, BACPAC Resources Center) using primers, P1 and P2. The PCR fragment was digested with SacII (introduced by P1) and NotI (introduced by P2 along with HindIII), and was subcloned into the SacII and NotI sites of the PGKneoF2L2DTA vector (pLM5). The right homologous region (3,131 bp) containing partial intron 29, exon 30, and partial intron 30 of LRRK1 was PCR amplified from mouse BAC DNA (clone RP23-213J23) using primers, P7 and P8. The PCR fragment was digested with SalI (introduced by P7) and HindIII (introduced by P8), and was subcloned into the SalI and HindIII sites of the PGKneoF2L2DTA vector (pLM6).

The left homologous region in the PGKneoF2L2DAT vector (pLM5) was digested with SacII and NotI, and subcloned into the SacII and NotI sites of the right homologous region containing pGKneoF2L2DTA plasmid (pLM6) to generate pLM7. Finally, the loxP-middle homologous region-PGK-Neo-loxP fragment was released from pLM4 by NotI and SalI digestions, and subcloned into the NotI and SalI sites of pLM7 to generate the final target vector (pLM8), which contains two loxP sites (intron 26 and intron 29). Upon Cre-mediated recombination, the endogenous LRRK1 genomic sequence flanked by the 5’ loxP site (1,288 bp upstream of LRRK1 exon 27) to the 3’ loxP site (1,023 bp downstream of exon 29) is excised. The PGK-neo cassette flanked by two FRT sites is under the control of the mouse phosphoglycerate kinase 1 (PGK) promoter and contains the bovine growth hormone polyA signal. To enhance the ratio of ES cells carrying homologous recombination events instead of random insertion of the targeting vector (68), the negative selection PGK-DTA cassette, which encodes diphtheria toxin A chain, was also included in the LRRK1 targeting vector.

The LRRK1 targeting vector was linearized by XhoI digestion, and then electroporated into MKV6.5 embryonic stem (ES) cells (a gift from R. Jaenisch), which were derived from B6/129 F1 mice (The Jackson Laboratory, IMSR_JAX:101043). G418 was applied to the culture at 150 μg/ml 24 h later, and after 6 days of G418 selection, the surviving ES clones (480) were picked and screened by Southern analysis using HindIII digestion of genomic DNA followed by hybridization with the 5’ external and 3’ external probes to confirm proper recombination events in the 5’ and 3’ homologous regions, respectively.

25 ES cell clones were confirmed to carry the proper homologous recombination events at the 5’ homologous region, giving rise to the 17.0 kb and the 4.8 kb bands, which represent the wild-type and the targeted allele, respectively (Figure 1D), and the 3’ homologous region, giving rise to the 17.0 kb and the 14.1 kb bands, which represent the wild-type and the targeted allele, respectively. We then expanded the selected 4 ES cell clones (3A11, 3D8, 3H1, and 3H6) and further verified by Southern using the 5’ and 3’ external, and neo probes. Two ES cell clones (3D8 and 3H6) were transfected with pCAGGS-flpE-puro (#20733, addgene) to delete the PGK-neo cassette by FLP recombinase. Three resulting ES cell clones (3D8C5, 3D8E5, and 3H6G7) were confirmed by Southern analysis using the 5’ and 3’ external probes and the neo probe to confirm the floxed LRRK1 allele by the deletion of the PGK-neo cassette.

LRRK2

To generate the LRRK2 targeting vector, the left homologous region (2,579 bp) containing the LRRK2 promoter region was PCR amplified from mouse BAC DNA (clone RP23-526A2, BACPAC Resources Center) using primers, P9 and P10, and subcloned into the pGEM-T vector to generate pLRRK2#1 (for details see Figure 1-figure supplement 4). The BamHI-loxP-NheI-SpeI fragment (56 bp), which was generated by annealing two complementary oligos, P39 and P40, was introduced to the BamHI and SpeI sites of pLRRK2#1 to generate pLRRK2#2. The middle homologous region, containing genomic sequences from the promoter region to partial intron 2, was PCR amplified from mouse BAC DNA (clone RP23-526A2) using primers, P11 and P12, digested with XbaI and NotI, and subcloned into the XbaI and NotI sites of the pGEM-T vector to generate pLRRK2#3, which was digested with XbaI and NotI, and subcloned into the XbaI and NotI sites of pLRRK2#2 to generate pLRRK2#4. The right homologous region (3,503 bp), which contains partial intron 1, exon 2, and partial intron 2, was PCR amplified from mouse BAC DNA (clone RP23-526A2) using primers, P13 and P14, and the PCR fragment was subcloned into the EcoRV site of pBluescript II KS (+) vector (212207, Agilent) to generate pLRRK2#5. The right homologous region in pLRRK2#5 was digested with BamHI followed by Klenow, and then digested with ClaI. The released fragment was subcloned into the EcoRV and ClaI sites of the PGKneoF2L2DTA vector to generate pLRRK2#6. LFNT-tk/pBS plasmid (a gift from S. Tonegawa) was digested with SacII (followed by Klenow) and NotI (followed by SspI digestion to make it easier to isolate the NotI-SacII/KN fragment) to release the loxP-FRT-PKG-neo-loxP-FRT fragment (2,928 bp), and then subcloned into the XbaI (followed by Klenow) and NotI sites of pLRRK2#6 to generate pLRRK2#7. Finally, pLRRK2#4 containing the left-loxP-middle homologous regions was digested with EcoRV and NotI, and subcloned into AleI and NotI sites of pLRRK2#7 containing the loxP-FRT-PKG-neo-loxP-FRT-right homologous region to generate the final targeting vector, which contains the loxP sites in the promoter region (1,768 bp upstream of the transcription initiation site) and in LRRK2 intron 2 (878 bp downstream of exon 2). Upon Cre-mediated recombination, the floxed endogenous LRRK2 genomic sequences from the 5’ loxP site (1,768 bp upstream of LRRK2 transcription) to the 3’ loxP site in intron 2 (878 bp downstream of exon 2) are excised. The negative selection PGK-DTA cassette is also included in the LRRK2 targeting vector.

The LRRK2 targeting vector was linearized by AhdI digestion, and then electroporated into MKV6.5 ES cells. G418 was applied to the culture at 150 ug/ml 24 h later, and after 6 days of G418 selection, the surviving ES clones (480) were picked and screened by Southern analysis using NheI digestion of genomic DNA followed by hybridization with the 5’ external probe (753 bp, PCR amplified using P19 and P20). Southern analysis confirmed that two ES cell clones (N24 and N34) carry the proper recombination event in the 5’ homologous arm, giving rise to 11.6 kb and 3.6 kb bands, which represent the wild-type and targeted alleles, respectively (Figure 1F), followed by genomic PCR confirmation for the proper recombination in the right arm.

Three (3D8C5, 3A11, 3H6) and one (N24) ES clones for LRRK1 and LRRK2, respectively, were microinjected into C57BL/6 mouse blastocysts to generate chimera mice, which were bred with B6/129 F1 mice to produce heterozygous floxed LRRK1 and targeted LRRK2 mice. Floxed LRRK1 mice were confirmed by Southern analysis using the 5’ and 3’ external probes (Figure 1D). Targeted LRRK2 mice were confirmed by Southern (data shown in Figure 1-figure supplement 6; NheI digestion followed by hybridization with the 5’ external probe and SphI digestion followed by hybridization with the 3’ external probe). Targeted LRRK2 mice were then bred with Actin-FLP deleter mice (IMSR_JAX:005703) (33) to generate floxed LRRK2 mice, which were confirmed by Southern analysis using the 5’ and 3’ external probes following NheI digestion (Figure 1F). Heterozygous fLRRK1/+ and fLRRK2/+ mice were crossed with each other to obtain homozygous single floxed mice (fLRRK1/fLRRK1 and fLRRK2/fLRRK2) and double floxed mice (fLRRK1/fLRRK1; fLRRK2/fLRRK2).

Generation of deleted LRRK1/LRRK2 alleles and DA neuron-specific cDKO mice

In order to ensure that Cre-mediated deletion of the floxed LRRK1 and LRRK2 alleles results in null alleles, we crossed floxed LRRK1 and LRRK2 mice with germline deleter, CMV-Cre transgenic mice (B6.C-Tg(CMV-cre)1Cgn/J; IMSR_JAX:006054) (34), to generate LRRK1 deleted (Δ/Δ) mice (by removing exons 27-29) and LRRK2 deleted (Δ/Δ) mice (by removing the promoter region and exons 1-2) for further molecular characterization. To generate DA neuron-specific LRRK1/LRRK2 cDKO mice, we used DAT-Cre KI mice (B6.SJL-Slc6a3tm1.1(cre)Bkmn/J; IMSR_JAX:006660), which express Cre recombinase under the control of the endogenous DAT promoter (32). We crossed double floxed mice with DAT-Cre mice to generate LRRK cDKO mice (fLRRK1/fLRRK1; fLRRK2/fLRRK2; DAT-Cre/+). LRRK cDKO and littermate control mice used in the phenotypic analysis were obtained by crossing fLRRK1/fLRRK1; fLRRK2/fLRRK2; DAT-Cre/+ with fLRRK1/fLRRK1; fLRRK2/fLRRK2 mice. We only used cDKO and control mice that carry all floxed LRRK1 and LRRK2 alleles for phenotypic analysis.

Southern analysis

For the identification and validation of the targeted and floxed LRRK1 alleles, we used the 5’ external, 3’ external, and neo probes. Genomic DNA from ES cells or mouse tails was digested with HindIII. The 5’ external probe (377 bp), which is 839 bp upstream of the 5’ homologous region, was PCR amplified from mouse BAC DNA (clone RP23-213J23) using primers P15 and P16. The 3’ external probe (305 bp), which is 2,736 bp downstream of the 3’ homologous region, was PCR amplified from mouse BAC DNA (clone RP23-213J23) using primers P17 and P18. The neo probe (363 bp) was PCR amplified from pSoriano plasmid using primers P23 and P24. Following HindIII digestion, the presence of the 17.0 kb product using either the 5’ or 3’ external probe represents the wild-type allele, whereas the 4.8 kb (the 5’ external probe) and the 12.2 kb (the 3’ external probe) products represent the floxed LRRK1 allele. Genomic DNA digested by HindIII and hybridized with the neo probe further confirmed the wild-type and floxed alleles (no band) and the targeted LRRK1 allele (14.1 kb).

For the identification and validation of the targeted and floxed LRRK2 alleles, we used the 5’ and 3’ external probes as well as the neo probe. Genomic DNA from ES cells or mouse tails was digested with NheI or SphI followed by hybridization with the 5’ or 3’ external probe or the neo probe. The 5’ external probe (753 bp), which is 84 bp upstream of the 5’ homologous region, was PCR amplified from mouse BAC DNA (clone RP23-526A2) using primers P19 and P20. The 3’ external probe (622 bp), which is 38 bp downstream of the 3’ homologous region, was PCR amplified from mouse BAC DNA (clone RP23-526A2) using primers P21 and P22. Following NheI digestion, the presence of the 11.5 kb product using either the 5’ or 3’ external probe represents the wild-type allele, whereas the 3.6 kb (the 5’ external probe) and the 5.2 kb (the 3’ external probe) products represent the floxed LRRK2 allele.

PCR genotyping

Genomic PCR was performed to determine the presence of the deleted, the floxed, and/or the wild-type alleles. For LRRK1, the following primers were used: P25 (5’-ATTGGTCTTTGAAGAGACAGCATCTGG, forward primer, 392 nt downstream of exon 26), P26 (5’-TTTCCCTGAGGTGGAGAAGTGACTGG, reverse primer, 567 nt downstream of exon 26), and P27 (5’-TCACGTCGTCTAAGCCTCCT, reverse primer, 1,218 nt downstream of exon 29). The PCR products from P25 and P26 are 266 bp and 405 bp, which represent the wild-type and the fLRRK1 alleles, respectively, whereas the PCR product from P25 and P27 is 583 bp, which represents the deleted LRRK1 allele. For LRRK2, the following primers were used: P28 (5’-CTTCCTCAGAAGTTAGGTAAACATTGAGTG, forward primer, 2,069 nt upstream of exon 1), P29 (5’-CTAAGTGACACCGTGTTTCCAAAGTC, reverse primer, 1,739 nt upstream of exon 1), and P30 (5’-GGAAAGTTTCACAATTGGAAAAATAAAAATATTTACTGCAGATA, reverse primer at 2,848 nt downstream of exon 2). The PCR products from P28 and P29 are 305 bp and 367 bp, representing the wild-type and the fLRRK2 alleles, respectively, whereas the PCR product from P28 and P30 is 587 bp, which represents the deleted LRRK2 allele.

For DAT-IRES-Cre, the following primers were used: JKM1823 (5’-TGGCTGTTGGTGTAAAGTGG, forward primer at exon 16 and 3’-UTR), JKM1824 (5’-GGACAGGGACATGGTTGACT, reverse primer at 3’-UTR), and JKM1825 (5’-CCAAAAGACGGCAATATGGT, reverse primer at IRES sequence). The PCR product from JKM1823 and JKM1824 is 264 bp, which represents the wild-type allele, whereas the PCR product from JKM1823 and JKM1825 is 152 bp, which represents the DAT-IRES-Cre KI allele.

Northern analysis

Total RNA was isolated from brains, kidneys, or lungs using TRI reagent (T9424, Millipore-Sigma) according to the manufacturer’s instruction. For the LRRK1 Northern analysis, polyA+ RNA was enriched from ∼500 μg total RNA using the Poly(A)Purist™ MAG Kit (AM1922, ThermoFisher) according to the manufacturer’s instruction. For the LRRK2, ∼10 μg of total RNA was used for Northern analysis. RNA was separated in formaldehyde agarose gel, and transferred into Amersham Hybond-nylon membrane (RPN303N, GE Healthcare). Probes were synthesized using Prime-It II random labeling kit (#300385, Agilent) and then used for membrane hybridization at 55°C overnight.

The cDNA probe specific for LRRK1 exons 2-3 (383 bp) was PCR amplified using primers P31 (5’-CAGGATGAGCGTGTGTCTGCAG) and P32 (5’-CCTTCTCCTGTGAGGATTCGCTCT). The cDNA probe specific for LRRK1 exons 27-29 (550 bp) was PCR amplified using primers P33 (5’-CTGGCCTACCTGCACAAGAA) and P34 (5’-CCTTCCCATCCCAGAACACC).

The cDNA probe specific for LRRK2 exons 1-5 (437 bp) was PCR amplified using primers P35 (5’-AGGAAGGCAAGCAGATCGAG) and P36 (5’-GGCTGAATATCTGTGCATGGC). The probe specific for GAPDH exons 5-7 (452 bp) was PCR amplified using primers P37 (5’-ACCACAGTCCATGCCATCAC) and P38 (5’-TCCACCACCCTGTTGCTGTA).

Hybridization was performed using α-32P-dCTP-labeled probes specific to each gene. Specific signals were detected by autoradiography with Hyperfilm (E3018, Amersham).

RT-PCR

Total RNA was isolated from brains, kidneys, or lungs using TRI reagent (T9424, Millipore-Sigma) according to the manufacturer’s instruction. Approximately 1 μg of RNA was reverse-transcribed using Superscript III (18080093, Thermo Fisher Scientific) according to the manufacturer’s instructions. For RT-PCR analysis of LRRK1 transcripts in mice carrying the homozygous floxed or deleted alleles, we used primers P53 and P54 for exons 4-8 (714 bp), P57 and P58 for exons 11-17 (818 bp), P63 and P64 for exons 20-25 (922 bp), or P41 and P42 for exons 25-31 to confirm normal splicing of LRRK1 mRNA in fLRRK1/fLRRK1 mice and truncated LRRK1 transcripts lacking exons 27-29 in Δ/Δ mice (for details see Figure 1-figure supplement 9). For RT-PCR analysis of LRRK2 transcripts, we used primers P35 and P36 in exons 1-5 to confirm normal splicing of LRRK2 mRNA (437 bp) in fLRRK2/fLRRK2 mice and the absence of RT-PCR products in LRRK2 Δ/Δ mice (Figure 1-figure supplement 11). The identity of the PCR products was confirmed by sequencing.

Western analysis

Fresh tissues were collected and homogenized in an ice-cold stringent RIPA buffer (50 mM Tris-Cl (pH 7.6), 150 mM NaCl, 0.5 mM EDTA, 1% NP40, 0.5% sodium deoxycholate, 0.1% SDS, 1mM PMSF supplement with protease inhibitor cocktail (P8340, Sigma) and phosphatase inhibitor cocktail (P0044, Sigma)), followed by sonication. Homogenates were centrifuged at 14,000xg for 20 min at 4°C to separate supernatants (RIPA buffer-soluble fraction). An equal amount (10-40 μg per lane) of total proteins from each preparation were loaded and separated on NuPAGE gels (Invitrogen), and transferred to nitrocellulose membranes. The membranes were blocked in Intercept (TBS) Blocking Buffer (927-60001, Li-Cor) for 1 hr at room temperature, and incubated at 4°C overnight with specific primary antibodies. Primary antibodies used were rabbit anti-LRRK1 (ANR-101, Alomone Lab, RRID: AB_2756700), rabbit anti-LRRK2 (ab133474, abcam, RRID: AB_2713963), mouse anti-α-Vinculin (05-386, Millipore, RRID: AB_309711). Membranes were then incubated with dye-coupled secondary antibodies, goat anti-mouse IRdye680 (#925-68070, LI-COR, RRID: AB_2651128) or goat anti-rabbit IRdye800 (#925-32211, LI-COR, RRID: AB_2651127). Signals were quantified using the Odyssey Infrared Imaging System (LI-COR).

Histological analysis

Mice were anesthetized with ketamine (100 mg/kg) + xylazine (10 mg/kg) + acepromazine (3 mg/kg), and transcardially perfused with phosphate-buffered saline solution (PBS, pH 7.4) containing 0.25 g/L heparin (H3149, Sigma) and 5 g/L procaine (P9879, Sigma). Brains were post-fixed in 4% formaldehyde in PBS (pH 7.4) (15710, Electron Microscopy Sciences) at 4°C overnight and then processed for paraffin embedding following standard procedures. For frozen sections, post-fixed brains were immersed in a sucrose series solution (15% and 30% sucrose in PBS) at 4°C overnight for cryoprotection, and then brains were embedded in an OCT compound (4583, Sakura). Serial coronal sections (16 µm) of paraffinized brains or frozen brains were obtained using Leica RM2235 or Leica CM1860, respectively. Coronal sections containing the SNpc or LC were selected for immunohistochemical analysis.

Histological analyses were performed as described previously (30, 31, 67). Briefly, for DAB-derived TH-immunohistochemistry, coronal sections were deparaffinized, alcohol-dehydrated, and then subjected to permeabilization with a solution containing 0.1% Triton X-100 in TBS followed by antigen retrieval for 5 min in 10 mM sodium citrate buffer, pH 6.0. Endogenous peroxidase activity was quenched by incubating in 0.3% H2O2 in methanol. Sections were then blocked with a solution containing 5% normal goat serum (S-1000, Vector Laboratories) and 0.1% Triton X-100 in TBS for 1 hr at room temperature. After blocking, sections were incubated with the primary antibody, rabbit anti-TH (1:750, ab112, abcam, RRID: AB_297840) overnight at 4°C. Sections were washed three times in 0.1% Triton X-100 in TBS followed by 1 hr incubation with goat biotinylated anti-rabbit IgG secondary antibody (1:250, BA-1000, Vector Laboratories, RRID: AB_2313606) at room temperature and 30 min incubation with Vectastain Elite ABC reagent (PK-6100, Vector Laboratories) and then developed using chromogenic DAB substrate (SK-4100, Vector Laboratories).

For immunofluorescence staining of paraffin sections, coronal sections were deparaffinized, alcohol-dehydrated, and then subjected to permeabilization with a solution containing 0.1% Triton X-100 in PBS followed by antigen retrieval for 5 min in 10 mM sodium citrate buffer, pH 6.0 except those for cleaved-caspase3 immunostaining which were performed antigen retrieval for 10 min. Sections were then blocked with a solution containing 10% normal goat serum and 0.1% Triton X-100 in PBS for 1 hr at room temperature. After blocking, sections were incubated with primary antibodies overnight at 4°C. The primary antibodies used were mouse anti-TH (1:50, sc-25269, Santa Cruz, RRID: AB_628422), rabbit anti-NeuN (1:400, 12943S, Cell Signaling Technology, RRID: AB_2630395), rabbit anti-cleaved caspases-3 (1:150, 9661, Cell Signaling Technology, RRID: AB_2341188), or rabbit anti-Iba1 (1:500, 019-19,751, Wako, RRID: AB_839504). Sections were washed three times in 0.1% Triton X-100 in PBS followed by 1 hr incubation with fluorophore-conjugated secondary antibodies, goat anti-mouse IgG Alexa Fluor 488 (1:500, A11001, ThermoFisher, RRID: AB_2534069) and goat anti-rabbit IgG Alexa Fluor 555 (1:500, A32732, ThermoFisher, RRID: AB_2633281) at room temperature. Fluorescence images were taken (a stack of 2-3 confocal images spaced at 4 μm), projected with maximal intensity projection mode, and analyzed by FV1000 confocal microscope system (Olympus)

For immunofluorescence staining of cryopreserved sections, coronal brain sections were washed with PBS to rinse out OCT, and then blocked with a solution containing 5% normal goat serum and 0.1% Triton X-100 in TBS for 1 hr at room temperature. After blocking with 10% NGS, sections were incubated with primary antibodies, rabbit anti-GFP (1: 1000, ab290, abcam, RRID: AB_303395), and mouse anti-TH (1:50, sc-25269, Santa Cruz, RRID: AB_628422), overnight at 4°C. Sections were incubated with fluorophore-conjugated secondary antibodies, goat anti-rabbit IgG Alexa Fluor 488 (1:500, A11034, ThermoFisher, RRID: AB_2576217) and goat anti-mouse IgG Alexa Fluor 555 (1:250, A21424, ThermoFisher, RRID: AB_141780) for 1 hr at room temperature. Fluorescence images were taken and analyzed by FV1000 confocal microscope system (Olympus).

The number of GFP+ and GFP+/TH+ cells in the SNpc of DAT-IRES-Cre/+; Rosa26-CAG-LSL-ZsGreen1/+ reporter mice were quantified using 3 comparable coronal sections (16 µm in thickness, spaced 320 μm apart) per brain (n = 3 brains, one hemisphere). The percentage of GFP+ DA neurons in the SNpc was obtained by dividing the sum of GPF+/TH+ neurons by the sum of total TH+ neurons quantified.

Quantification of TH+ DA neurons in the SNpc or LC was performed as previously described (30, 31, 67, 69). Briefly, TH+ neurons in the SNpc, which were marked based on morphological features as previously described (70), were quantified in every 10th serial coronal section (16 µm in thickness) throughout the SNpc (a total of 6-9 sections, spaced 160 µm apart). Total number of TH+ cells in the SNpc was calculated as follows: [total number of TH+ DA neurons quantified in all 6-9 sections] x 10 (every 10th section sampled) x 2 (both hemispheres). Similar results were obtained independently by two investigators using coded samples. Total number of TH+ noradrenergic neurons in the LC (a total of 6-9 sections, spaced 80 µm apart) was calculated as follows: [total number of TH+ noradrenergic neurons quantified in all 6-9 sections] x 5 (every 5th section sampled) x 2 (both hemispheres).

NeuN+, active Caspase-3+, or Iba1+ cells in the SNpc, which was marked by TH immunoreactivity, were quantified using serial coronal sections (16 μm in thickness, every 10th section, a total of 6-9 sections per brain). Compared to chromogenic DAB immunostaining followed by counting under the stereomicroscope, which captures TH-positive dopaminergic neurons from the single plane layer of the sections, immunofluorescence immunostaining picks up more cells with higher sensitivity by maximal intensity projection mode throughout the entire layer of the section under the confocal microscope. Thus, more TH-positive dopaminergic neurons in the SNpc were captured and quantified under confocal microscopy compared to DAB staining. The total number of NeuN+, active Caspase-3+, or Iba1+ cells in the SNpc, which was marked by TH immunoreactivity, was calculated as follows: [total number of NeuN+, active Caspase-3+, or Iba1+ cells in all sections counted] x 10 (every 10th section sampled) x 2 (both hemispheres).

Quantitative EM analysis

The collection and quantification of the EM images were performed as described previously (30, 71). Mice were perfused with PBS containing 0.25 g/L heparin and 5 g/L procaine followed by a fixative solution containing 2.5% paraformaldehyde and 2.5% glutaraldehyde in 0.1 M sodium cacodylate buffer (pH 7.4) (#1549, Electron Microscopy Science). Brains were dissected and post-fixed overnight in a fixative solution at 4°C. The dissected tissues were trimmed to 1-2 mm3 cubes followed by osmication and uranyl acetate staining, dehydration in graded alcohol, and embedded in TABB 812 Resin (Marivac Ltd.) at the Harvard Medical School EM facility. 0.5 μm sections were stained with toluidine blue and viewed under the light stereomicroscope (Nikon Eclips E600) to find the SNpc area for EM viewing. Then adjacent sections were cut with 80 nm in thickness with the Leica Ultracut S microtome, picked up on formvar-carbon coated slot Copper grids, stained with 0.2% Lead Citrate, and viewed and imaged under the JEOL 1200x Electron Microscope. A minimum of 10 micrographs containing the entire cell body in the SNpc area were analyzed for each brain. The image was analyzed by using the Fiji version of ImageJ. The number of electron-dense autophagic and lysosomal vacuoles (> 0.5 μm in diameter) in individual neuronal profiles was quantified. We previously calculated the diameter of electron-dense autophagic/lysosomal vacuoles by measuring the longest side manually (30, 31). In the current study, we used Feret’s Diameter (72) to accurately measure the longest distance between any of two points in the electron-dense autophagic/lysosomal vacuoles, resulting in a higher number of electron-dense vacuoles quantified. The number of electron-dense vacuoles (> 0.5 μm in diameter) in individual neuronal profiles was quantified. The area of electron-dense vacuoles (> 0.5 μm in diameter) was quantified and averaged in individual neuronal profiles. Finally, the average number or area of vacuoles was calculated per genotype. Experiments were done in a genotype blind manner (after scarifying mice, the brain samples were coded and sent to the Harvard Medical School EM Core, where the images were captured, and the vacuoles were quantified by another independent experimenter).

Experimental design and statistical analysis

Group sizes were determined based on prior studies. Littermates of control and cDKO mice were randomly distributed between experimental groups. Data acquisition and quantification were performed in a genotype blind manner with the exception of molecular analyses (Southern, Northern, RT-PCR, and Western). All statistical analyses were performed using Prism 9 (GraphPad software) or Excel (Microsoft). All data are presented as the means ± SEM. The exact sample size, the number of mice or neurons, of each experiment is indicated in the figure or the legend.

Statistical analyses were conducted using a two-tailed Student’s t-test for the comparison of a given variable in two genotypes or two-way ANOVA followed by Bonferroni’s post hoc comparisons for the comparison of more than two conditions. Statistical outliers were identified and excluded using the ROUT method with 1% the maximum desired false discovery rate developed by Prism. All statistical comparisons were performed on data from ≥ 4 biologically independent samples and replicated on different experimental days. Significance is shown as *p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001, and not significantly different values are often not noted.

Data availability

All data generated or analyzed during this study are included in the manuscript and supporting file. Source Data files have been provided for Figures 1 to 7.

Funding disclosure

This work was supported by a grant from the NIH (R37NS071251 to J.S.).

eLife Assessment

This work describes new validated conditional double KO (cDKO) mice for LRRK1 and LRRK2 that will be useful for the field, given that LRRK2 is widely expressed in the brain and periphery, and many divergent phenotypes have been attributed previously to LRRK2 expression. The manuscript presents solid data demonstrating that it is the loss of LRRK1 and LRRK2 expression within the SNpc DA cells that is not well tolerated, as it was previously unclear from past work whether neurodegeneration in the LRRK double Knock Out (DKO) was cell autonomous or the result of loss of LRRK1/LRRK2 expression in other types of cells. Future studies may pursue the biochemical mechanisms underlying the reason for the apoptotic cells noted in this study, as here, the LRRK1/LRRK2 KO mice did not replicate the dramatic increase in the number of autophagic vacuoles previously noted in germline global LRRK1/LRRK2 KO mice.

Reviewer #1 (Public Review):

Summary:

This is an important work showing that loss of LRRK function causes late-onset dopaminergic neurodegeneration in a cell-autonomous manner. One of the LRRK members, LRRK2, is of significant translational importance as mutations in LRRK2 cause late-onset autosomal dominant Parkinson's disease (PD). While many in the field assume that LRRK2 mutant causes PD via increased LRRK2 activity (i.e., kinase activity), it is not a settled issue as not all disease-causing mutant LRRK2 exhibit increased activity. Further, while LRRK2 inhibitors are under clinical trials for PD, the consequence of chronic, long-term LRRK2 inhibition is unknown. Thus, studies evaluating the long-term impact of LRRK deficit have important translational implications. Moreover, because LRRK proteins, particularly LRRK2, are known to modulate immune response and intracellular membrane trafficking, the study's results and the reagents will be valuable for others interested in LRRK function.

Strengths:

This report describes a mouse model where the LRRK1 and LRRK2 gene is conditionally deleted in dopaminergic neurons. Previously, this group showed that while loss of LRRK2 expression does not cause brain phenotype, loss of both LRRK1 and LRRK2 causes a later onset, progressive degeneration of catecholaminergic neurons and dopaminergic (DAergic) neurons in the substantia nigra (SN), and noradrenergic neurons in the locus coeruleus (LC). However, because LRRK genes are widely expressed with some peripheral phenotypes, it was unknown if the neurodegeneration in the LRRK double knockout (DKO) was cell autonomous. To rigorously test this question, the authors have generated a double conditional (cDKO) allele where both LRRK1 and LRRK2 genes were targeted to contain loxP sites. In my view, this was beyond what is usually required, as most investigators might might combine one KO allele with another floxed allele. The authors provide a rigorous validation showing that the Driver (DAT-Cre) is expressed in most DAergic neurons in the SN and that LRRK levers are decreased selectively in the ventral midbrain. Using these mice, the authors show that the number of DAergic neurons is normal at 15 but significantly decreased at 20 months of age. Moreover, the authors show that the number of apoptotic neurons is increased by ~2X in aged SN, demonstrating increased ongoing cell death, as well as an increase in activated microglia. The degeneration is limited to DAergic neurons as LC neurons are not lost as this population does not express DAT. Overall, the mouse genetics and experimental analysis were performed rigorously, and the results were statistically sound and compelling.

Weaknesses:

I only have a few minor comments. First is that in PD and other degenerative conditions, loss of axons and terminals occurs prior to cell bodies. It might be beneficial to show the status of DAergic markers in the striatum. Second, previous studies indicate that very little, if any, LRRK1 is expressed in SN DAergic neurons. This also the case with the Allen Brain Atlas profile. Thus, authors should discuss the discrepancy as authors seem to imply significant LRRK1 expression in DA neurons.

Reviewer #2 (Public Review):

Summary:

In this manuscript, Shen and collaborators described the generation of cDKO mice lacking LRRK1 and LRRK2 selectively in DAT-positive DAergic neurons. The Authors asked whether selective deletion of both LRRK isoforms could lead to a Parkinsonian phenotype, as previously reported by the same group in germline double LRRK1 and LRRK2 knockout mice (PMID: 29056298). Indeed, cDKO mice developed a late reduction of TH+ neurons in SNpc that partially correlated with the reduction of NeuN+ cells. This was associated with increased apoptotic cell and microglial cell numbers in SNpc. Unlike the constitutive DKO mice described earlier, however, cDKO mice did not replicate the dramatic increase in the number of autophagic vacuoles. The study supports the authors' hypothesis that loss of function rather than gain of function of LRRK2 leads to PD.

Strengths:

The study described for the first time a model where both the PD-associated gene LRRK2 and its homolog LRRK1 are deleted selectively in DAergic neurons, offering a new tool to understand the physiopathological role of LRRK2 and the compensating role of LRRK1 in modulating DAergic cell function.

Weaknesses:

The model has no construct validity since loss of function mutations of LRRK2 are well-tolerated in humans and do not lead to PD. The evidence of a Parkinsonian phenotype in these cDKO mice is limited and should be considered preliminary.

Reviewer #3 (Public Review):

Kang, Huang, and colleagues investigated the impact of LRRK1 and LRRK2 deletion, specifically in dopaminergic neurons, using a novel cDKO mouse model. They observed a significant reduction in DAergic neurons in the substantia nigra in their conditional LRRK1 and LRRK2 KO mice and a corresponding increase in markers of apoptosis and gliosis. This work set out to address a long-standing question within the field around the role and importance of LRRK1 and LRRK2 in DAergic neurons and suggests that the loss of both proteins triggers some neurodegeneration and glial activation.

The studies included in this work are carefully performed and clearly communicated, but additional studies are needed to strengthen further the authors' claims around the consequences of LRRK2 deletion in DAergic neurons.

1. In Figures 2E and F, the authors assess the protein levels of LRRK1 and LRRK2 in their cDKO mouse model to confirm the deletion of both proteins. They observe a mild loss of LRRK1 and LRRK2 signals in the ventral midbrain compared to wild-type animals. While this is not surprising given other cell types that still express LRRK1 and LRRK2 would be present in their dissected ventral midbrain samples, it does not sufficiently confirm that LRRK1 and LRRK2 are not expressed in DAergic neurons. Additional data is needed to more directly demonstrate that LRRK1 and LRRK2 protein levels are reduced in DAergic neurons, including analysis of LRRK1 and LRRK2 protein levels via immunohistochemistry or FACS-based analysis of TH+ neurons.

2. The authors observed a significant but modest effect of LRRK1 and LRRK2 deletion on the number of TH+ neurons in the substantia nigra (12-15% loss at 20-24 months of age). It is unclear whether this extent of neuron loss is functionally relevant. To strengthen the impact of these data, additional studies are warranted to determine whether this translates into any PD-relevant deficits in the mice, including motor deficits or alterations in alpha-synuclein accumulation/aggregation.

3. The authors demonstrate that, unlike in the germline LRRK DKO mice, they do not observe any alterations in electron-dense vacuoles via EM. Given their data showing increased apoptosis and gliosis, it remains unclear how the loss of LRRK proteins leads to DAergic neuronal cell loss. Mechanistic studies would be insightful to understand better potential explanations for how the loss of LRRK1 and LRRK2 may impair cellular survival, and additional text should be added to the discussion to discuss potential hypotheses for how this might occur.

4. The authors discuss the potential implications of the neuronal cell loss observed in cDKO mice for LRRK1 and LRRK2 for therapeutic approaches targeting LRRK2 and suggest this argues that LRRK2 variants may exert their effects through a loss-of-protein function. However, all of the data generated in this work focus on a mouse in which both LRRK1 and LRRK2 have been deleted, and it is therefore difficult to make any definitive conclusions about the consequences of specifically targeting LRRK2. The authors note potential redundancy between the two LRRK proteins, and they should soften some of their conclusions in the discussion section around implications for the effects of LRRK2 variants. Human subjects that carry LRRK2 loss-of-function alleles do not have an increased risk for developing PD, which argues against the author's conclusions that LRRK2 variants associated with PD are loss-of-function. Additional text should be included in their discussion to better address these nuances and caution should be used in terms of extrapolating their data to effects observed with PD-linked variants in LRRK2.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment

This work describes new validated conditional double KO (cDKO) mice for LRRK1 and LRRK2 that will be useful for the field, given that LRRK2 is widely expressed in the brain and periphery, and many divergent phenotypes have been attributed previously to LRRK2 expression. The manuscript presents solid data demonstrating that it is the loss of LRRK1 and LRRK2 expression within the SNpc DA cells that is not well tolerated, as it was previously unclear from past work whether neurodegeneration in the LRRK double Knock Out (DKO) was cell autonomous or the result of loss of LRRK1/LRRK2 expression in other types of cells. Future studies may pursue the biochemical mechanisms underlying the reason for the apoptotic cells noted in this study, as here, the LRRK1/LRRK2 KO mice did not replicate the dramatic increase in the number of autophagic vacuoles previously noted in germline global LRRK1/LRRK2 KO mice.

We thank the editors for handling our manuscript and for the succinct summary that recognizes the significance of our findings and points out interesting directions for future studies. We also thank the reviewers for their helpful comments and positive evaluation of our work. Below, we have provided point-by-point responses to the reviewers’ comments.

Reviewer #1 (Public Review):

Summary:

This is an important work showing that loss of LRRK function causes late-onset dopaminergic neurodegeneration in a cell-autonomous manner. One of the LRRK members, LRRK2, is of significant translational importance as mutations in LRRK2 cause late-onset autosomal dominant Parkinson's disease (PD). While many in the field assume that LRRK2 mutant causes PD via increased LRRK2 activity (i.e., kinase activity), it is not a settled issue as not all disease-causing mutant LRRK2 exhibit increased activity. Further, while LRRK2 inhibitors are under clinical trials for PD, the consequence of chronic, long-term LRRK2 inhibition is unknown. Thus, studies evaluating the long-term impact of LRRK deficit have important translational implications. Moreover, because LRRK proteins, particularly LRRK2, are known to modulate immune response and intracellular membrane trafficking, the study's results and the reagents will be valuable for others interested in LRRK function.

Strengths:

This report describes a mouse model where the LRRK1 and LRRK2 gene is conditionally deleted in dopaminergic neurons. Previously, this group showed that while loss of LRRK2 expression does not cause brain phenotype, loss of both LRRK1 and LRRK2 causes a later onset, progressive degeneration of catecholaminergic neurons and dopaminergic (DAergic) neurons in the substantia nigra (SN), and noradrenergic neurons in the locus coeruleus (LC). However, because LRRK genes are widely expressed with some peripheral phenotypes, it was unknown if the neurodegeneration in the LRRK double knockout (DKO) was cell autonomous. To rigorously test this question, the authors have generated a double conditional (cDKO) allele where both LRRK1 and LRRK2 genes were targeted to contain loxP sites. In my view, this was beyond what is usually required, as most investigators might might combine one KO allele with another floxed allele. The authors provide a rigorous validation showing that the Driver (DAT-Cre) is expressed in most DAergic neurons in the SN and that LRRK levers are decreased selectively in the ventral midbrain. Using these mice, the authors show that the number of DAergic neurons is normal at 15 but significantly decreased at 20 months of age. Moreover, the authors show that the number of apoptotic neurons is increased by ~2X in aged SN, demonstrating increased ongoing cell death, as well as an increase in activated microglia. The degeneration is limited to DAergic neurons as LC neurons are not lost as this population does not express DAT. Overall, the mouse genetics and experimental analysis were performed rigorously, and the results were statistically sound and compelling.

Weaknesses:

I only have a few minor comments. First is that in PD and other degenerative conditions, loss of axons and terminals occurs prior to cell bodies. It might be beneficial to show the status of DAergic markers in the striatum. Second, previous studies indicate that very little, if any, LRRK1 is expressed in SN DAergic neurons. This also the case with the Allen Brain Atlas profile. Thus, authors should discuss the discrepancy as authors seem to imply significant LRRK1 expression in DA neurons.

We appreciate the reviewer’s recognition of the importance of the study as well as our rigorous experimental approaches and compelling results. Our responses to the reviewer's two minor comments are below.

1. DAergic markers in the striatum: We performed TH immunostaining in the striatum and quantified TH+ DA terminals in the striatum of DA neuron-specific LRRK cDKO and littermate control mice at the ages of 15 and 24 months. We found similar levels of TH immunoreactivity in the striatum of LRRK cDKO and littermate control mice at the age of 15 months (p = 0.6565, unpaired Student’s t-test) and significantly reduced levels of TH immunoreactivity in the striatum of LRRK cDKO, compared to control mice at the age of 24 months (~19%, p = 0.0215), suggesting an age-dependent loss of dopaminergic terminals in the striatum of DA neuron-specific LRRK cDKO mice. These results are now included as Figure 5 of the revised manuscript.

2. LRRK1 expression in the SNpc: It is shown in the Mouse brain RNA-seq dataset and the Allen Mouse brain ISH dataset (https://www.proteinatlas.org/ENSG00000154237-LRRK1/brain) that LRRK1 is broadly expressed in the mouse brain and is expressed at modest levels in the midbrain, comparable to the cerebral cortex. Indeed, our Western analysis also showed that levels of LRRK1 detected in the dissected ventral midbrain and the cerebral cortex of control mice are similar (40µg total protein loaded per lane; Figure 2E). Furthermore, we previously demonstrated that deletion of LRRK2 (or LRRK1) alone does not cause age-dependent loss of DA neurons in the SNpc, but deletions of both LRRK1 and LRRK2 result in age-dependent loss of DA neurons in LRRK DKO mice, indicating the functional importance of LRRK1 in the protection of DA neuron survival in the aging mouse brain (Tong et al., PNAS 2010, 107: 9879-9884, Giaime et al., Neuron 2017, 96: 796-807).

Reviewer #2 (Public Review):

Summary:

In this manuscript, Shen and collaborators described the generation of cDKO mice lacking LRRK1 and LRRK2 selectively in DAT-positive DAergic neurons. The Authors asked whether selective deletion of both LRRK isoforms could lead to a Parkinsonian phenotype, as previously reported by the same group in germline double LRRK1 and LRRK2 knockout mice (PMID: 29056298). Indeed, cDKO mice developed a late reduction of TH+ neurons in SNpc that partially correlated with the reduction of NeuN+ cells. This was associated with increased apoptotic cell and microglial cell numbers in SNpc.

Unlike the constitutive DKO mice described earlier, however, cDKO mice did not replicate the dramatic increase in the number of autophagic vacuoles. The study supports the authors' hypothesis that loss of function rather than gain of function of LRRK2 leads to PD.

Strengths:

The study described for the first time a model where both the PD-associated gene LRRK2 and its homolog LRRK1 are deleted selectively in DAergic neurons, offering a new tool to understand the physiopathological role of LRRK2 and the compensating role of LRRK1 in modulating DAergic cell function.

Weaknesses:

The model has no construct validity since loss of function mutations of LRRK2 are well-tolerated in humans and do not lead to PD. The evidence of a Parkinsonian phenotype in these cDKO mice is limited and should be considered preliminary.

We thank the reviewer for commenting on the usefulness of this new PD mouse model.

The reviewer did not include a reference citation for the statement "loss of function mutations of LRRK2 are well-tolerated in humans and do not lead to PD." It is possible that the reviewer was referring to a human population study (Whiffin et al., Nat Med 2020, 26: 869-877), entitled "The effect of LRRK2 lossof-function variants in humans." In this study, the authors analyzed 141,456 individuals sequenced in the Genome Aggregation Database, 49,960 exome-sequenced individuals from the UK Biobank, and more than 4 million participants in the 23andMe genotyped dataset, and they looked for human genetic variants predicted to cause loss-of-function of protein-coding genes (pLoF variants). The reported findings were interesting, and the authors were careful in stating their conclusions. However, this is not a linkage study of large pedigrees carrying a single, clear-cut loss-of-function mutation (e.g. large deletions of most exons and coding sequences). Therefore, the experimental evidence is not compelling enough to conclude whether loss-of-function mutations in LRRK2 cause PD or do not cause PD.

The current report is an unbiased genetic study in an effort to reveal the normal physiological role of LRRK in dopaminergic neurons. It was not intended to produce Parkinsonian phenotypes in LRRK cDKO mice, which would be a biased effort. However, the unequivocal discovery of the cell intrinsic role of LRRK in the protection of DA neurons from age-dependent degeneration and apoptotic cell death should be considered seriously, while we contemplate the disease mechanism and how LRRK2 mutations may cause DA neuron loss and PD.

Reviewer #3 (Public Review):

Kang, Huang, and colleagues investigated the impact of LRRK1 and LRRK2 deletion, specifically in dopaminergic neurons, using a novel cDKO mouse model. They observed a significant reduction in DAergic neurons in the substantia nigra in their conditional LRRK1 and LRRK2 KO mice and a corresponding increase in markers of apoptosis and gliosis. This work set out to address a longstanding question within the field around the role and importance of LRRK1 and LRRK2 in DAergic neurons and suggests that the loss of both proteins triggers some neurodegeneration and glial activation.

The studies included in this work are carefully performed and clearly communicated, but additional studies are needed to strengthen further the authors' claims around the consequences of LRRK2 deletion in DAergic neurons.

1. In Figures 2E and F, the authors assess the protein levels of LRRK1 and LRRK2 in their cDKO mouse model to confirm the deletion of both proteins. They observe a mild loss of LRRK1 and LRRK2 signals in the ventral midbrain compared to wild-type animals. While this is not surprising given other cell types that still express LRRK1 and LRRK2 would be present in their dissected ventral midbrain samples, it does not sufficiently confirm that LRRK1 and LRRK2 are not expressed in DAergic neurons. Additional data is needed to more directly demonstrate that LRRK1 and LRRK2 protein levels are reduced in DAergic neurons, including analysis of LRRK1 and LRRK2 protein levels via immunohistochemistry or FACS-based analysis of TH+ neurons.

We thank the reviewer for highlighting this incredibly important but often overlooked issue. We agree that the data in Figure 2E, F alone would be inadequate to validate DA neuron-specific LRRK cDKO mice.

Cell type-specific conditional knockouts are a mosaic with KO cells mixed with other cell types expressing the gene normally. DA neuron-specific cDKO is particularly challenging, as DA neurons are a subset of cells embedded in the ventral midbrain. Rather than using immunostaining, which relies upon specific, good LRRK1 and LRRK2 antibodies for IHC, or FACS sorting of TH+ neurons followed by Western blotting (few cells, mixed cell populations, etc.), we chose a clean genetic approach by generating germline mutant mice carrying the deleted LRRK1 and LRRK2 alleles in all cells from the floxed LRRK1 and LRRK2 alleles. This approach permits characterization of these deletion mutations in germline mutant mice using molecular approaches that yield unambiguous results.

We crossed CMV-Cre deleter mice with floxed LRRK1 and LRRK2 mice to generate respective germline LRRK1 KO and LRRK2 KO mice, in which all cells carry the LRRK1 or LRRK2 deleted alleles that are identical to those in DA neurons of cDKO mice. We then performed Northern, extensive RTPCR followed by sequencing, and Western analyses to show the absence of the full length LRRK1 and LRRK2 mRNA (Figure 1G, H, Figure 1-figure supplement 8 and 10), and the expected truncation of LRRK1 and LRRK2 mRNA (Figure 1-figure supplement 9 and 11), and the absence of LRRK1 and LRRK2 proteins (Figure 1I). These analyses together demonstrate that in the presence of Cre, either CMV-Cre expressed in all cells or DAT-Cre expressed selectively in DA neurons, the floxed LRRK1 and LRRK2 exons are deleted, resulting in null alleles. We further demonstrated the specificity of DAT-Cremediated recombination (deletion) by crossing DAT-Cre mice with a GFP reporter, showing that 99% TH+ DA neurons in the SNpc are also GFP+ (Figure 2A, B), indicating that DAT-Cre-mediated recombination of the floxed alleles occurs in essentially all TH+ DA neurons in the SNpc.

1. The authors observed a significant but modest effect of LRRK1 and LRRK2 deletion on the number of TH+ neurons in the substantia nigra (12-15% loss at 20-24 months of age). It is unclear whether this extent of neuron loss is functionally relevant. To strengthen the impact of these data, additional studies are warranted to determine whether this translates into any PD-relevant deficits in the mice, including motor deficits or alterations in alpha-synuclein accumulation/aggregation.

Yes, the reduction of DA neurons in the SNpc of cDKO mice at the age of 20-24 months is modest. At 15 months of age, the number of TH+ DA neurons in the SNpc is similar between LRRK cDKO mice (10,000 ± 141) and littermate controls (10,077 ± 310, p > 0.9999). At 20 months of age, the number of DA neurons in the SNpc of LRRK cDKO mice (8,948 ± 273) is significantly reduced (-12.7%), compared to control mice (10,244 ± 220, F1,46 = 16.59, p = 0.0002, two-way ANOVA with Bonferroni’s post hoc multiple comparisons, p = 0.0041). By 24 months of age, the number of DA neurons in the SNpc of LRRK cDKO mice (8,188 ± 452) relative to controls (9,675 ± 232, p = 0.0010) is further reduced (15.4%).

Similar results were obtained by an independent quantification by another investigator, also conducted in a genotype blind manner, using the fractionator and optical dissector method, by which TH+ cells were quantified in 25% areas. These results are included as Figure 3-figure supplement 1 in the revised manuscript. Because of the more limited sampling, the quantification data are more variable, compared to quantification of TH+ cells in all areas of the SNpc, shown in Figure 3. With both methods, we quantified TH+ cells in every 10th sections encompassing the entire SNpc (3D structure), as sampling using every 5th or every 10th sections yielded similar results.

We also performed behavioral analysis of LRRK cDKO mice and littermate controls at the ages of 10 and 25 months using the beam walk test (10 mm and 20 mm beam) and the pole test, which are sensitive to impairment of motor coordination. We found that LRRK cDKO mice at 10 months of age showed significantly more hindlimb errors (p = 0.0005, unpaired two-tailed Student’s t-test) and longer traversal time (p = 0.0075) in the 10mm beam walk test, compared to control mice, though their performance is similar in the 20 mm beam walk (hindlimb slips: p = 0.0733, traversal time: p = 0.9796) and in the pole test. At 22 months of age, the performance of LRRK cDKO mice and littermate controls is more variable and worse, compared to the younger mice, and is not significantly different between the genotypic groups. These results are now included as Figure 9 of the revised manuscript.

1. The authors demonstrate that, unlike in the germline LRRK DKO mice, they do not observe any alterations in electron-dense vacuoles via EM. Given their data showing increased apoptosis and gliosis, it remains unclear how the loss of LRRK proteins leads to DAergic neuronal cell loss. Mechanistic studies would be insightful to understand better potential explanations for how the loss of LRRK1 and LRRK2 may impair cellular survival, and additional text should be added to the discussion to discuss potential hypotheses for how this might occur.

We agree that this phenotypic difference between germline DKO and DA neuron-specific cDKO mice is intriguing, suggesting a non-cell autonomous contribution of LRRK in age-dependent accumulation of autophagic and lysosomal vacuoles in SNpc neurons of germline LRRK DKO mice. We will discuss the phenotypic difference further in the revised manuscript. We are generating microglial specific LRRK cDKO mice to investigate the role of LRRK in microglia and whether microglia contribute in a cell extrinsic manner to the regulation of the autophagy-lysosomal pathway in DA neurons.

1. The authors discuss the potential implications of the neuronal cell loss observed in cDKO mice for LRRK1 and LRRK2 for therapeutic approaches targeting LRRK2 and suggest this argues that LRRK2 variants may exert their effects through a loss-of-protein function. However, all of the data generated in this work focus on a mouse in which both LRRK1 and LRRK2 have been deleted, and it is therefore difficult to make any definitive conclusions about the consequences of specifically targeting LRRK2. The authors note potential redundancy between the two LRRK proteins, and they should soften some of their conclusions in the discussion section around implications for the effects of LRRK2 variants. Human subjects that carry LRRK2 loss-of-function alleles do not have an increased risk for developing PD, which argues against the author's conclusions that LRRK2 variants associated with PD are loss-o-ffunction. Additional text should be included in their discussion to better address these nuances and caution should be used in terms of extrapolating their data to effects observed with PD-linked variants in LRRK2.

We will modify the discussion accordingly in the revised manuscript.

Version 2

Abstract

Abstract

Mutations in leucine-rich repeat kinase 2 (LRRK2) are the most common genetic cause of Parkinson’s disease (PD), which is the leading neurodegenerative movement disorder characterized by the progressive loss of dopaminergic (DA) neurons in the substantia nigra pars compacta (SNpc). However, whether LRRK2 mutations cause PD and degeneration of DA neurons via a toxic gain-of-function or a loss-of-function mechanism is unresolved and has pivotal implications for LRRK2-based PD therapies. In this study, we investigate whether LRRK2 and its functional homologue LRRK1 play an essential, intrinsic role in DA neuron survival through the development of DA neuron-specific LRRK conditional double knockout (cDKO) mice. We first generated and characterized floxed LRRK1 and LRRK2 mice and then confirmed that germline deletions of the floxed LRRK1 and LRRK2 alleles result in null mutations, as evidenced by the absence of LRRK1 and LRRK2 mRNA and protein in the respective homozygous deleted mutant mice. We further examined the specificity of Cre-mediated recombination driven by the dopamine transporter-Cre (DAT-Cre) knockin (KI) allele using a GFP reporter line and confirmed that DAT-Cre-mediated recombination is restricted to DA neurons in the SNpc. Crossing these validated floxed LRRK1 and LRRK2 mice with DAT-Cre KI mice, we then generated DA neuron-restricted LRRK cDKO mice and further showed that levels of LRRK1 and LRRK2 are reduced in dissected ventral midbrains of LRRK cDKO mice. While DA neuron-restricted LRRK cDKO mice of both sexes exhibit normal mortality and body weight, they develop age-dependent loss of DA neurons in the SNpc, as demonstrated by the progressive reduction of DA neurons in the SNpc of LRRK cDKO mice at the ages of 20 and 24 months but the unaffected number of DA neurons at the age of 15 months. Moreover, DA neurodegeneration is accompanied with increases of apoptosis and elevated microgliosis in the SNpc as well as decreases of DA terminals in the striatum, and is preceded by impaired motor coordination. Taken together, these findings provide the unequivocal evidence for the importance of LRRK in DA neurons and raise the possibility that LRRK2 mutations may impair its protection of DA neurons, leading to DA neurodegeneration in PD.

Introduction

Parkinson’s disease (PD) is the most common movement disorder and is characterized by the progressive loss of dopaminergic (DA) neurons in the substantia nigra pars compacta (SNpc). Dominantly inherited missense mutations in the leucine-rich repeat kinase 2 (LRRK2) gene are the most common cause of both familial and sporadic PD, highlighting the importance of LRRK2 in PD pathogenesis (1-16). LRRK2 is a large protein of 2,527 amino acid residues containing multiple functional domains including several leucine-rich repeats (LRRs), a GTPase-like domain of Ras-of-complex (Roc), a C-terminal of Roc (COR) domain, and a serine/threonine MAPKKK-like kinase domain. LRRK1, a homologue of LRRK2, belongs to the evolutionarily conserved Roco protein family and contains similar LRRs, Roc, COR, and kinase domains (17-19). LRRK proteins are broadly expressed with LRRK2 being most abundant in the kidney (20). While most PD mutations are found in LRRK2, rare variants in the Roc, COR, and kinase domains of LRRK1 have been reported and may be associated with PD (21). Furthermore, the Roc-COR domain of LRRK2 forms dimers and exhibits conventional Ras-like GTPase properties, and the R1441/C/G/H and I1371V mutations destabilize dimer formation and decrease GTPase activity (22, 23). Recent high-resolution cryoEM structural studies of full-length LRRK2 demonstrated its existence as dimers and pathogenic mutations such as R1441/C/G/H and Y1699I at the Roc-COR interface, whereas the G2910S mutant and wild-type LRRK2 share similar structure (24).

Previous genetic studies demonstrated that LRRK2 plays essential roles in the autophagy-lysosomal pathway (25-28). Consistent with high levels of LRRK2 expression in kidneys, LRRK2-/- mice develop age-dependent phenotypes in the kidney, including autophagy-lysosomal impairments and increases of α-synuclein, apoptosis, and inflammatory responses (25, 26, 29). It has also been reported that LRRK2 is important for maintaining lung homeostasis, and LRRK2 deficiency results in impaired autophagy in alveolar type II epithelial cells (28). It was proposed that the lack of brain phenotypes in LRRK2-/- mice might be due to the presence of LRRK1 that compensates functionally in the absence of LRRK2 (25, 26). Indeed, LRRK double knockout (DKO) mice develop age-dependent, progressive loss of DA neurons in the SNpc, beginning at 14 months of age (30). However, LRRK DKO mice also exhibit lower body weight and increased mortality, raising the possibility that DA neurodegeneration in LRRK DKO mice may be secondary to poor health.

In this study, we investigate whether LRRK2 and its functional homologue LRRK1 play an essential, intrinsic role in DA neuron survival through the development of DA neuron-specific LRRK conditional DKO (cDKO) mice. We first generated and confirmed floxed LRRK1 and LRRK2 mice, and then crossed them with the CMV-Cre deleter to create germline deletions of the floxed LRRK1 and LRRK2 alleles, followed by Northern and Western analyses to confirm the absence of LRRK1 and LRRK2 mRNA and protein in the respective homozygous deleted mutant mice. We also crossed a GFP reporter mouse with DAT-Cre KI mice and confirmed that Cre-mediated recombination occurs in most, if not all, DA neurons in the SNpc and is restricted to DA neurons. We then crossed these thoroughly validated floxed LRRK1 and LRRK2 mice with DAT-Cre KI mice to generate DA neuron-restricted LRRK cDKO mice and further confirmed the reduction of LRRK1 and LRRK2 in dissected ventral midbrains of LRRK cDKO mice. While DA neuron-restricted LRRK cDKO mice of both sexes exhibit normal mortality and body weight, they develop age-dependent loss of DA neurons in the SNpc, as demonstrated by the progressive reduction of TH+ DA neurons or NeuN+ neurons in the SNpc of cDKO mice at the ages of 20 and 24 months. Moreover, DA neurodegeneration is accompanied with increases of apoptosis and elevated microgliosis in the SNpc as well as decreases of DA terminals in the striatum, and is preceded by impaired motor coordination. Interestingly, quantitative electron microscopy (EM) analysis showed a similar number of electron-dense vacuoles in SNpc neurons of LRRK cDKO mice relative to controls, in contrast to age-dependent increases of vacuoles in SNpc neurons of germline LRRK DKO mice (30, 31). These findings provide the unequivocal evidence for the importance of LRRK in DA neurons and raise the possibility that LRRK2 mutations may impair this crucial physiological function, leading to the loss of DA neurons in Parkinson’s disease.

Results

Generation and molecular characterization of the floxed and deleted LRRK1 and LRRK2 alleles

LRRK2 and its homologue LRRK1 share several functional (LRRs, GTPase Roc, COR, and kinase) domains (Figure 1A). To investigate the intrinsic role of LRRK in DA neurons, we generated floxed LRRK1 (fLRRK1) and floxed LRRK2 (fLRRK2) mice, which permit deletions of LRRK1 and LRRK2 selectively in DA neurons by DAT-Cre, in which Cre recombinase is expressed under the control of the endogenous DAT promoter (32). We introduced two loxP sites in introns 26 and 29 of LRRK1 through homologous recombination and site-specific recombination by FLP recombinase to remove the positive selection PGK-neo cassette, which is flanked by two FRT sites (Figure 1B, C; Figure 1-figure supplement 1, 2). The embryonic stem (ES) cells carrying the targeted allele or the floxed allele were identified and validated by Southern analysis using the 5’ and 3’ external probes as well as the neo probe (Figure 1D; Figure 1-figure supplement 3). The validated ES cells carrying the floxed LRRK1 allele were injected into mouse blastocysts to generate floxed LRRK1 mice, which were further confirmed by Southern using the 5’ and 3’ external probes (Figure 1D). In the presence of Cre recombinase, the floxed LRRK1 genomic region containing part of intron 26 (1,288 bp), exons 27-29, which encode the kinase domain, and part of intron 29 (1,023 bp) is deleted (Figure 1C), and the removal of exons 27-29 (625 bp) results in a frameshift of downstream codons.

Generation and characterization of floxed and deleted LRRK1 and LRRK2 alleles

(A) Schematic illustrations of human LRRK1 and LRRK2 proteins showing similar functional domains. LRRK12015 protein is derived from exons 2-34 (Ensembl Genome Database: ENSG00000154237). LRRK22527 protein is derived from exons 1-51 (ENSG00000188906). LRR: Leucine rich repeats; Roc: Ras-of-complex; COR: C-terminal of Roc; KIN: Kinase domain.

(B) Schematic illustrations of the gene structures of mouse LRRK1 and LRRK2. The boxes in blue are exons that encode the LRRK1 and LRRK2 proteins, and the gray boxes represent the 5’ and 3’ UTRs. The exons are not drawn in scale. The start codon ATG is in exon 2 of LRRK1 and exon 1 of LRRK2. The exons 27-29 of LRRK1 and the promoter/exons 1-2 of LRRK2 are flanked with loxP sites (green arrowheads).

(C) Targeting strategy for the generation of the targeted, floxed, and deleted LRRK1 alleles. The red boxes represent the targeted exons 27-29, and the blue boxes represent the untargeted exons. The locations and sizes of the 5’ and 3’ external probes are shown. The targeting vector contains the 5’ and 3’ homologous regions (marked by dashed lines) and the middle region (from intron 26 to intron 29), which includes a loxP site (green arrowhead) in intron 26 (1,288 bp upstream of exon 27) and the PGK-neo selection cassette flanked by two FRT (FLP recognition target) sequences (gray circles) followed by another loxP site in intron 29 (1,023 bp downstream of exon 29). A negative selection cassette encoding diphtheria toxin fragment A (DTA) is also included in the targeting vector to reduce ES cells bearing randomly inserted targeting vectors. ES cells carrying the correctly targeted LRRK1 allele were transfected with pCAG-FLP to remove the PGK-neo cassette and generate the floxed LRRK1 allele. Floxed LRRK1 mice were bred with CMV-Cre transgenic mice to generate germline deleted LRRK1 mice. Detailed strategy for generating targeting vector and DNA sequence of floxed LRRK1 allele can be found in Figure 1-figure supplement 1 and 2, respectively.

(D) Southern analysis of the targeted and floxed LRRK1 alleles. Genomic DNA from ES cells or mouse tails was digested with HindIII and hybridized with the 5’ or 3’ external probe. For the 5’ probe, the resulting 17.0 kb and 4.8 kb bands represent the wild-type (WT) and the targeted (T) or floxed (F) alleles, respectively. For the 3’ probe, the resulting 17.0 kb and 12.2 kb bands represent the WT and the floxed alleles, respectively. Detailed Southern strategy can be found in Figure 1-figure supplement 3.

(E) Targeting strategy for the generation of the targeted, floxed, and deleted LRRK2 alleles. The red boxes represent LRRK2 exons 1 and 2, and the start codon ATG resides in exon 1. The locations and sizes of the 5’ and 3’ external probes are shown. The targeting vector contains the 5’ and 3’ homologous regions (marked by dashed lines) and the middle region (from the promoter to intron 2), which includes a loxP site (green arrowhead) upstream (1,768 bp) of the transcription initiation site and the PGK-neo selection cassette flanked by two FRT sequences (gray circles) and two loxP sites (green arrowheads) in intron 2 (878 bp downstream of exon 2). A negative selection cassette encoding DTA is also included in the targeting vector. Mice carrying the correctly targeted LRRK2 allele were crossed with Actin-FLP deleter mice to generate floxed LRRK2 mice, which were bred with CMV-Cre transgenic mice to generate germline deleted LRRK2 mice. Detailed strategy for generating targeting vector and DNA sequence of floxed LRRK2 allele can be found in Figure 1-figure supplement 4 and 5, respectively.

(F) Southern analysis of the targeted and floxed LRRK2 alleles. Genomic DNA from ES cells or mouse tails was digested with NheI and hybridized with the 5’ or 3’ external probe. For the 5’ probe, the resulting 11.5 kb and 3.6 kb bands represent the wild-type (WT) and the targeted (T) or floxed (F) alleles, respectively. For the 3’ probe, the resulting 11.5 kb and 5.2 kb bands represent the WT and the floxed alleles, respectively. Detailed Southern strategy can be found in Figure 1-figure supplement 6.

(G) Northern analysis of poly(A)+ RNA prepared from the lung of mice carrying homozygous germline deleted (Δ/Δ) LRRK1 alleles derived from the floxed LRRK1 alleles using the cDNA probe of exons 2-3 (left) and exons 27-29 (right). Using the upstream exons 2-3 probe, the LRRK1 transcripts in wild-type mice are the expected size of ∼7.4 kb, whereas the detected LRRK1 transcripts in LRRK1 Δ/Δ mice are truncated, consistent with the deletion of exons 27-29 (625 bp), which results in a frameshift, and are expressed at lower levels, likely due to nonsense mediated degradation of the truncated LRRK1 mRNA. Using a probe specific for exons 27-29, there is no LRRK1 transcript in LRRK1 Δ/Δ mice, as expected. Both blots were hybridized with a GAPDH probe as loading controls. Detailed Northern strategy and full-size blots are included in Figure 1-figure supplement 7 and 8, respectively. Extensive RT-PCR analysis of LRRK1 transcripts in Δ/Δ mice are shown in Figure 1-figure supplement 9.

(H) Northern analysis of total RNA prepared from the neocortex of mice carrying homozygous germline LRRK2 deleted (Δ/Δ) alleles using the cDNA probe of exons 1-5 shows the absence of LRRK2 transcripts. The blot was hybridized with a GAPDH probe as a loading control. The full-size blot is included in Figure 1-figure supplement 10. RT-PCR analysis of LRRK2 transcripts in Δ/Δ mice are shown in Figure 1-figure supplement 11

(I) Left: Western analysis of wild-type (+/+) and homozygous LRRK1 Δ/Δ brains show the absence of LRRK1 protein. Right: Western analysis of the neocortex of wild-type (+/+) and homozygous LRRK2 Δ/Δ mice shows the absence of LRRK2 protein. Vinculin was used as a loading control. Full-size blots can be found in Figure 1-Source Data 1.

The LRRK2 targeting vector contains the 5’ homologous region, a loxP site 1,768 bp upstream of the transcription initiation site, exons 1-2, the PGK-neo cassette flanked by two loxP sites and two FRT sequences, and the 3’ homologous region (Figure 1B, E; Figure 1-figure supplement 4, 5). The ES cell clones carrying the targeted allele were identified and validated by Southern analysis using the 5’ external probe (Figure 1F; Figure 1-figure supplement 6) and genomic PCR followed by sequencing. Mice carrying the LRRK2 targeted allele were further verified by Southern analysis using the 5’ and 3’ external probes (Figure 1-figure supplement 6) as well as the neo probe and were then crossed with the Actin-FLP mice (33) to remove the PGK-neo cassette flanked by two FRT sites to generate floxed LRRK2 mice (Figure 1E). The resulting floxed LRRK2 mice were confirmed by Southern analysis using the 5’ and 3’ external probes (Figure 1F). In the presence of Cre recombinase, the floxed LRRK2 region containing the promoter and exons 1-2 are deleted, likely resulting in a null allele, as we previously targeted a very similar region (∼2.5 kb upstream of the transcription initiation site and exons 1-2) to generate germline deletion of LRRK2 (25).

Dopaminergic neurons in the SNpc are a small neuronal population embedded in the ventral midbrain, making it difficult to confirm whether DA neuron-specific deletions of the floxed LRRK1 and LRRK2 regions result in null alleles. We previously generated three independent LRRK1 knockout (KO) mice, and only one KO line (Line 2) represents a LRRK1 null allele (30), whereas deletion of LRRK2 promoter region and exons 1-2 resulted in a LRRK2 null allele (25). We therefore generated germline deleted LRRK1 and LRRK2 mice from floxed LRRK1 and LRRK2 mice, respectively, by crossing them to germline deleter CMV-Cre transgenic mice (34).

We then performed Northern analysis of LRRK1 using both an upstream probe specific for exons 2-3 and a downstream probe specific for exons 27-29 (Figure 1G; Figure 1-figure supplement 7, 8). Because of the low expression level of LRRK1 mRNA and the relative abundance of LRRK1 mRNA in the lung (20), we enriched polyA+ RNA from the lung of the mice carrying homozygous germline deleted (Δ/Δ) LRRK1 alleles derived from the floxed LRRK1 alleles. Using the exons 2-3 probe, the LRRK1 transcripts in wild-type mice are of the expected size of ∼7.4 kb, whereas the detected LRRK1 transcripts in LRRK1 Δ/Δ mice are smaller and expressed at lower levels, consistent with the deletion of exons 27-29 (625 bp), resulting in a frameshift of downstream codons and the likely degradation of the truncated LRRK1 mRNA (Figure 1G; Figure 1-figure supplement 8). Using a probe specific for exons 27-29, there is no LRRK1 transcript in LRRK1 Δ/Δ mice (Figure 1G; Figure 1-figure supplement 8). Extensive RT-PCR analysis of total RNA isolated from the kidney, brain, and lung of LRRK1 Δ/Δ mice using an exon 32-specific primer for RT and exon-specific primer sets for PCR (e.g. exons 4-8, 11-17, 20-25, and 25-31), followed by sequencing confirmation of the PCR products, indicated normal LRRK1 splicing in fLRRK1/fLRRK1 mice and the lack of LRRK1 exons 27-29 in LRRK1 Δ/Δ mice (Figure 1-figure supplement 9).

Similarly, Northern analysis of LRRK2 using a probe specific for exons 1-5 and RT-PCR followed by sequencing confirmed the absence of LRRK2 mRNA in LRRK2 Δ/Δ brains and normal LRRK2 transcripts in fLRRK2/fLRRK2 brains (Figure 1H; Figure 1-figure supplement 10, 11). Furthermore, Western analysis confirmed the absence of LRRK1 and LRRK2 proteins in the brain of LRRK1 Δ/Δ and LRRK2 Δ/Δ mice, respectively (Figure 1I). Taken together, our Northern, RT-PCR followed by sequencing, and Western analyses demonstrated that deletion of the floxed LRRK1 and LRRK2 alleles results in null mutations. Thus, floxed LRRK1 and LRRK2 alleles can be used to generate DA neuron-specific LRRK cDKO mice.

Generation and molecular characterization of DA neuron-specific LRRK cDKO mice

To generate DA neuron-specific LRRK cDKO mice, we used DAT-Cre KI mice, which express Cre recombinase under the control of the endogenous DAT promoter (32). To confirm if Cre-mediated recombination occurs broadly and specifically in DA neurons of the SNpc, we crossed DAT-Cre mice with the Rosa26-CAG-LSL-ZsGreen1 reporter mouse (35). Upon Cre expression, Cre recombinase removes the floxed “stop” cassette, resulting in the expression of EGFP. We found that Cre-mediated recombination (GFP+) occurs in TH+ cells in the SNpc (Figure 2A). Quantification of GFP+ and/or TH+ cells in the SNpc showed that 99% of TH+ DA neurons are also GFP+, demonstrating that Cre-mediated recombination takes place in essentially all DA neurons in the SNpc (Figure 2B).

Generation and characterization of DA neuron-specific LRRK cDKO mice

(A) Immunostaining of GFP and/or TH in the SNpc of DAT-Cre/+; Rosa26-CAG-LSL-ZsGreen1/+ mice at 2 months of age. Cre recombinase is expressed under the control of the DAT endogenous promoter and removes the floxed “stop” cassette, resulting in the expression of EGFP under the control of the ubiquitous CAG promoter.

(B) Quantification of GFP+/TH+ and TH+ cells shows that 99% of TH+ DA neurons (722 ± 46 TH+ cells) in the SNpc are also GFP+ (713 ± 46 cells), indicating that DAT-Cre mediated recombination occurs in essentially all TH+ DA neurons. N = 3 mice, 3 comparable sections per hemisphere, 320 μm apart.

(C) Similar body weight between LRRK cDKO mice and littermate controls at all ages examined (F1,68 = 0.001310, p = 0.9712; 2M, 20M: p > 0.9999; 15M: p = 0.7857, 25M: p = 0.8084, two-way ANOVA with Bonferroni’s post hoc multiple comparisons).

(D) Similar brain weight between LRRK cDKO and control mice (F1,68 = 3.603, p = 0.0619; 2M: p = 0.3893; 15M: p > 0.9999; 20M: p > 0.9999; 25M: p = 0.3223, two-way ANOVA with Bonferroni’s post hoc multiple comparisons).

(E) Western analysis of LRRK1 and LRRK2 proteins in the dissected cerebral cortex (CTX) and ventral midbrain (vMB) of LRRK cDKO and littermate controls at 2 months of age.

(F) Quantification shows significant decreases of LRRK1 and LRRK2 in the dissected ventral midbrain of LRRK cDKO mice (LRRK1, p = 0.0432; LRRK2, p = 0.0162, unpaired two-tailed Student’s t-test), compared to controls, but not in the dissected cortex of cDKO mice (LRRK1: p = 0.7648; LRRK2: p = 0.2325).

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. *p < 0.05. Scale bar: 100 μm.

Having confirmed fLRRK1 and fLRRK2 mice as well as DAT-Cre mice, we then bred them together to generate LRRK cDKO mice (fLRRK1/fLRRK1; fLRRK2/fLRRK2; DAT-Cre/+), which were further bred with fLRRK1/fLRRK1; fLRRK2/fLRRK2 mice to generate cDKO and littermate controls (fLRRK1/fLRRK1; fLRRK2/fLRRK2). It was previously reported that germline LRRK DKO mice failed to gain weight as they aged (30). However, DA neuron-specific LRRK cDKO and littermate control mice have similar body and brain weights at the ages of 2-24 months (Figure 2C, D). Western analysis showed a significant reduction of LRRK1 and LRRK2 proteins in the dissected ventral midbrain but not in the cerebral cortex of DA neuron-specific LRRK cDKO mice at 2-3 months of age, relative to littermate controls (Figure 2E, F), further validating these DA neuron-specific LRRK cDKO mice.

Age-dependent loss of TH+ DA neurons in the SNpc of LRRK cDKO mice

To determine whether the inactivation of LRRK selectively in DA neurons of the SNpc affects their survival, we performed TH immunostaining and quantified TH+ DA neurons in the SNpc of LRRK cDKO mice and littermate controls using stereological methods. The morphology of TH+ DA neurons in LRRK cDKO mice at the ages of 15, 20, and 24 months appears normal (Figure 3A). Quantification of TH+ neurons in the SNpc using serial sections emcompassing the entire SNpc revealed that the number of DA neurons in the SNpc at the age of 15 months is similar between cDKO mice (10,000 ± 141) and littermate controls (10,077 ± 310; F1,46 = 16.59, p = 0.0002, two-way ANOVA with Bonferroni’s post hoc multiple comparisons, p > 0.9999; Figure 3B). However, at the age of 20 months, the number of TH+ neurons in the SNpc of cDKO mice (8,948 ± 273) is significantly reduced, compared to controls (10,244 ± 220; F1,46 = 16.59, p = 0.0002, two-way ANOVA with Bonferroni’s post hoc multiple comparisons, p = 0.0041), and is further decreased at the age of 24 months (Control: 9,675 ± 232, cDKO: 8,188 ± 452; p = 0.0010; Figure 3B). Similar genotypic differences were observed in an independent quantification by another investigator, also conducted in a genotype blind manner, using the fractionator and optical dissector to randomly sample 25% area of the SNpc (Figure 3-figure supplement 1).

Age-dependent loss of DA neurons in the SNpc of LRRK cDKO mice

(A) TH immunostaining shows TH+ DA neurons in the SNpc of LRRK cDKO and littermate controls at the age of 15, 20, and 24 months. Higher power views of the boxed areas show grossly normal DA neuron morphology in LRRK cDKO mice.

(B) Quantification of TH+ DA neurons in the SNpc reveals similar numbers of DA neurons in LRRK cDKO mice (10,000 ± 141) and littermate controls (10,077 ± 310, p > 0.9999) at 15 months of age. At 20 months of age, the number of DA neurons in the SNpc of LRRK cDKO mice (8,948 ± 273) is significantly reduced, compared to control mice (10,244 ± 220, F1,46 = 16.59, p = 0.0002; p = 0.0041, two-way ANOVA with Bonferroni’s post hoc multiple comparisons). By 24 months of age, the reduction of DA neurons in the SNpc of LRRK cDKO mice (8,188 ± 452) relative to controls (9,675 ± 232, p = 0.0010) is greater, compared to 20 months of age. Raw quantification data are included in Figure 3-Source Data 1.

(C) Immunohistological analysis of TH and NeuN shows TH+ DA neurons (green) and NeuN+ neurons (red) in the SNpc of LRRK cDKO mice and controls at 24 months of age.

(D) Quantification of NeuN+ cells in the SNpc shows that the number of NeuN+ neurons in LRRK cDKO mice (17,923 ± 813) is significantly lower than that in control mice (21,907 ± 469, p = 0.0006, Student’s t-test), indicating loss of neurons in the SNpc of LRRK cDKO mice. All TH+ cells are NeuN+. The number of TH+/NeuN+ cells in the SNpc of LRRK cDKO mice (10,500 ± 644) is also lower, compared to control mice (14,102 ± 310, p = 0.0001). There is no significant difference in the number of NeuN+/TH-neurons between littermate controls (7,804 ± 249) and cDKO mice (7,423 ± 344, p = 0.3747).

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. **p < 0.01, ***p < 0.001. Scale bar: 100 μm.

We further performed TH and NeuN double immunostaining of LRRK cDKO and control mice at 24 months of age (Figure 3C). Quantification of NeuN+ and TH+/NeuN+ cells in the SNpc using serial sections encompassing the entire SNpc showed that the number of NeuN+ neurons is also significantly reduced in the SNpc of cDKO mice (17,923 ± 813), compared to controls (21,907 ± 469, p = 0.0006, Student’s t-test; Figure 3D). The number of TH+/NeuN+ cells is also lower in the SNpc of LRRK cDKO mice (10,500 ± 644), compared to control mice (14,102 ± 310, p = 0.0001; Figure 3D). These data indicate that the reduction in TH+ cells is not due to decreases of TH expression in DA neurons, but rather a result of the loss of DA neuron cell bodies in the SNpc of LRRK cDKO mice.

We further evaluated apoptosis in the SNpc of LRRK cDKO and littermate controls at the age of 24 months using an antibody specific for active Caspase-3 to label apoptotic cells. We observed increases in apoptotic DA neurons, labeled by active Caspase-3+ and TH+ immunoreactivity, in the SNpc of LRRK cDKO mice (Figure 4A). Quantification of active Caspase-3+/TH+ apoptotic DA neurons shows a significant increase in the SNpc of LRRK cDKO mice (323 ± 38), compared to controls (157 ± 8, p = 0.0004, Student’s t-test; Figure 4B). These results further support that LRRK plays an intrinsic role in the survival of DA neurons in the SNpc during aging.

Increases in apoptotic DA neurons in the SNpc of LRRK cDKO mice

(A) Representative images of TH and active Caspase-3 immunostaining show TH+ DA neurons (green) and active Caspase-3+ apoptotic cells (red) in the SNpc of LRRK cDKO and control mice at the age of 24 months.

(B) Quantification of active Caspase-3+/TH+ cells shows significant increases of apoptotic DA neurons in the SNpc of LRRK cDKO mice (323 ± 38) at 24 months of age, relative to controls (157 ± 8, p = 0.0004, unpaired Student’s t-test). Raw quantification data are included in Figure 4-Source Data 1.

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. ***p < 0.001. Scale bar: 100 μm.

Age-dependent loss of TH+ DA terminals in the striatum of LRRK cDKO mice

To determine whether loss of DA neurons in the SNpc is accompanied with loss of DA terminals in the striatum, we performed TH immunostaining and quantified TH immunoreactivity in the striatum of LRRK cDKO and littermate control mice at the ages of 15 and 24 months (Figure 5A). Quantitative analysis showed normal levels of TH immunoreactivity in the striatum of cDKO mice at 15 months of age but reduced levels of TH immunoreactivity in the striatum of cDKO mice at 24 months of age (−19%, p = 0.0215, unpaired Student’s t-test), suggesting an age-dependent loss of TH+ dopaminergic terminals in the striatum (Figure 5B).

Age-dependent loss of TH+ DA terminals in the striatum of LRRK cDKO mice

(A) Representative TH immunostaining images in the striatum of LRRK cDKO mice and littermate controls at the ages of 15 and 24 months.

(B) Quantification of TH immunoreactivity in the striatum of LRRK cDKO and control mice shows similar TH immunoreactivity at the age of 15 months (p = 0.8766, unpaired Student’s t-test), but there is a significant decrease in TH immunoreactivity in the striatum of LRRK cDKO mice at 24 months of age (- 19%, p = 0.0215), compared to controls.

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. *p < 0.05. Scale bar: 1 mm.

Unaffected TH+ noradrenergic neurons in the LC of LRRK cDKO mice

Previously, we reported the reduction of noradrenergic neurons in the locus coeruleus (LC) of LRRK DKO mice (30). We therefore quantified TH+ noradrenergic neurons in the LC of LRRK cDKO mice at the age of 24 months. The number of TH+ cells in the LC is similar between cDKO and littermate controls (Control: 3,418 ± 86, cDKO: 3,350 ± 99, p = 0.6110, Student’s t-test; Figure 6A, B). Further examination of the GFP reporter line crossed with DAT-Cre showed the lack of Cre-mediated recombination in the LC (Figure 6C), further support the cell intrinsic nature of DA neuron loss in the SNpc of aged LRRK cDKO mice.

Normal number of TH+ noradrenergic neurons in the LC of LRRK cDKO mice

(A) Representative images of TH+ noradrenergic neurons in the locus coeruleus (LC) of LRRK cDKO mice and littermate controls at 24 months of age.

(B) Quantification of TH+ cells shows similar number of TH+ noradrenergic neurons in the LC of LRRK cDKO mice (3,350 ± 99) and controls (3,418 ± 86, p = 0.6110, unpaired Student’s t-test).

(C) Top: Immunostaining of TH and GFP in the LC of DAT-Cre/+; Rosa26-CAG-LSL-ZsGreen1/+ mice at 2 months of age. There is no GFP+ (green) cell in the LC, indicating that DAT-Cre is not expressed in the LC. Bottom: Higher power views of the boxed areas.

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. Scale bar: 100 μm.

Quantitative EM analysis of the SNpc in LRRK cDKO mice

We previously reported age-dependent increases of electron-dense autophagic and autolysosomal vacuoles as well as the presence of large lipofuscin granules in the surviving SNpc neurons of LRRK DKO mice beginning at 10 months of age (30, 31). To determine whether selective inactivation of LRRK in DA neurons similarly results in the accumulation of electron-dense vacuoles in the SNpc, we performed EM analysis in the SNpc of LRRK cDKO mice and littermate controls at the age of 25 months (Figure 7). We observed various electron-dense double membrane autophagosomes and single membrane autolysosomes as well as lipofuscin granules composed of lipid-containing residues of lysosomal digestion in the SNpc of LRRK cDKO and littermate control mice (Figure 7C-F). Interestingly, the number of electron-dense vacuoles in the SNpc is similar between LRRK cDKO mice and littermate controls at the age of 25 months (Control: 6.72 ± 0.43, cDKO: 6.99 ± 0.52, p = 0.6839, Student’s t-test; Figure 7G). We also found no significant difference in the area of electron-dense vacuoles in the SNpc between LRRK cDKO and littermate controls (Control: 4.43 ± 0.44 μm2; cDKO: 4.60 ± 0.49 μm2, p = 0.8048; Figure 7G). The difference in accumulation of electron-dense vacuoles in the SNpc between germline DKO mice and DA neuron-restricted cDKO suggests that LRRK in non-DA neurons, possibly microglia, may play a more prominent role in the regulation of the autophagy-lysosomal pathway.

Unchanged number of electron-dense vacuoles in the SNpc of LRRK cDKO mice

(A, B) Representative EM images showing electron-dense vacuoles (arrowheads) in SNpc neurons of cDKO mice and littermate controls at the age of 25 months.

(C-F) Higher power views showing various electron-dense vacuoles, autolysosomes (single arrows), autophagosomes (double arrows), and lipid-containing vacuoles (asterisks) in SNpc neurons of littermate control (C, D) and cDKO (E, F) mice.

(G) Left: The average number of electron-dense vacuoles (>0.5 μm in diameter) in the SNpc neuronal profiles per mouse is not significantly different between LRRK cDKO mice and littermate controls at the age of 25 months (Control: 6.72 ± 0.43; cDKO: 6.99 ± 0.52, p = 0.6839, unpaired two-tailed Student’s t-test). Right: The total area of electron-dense vacuoles (>0.5 μm in diameter) in the SNpc neuronal profiles per mouse is similar between LRRK cDKO and littermate controls (Control: 4.43 ± 0.44 μm2; cDKO: 4.60 ± 0.49 μm2, p = 0.8048).

The value in parentheses indicates the number of mice (left) and neuron profiles (right) used in the quantification. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. Scale bar: 1 μm.

Enhanced microgliosis in the SNpc of LRRK cDKO mice

To determine whether selective inactivation of LRRK in DA neurons results in elevated microgliosis in the SNpc of LRRK cDKO mice. We performed immunohistochemical analysis of Iba1, which labels microglia, and TH, which marks DA neurons and processes, thus showing the boundary of the SNpc (Figure 8A-C). We found that the number of Iba1+ immunoreactive microglia is significantly increased in the SNpc of LRRK cDKO mice at 15 months of age (2541 ± 193), compared to controls (1737 ± 83; F1,45 = 102.6, p < 0.0001, two-way ANOVA with Bonferroni’s post hoc multiple comparisons, p = 0.0017; Figure 8A, D). The number of Iba1+ immunoreactive microglia in the SNpc of LRRK cDKO is further increased compared to controls at the age of 20 (Control: 2426 ± 68, cDKO: 3639 ± 127, p < 0.0001; Figure 8B, D) and 25 months (Control: 2640 ± 187, cDKO: 4089 ± 100, p < 0.0001; Figure 8C, D). These results show that despite the selective inactivation of LRRK in DA neurons of LRRK cDKO mice, microgliosis accompanies DA neuronal loss in the SNpc.

Elevated microgliosis in the SNpc of LRRK cDKO mice

(A-C) Representative images of Iba1+ microglia (red, marked by yellow arrowheads) and TH+ dopaminergic neurons (green) in the SNpc of LRRK DKO mice and controls at 15 (A), 20 (B), and 24 (C) months of age.

(D) Quantification of Iba1+ microglia shows significant increases in the number of Iba1+ microglia in the SNpc of LRRK cDKO mice, compared to control mice at the ages of 15 months (Control: 1737 ± 83, cDKO: 2541 ± 193; F1,45 = 102.6, p < 0.0001, p = 0.0017, two-way ANOVA with Bonferroni’s post hoc multiple comparisons), 20 months (Control: 2426 ± 68, cDKO: 3639 ± 127, p < 0.0001), and 24 months (Control: 2640 ± 187, cDKO: 4089 ± 100, p < 0.0001). Raw quantification data are included in Figure 8-Source Data 1.

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. **p < 0.01, ****p < 0.0001. Scale bar: 100 μm.

Impaired motor coordination of LRRK cDKO mice

To determine whether LRRK cDKO mice show motor deficits, we performed behavioral analysis of LRRK cDKO and littermate controls at the ages of 10 and 22 months using two versions of the elevated beam walk test with varying width of the beam (Figure 9). LRRK cDKO mice at 10 months of age displayed significantly more hindlimb slips/errors (4.4 ± 0.5) and longer traversal time (7.3 ± 0.3) in the 10 mm beam walk test, relative to littermate controls, which exhibited fewer slips (2.0 ± 0.3, p = 0.0005, unpaired two-tailed Student’s t-test) and shorter traversal time (5.8 ± 0.4, p = 0.0075; Figure 9A). In the 20 mm beam walk, which is less challenging than the narrower beam walk, both LRRK cDKO (1.5 ± 0.2) and littermate controls (1.0 ± 0.2, p = 0.0733) at 10 months of age performed well with few hindlimb slips and with no difference between LRRK cDKO and control mice. In addition, the traversal time is similar between LRRK cDKO (5.2 ± 0.3) and control littermates (5.1 ± 0.4, p = 0.9796; Figure 9A). These results show that LRRK cDKO mice at 10 months of age already exhibit deficits in motor coordination. However, in the pole test LRRK cDKO and control mice showed similar turning time (Control: 1.2 ± 0.1; cDKO: 1.5 ± 0.1; p = 0.1219) and descending time (Control: 4.7 ± 0.1; cDKO: 4.7 ± 0.2; p = 0.8620; Figure 9B). The LRRK cDKO and control mice at 10 months of age exhibit similar body weight (Figure 9C).

Impairment of motor coordination in LRRK cDKO mice

(A) In the 10 mm beam walk test, compared to control mice, LRRK cDKO mice at 10 months of age exhibit markedly more hindlimb slips (Control: 2.0 ± 0.3; cDKO: 4.4 ± 0.5; p = 0.0005, unpaired two-tailed Student’s t-test) and longer traversal time (Control: 5.8 ± 0.4; cDKO: 7.3 ± 0.3; p = 0.0075). In the less challenging 20 mm beam walk test, there is no significant difference in the number of hindlimb slips (Control: 1.0 ± 0.2; cDKO: 1.5 ± 0.2; p = 0.0733) and traversal time (Control: 5.1 ± 0.4; cDKO: 5.2 ± 0.3; p = 0.9796) between LRRK cDKO and control mice.

(B) In the pole test, LRRK cDKO and control mice at 10 months of age display similar turning time (Control: 1.2 ± 0.1; cDKO: 1.5 ± 0.1; p = 0.1219) and descending time (Control: 4.7 ± 0.1; cDKO: 4.7 ± 0.2; p = 0.8620).

(C) LRRK cDKO (31.1 ± 1.3) and control (32.0 ± 1.5; p = 0.6410) mice at 10 months of age show similar body weight.

(D) LRRK cDKO mice and control mice at 22 months of age in the 10 mm beam walk test show similar hindlimb slips (Control: 12.0 ± 3.0; cDKO: 13.8 ± 3.3; p = 0.7022) and traversal time (Control: 15.3 ± 2.0; cDKO: 16.2 ± 2.8; p = 0.8139). In the 20 mm team walk test, there is also no difference in hindlimb slips (Control: 4.8 ± 1.6; cDKO: 4.1 ± 1.0; p = 0.7142) and traversal time (Control: 8.7 ± 1.6; cDKO: 6.4 ± 0.7; p = 0.2223) between LRRK cDKO and control mice.

(E) In the pole test, LRRK cDKO mice at 22 months of age exhibit similar turning time (Control: 1.7 ± 0.2; cDKO: 2.8 ± 0.7; p = 0.1184) and descending time (Control: 6.6 ± 0.7; cDKO: 5.4 ± 0.3; p = 0.1413) compared to control mice.

(F) LRRK cDKO mice (34.8 ± 1.5) have similar body weight as control mice (39.6 ± 2.4; p = 0.1194).

The number in parentheses indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. **p < 0.01, ***p < 0.001. Raw behavior data are included in Figure 9-Source Data 1.

We then performed the beam walk test in naïve LRRK cDKO and control mice at 22 months of age (Figure 9D). The aged LRRK cDKO and control mice performed similarly in hindlimb errors (p = 0.7022) and traversal time (p = 0.8139) in 10 mm beam walk test. The hindlimb slips (Control: 12.0 ± 3.0; cDKO: 13.8 ± 3.3) and traversal time (Control: 15.3 ± 2.0; cDKO: 16.2 ± 2.8) of both genotypic groups were much higher, compared to the respective younger mice (p <0.0001; Figure 9D). Moreover, higher percentage of the aged mice (2-3 out of 8-9 per genotypic group) failed the test, compared to the younger mice (0-2 mice out of 18-20 per genotypic group, Figure 9A, D). These results show that aged mice of both genotypes perform poorly in the challenging narrow beam walk test, suggesting that it is not optimal to reveal subtle differences in motor coordination between the genotypic groups. The performance of LRRK cDKO and control mice at 22 months of age in the 20 mm beam walk test was also not significantly different with similar hindlimb errors (p = 0.7142) and traversal time (p = 0.2223; Figure 9D). In the pole test, LRRK cDKO mice and littermate controls also displayed similar turning time (p = 0.1184) and descending time to their home cage (p = 0.1413; Figure 9E) as well as body weight (Figure 9F).

Discussion

LRRK2 mutations represent ∼5% familial and 1-2% of sporadic PD cases, and the G2019S mutation is the most common mutation but with higher age of onset and incomplete penetrance (74% at age 79) (36). Despite the importance of LRRK2 in PD pathogenesis, whether LRRK2 mutations cause PD via a loss-of-function or a gain-of-function mechanism remains highly debated and unresolved. Neither LRRK2-deficient mice nor LRRK2 R1441C and G2019S KI mice develop dopaminergic neurodegeneration (25, 37, 38). It was proposed that the lack of brain phenotypes in LRRK2-/- mice might be due to the presence of LRRK1 (25, 26), which is broadly expressed in the brain including the midbrain (https://www.proteinatlas.org/ENSG00000154237-LRRK1/brain). Indeed, germline deletions of LRRK2 and LRRK1 results in age-dependent loss of DA neurons and increases of apoptosis and microgliosis in the SNpc of LRRK DKO mice (30, 31). However, the pleiotropic roles of LRRK1 and LRRK2 in the kidney (25, 26, 29), lung (28), and bone (39, 40), and the observed earlier mortality and lower body weight in LRRK DKO mice raised the possibility that DA neurodegeneration in LRRK DKO mice may be due to poor health.

In this study, we address this important question through the development of floxed LRRK1 and LRRK2 mice (Figure 1, Figure1-figure supplement 1-10), which permit dissection of cell intrinsic function of LRRK in DA neurons. We generated DA neuron-specific LRRK cDKO mice using the DAT-Cre KI allele (32) to delete floxed LRRK1 and LRRK2 regions selectively in DA neurons. Indeed, using an EGFP reporter line, we found that DAT-Cre drives Cre-mediated recombination in almost all DA neurons in the SNpc (Figure 2). Unlike the LRRK germline DKO mice (30), DA neuron-specific LRRK cDKO mice exhibit normal body weight and mortality during mouse lifespan (Figure 2). Importantly, LRRK cDKO mice develop age-dependent, progressive DA neurodegeneration, as evidenced by the normal number of DA neurons at the age of 15 months and the progressive reduction of DA neurons in the SNpc at the ages of 20 and 24 months (Figure 3), accompanied with increases of apoptotic DA neurons in the SNpc (Figure 4) and decreases of DA terminals in the striatum (Figure 5), whereas the number of noradrenergic neurons in the LC is unchanged, consistent with the lack of Cre-mediated recombination in the LC by DAT-Cre (Figure 6). Interestingly, quantitative EM analysis showed similar numbers of electron-dense vacuoles in the SNpc of LRRK cDKO and control mice at 25 months of age (Figure 7), in contrast to age-dependent increases of vacuoles in the SNpc of LRRK DKO mice (30, 31). Lastly, DA neurodegeneration in the SNpc of LRRK cDKO mice is accompanied with elevated microgliosis (Figure 8) and is preceded by impaired motor coordination (Figure 9).

The toxic gain-of-function pathogenic mechanism was initially proposed based on in vitro kinase assays using recombinant mutant LRRK2, which showed that R1441C and G2019S resulted in increases of autophosphorylation and phosphorylation of a generic substrate (41). Subsequent biochemical studies further reported that LRRK2 mutations (e.g. R1441C/G, G2019S) enhanced kinase activity, leading to elevated levels of pSer1292-LRRK2 (42), pSer935-LRRK2 (43, 44), and pT73-Rab10 and pS106-Rab12 (45). However, most transgenic mice overexpressing mutant LRRK2 did not produce neurodegeneration or DA neuron loss (46-54), though two reports showed that overexpression of LRRK2 G2019S but not R1441C resulted in DA neurodegeneration (55, 56). Thus, further studies are needed to elucidate the differences of these seemingly conflicting transgenic findings and to understand the differences between the R1441C and G2019S mutations.

The loss-of-function pathogenic mechanism was supported by mouse knockout findings showing that germline inactivation of LRRK results in age-dependent loss of DA neurons in the SNpc and DA terminals in the striatum (30, 31). The current study demonstrated a cell intrinsic role of LRRK in support of DA neuron survival in the SNpc of the aging brain, providing further genetic evidence of an essential, cell autonomous requirement of LRRK in DA neuron survival. However, highly penetrant PD-linked mutations are dominantly-inherited missense mutations in LRRK2 (6), and there have been no reported deletion mutations that are linked to or causal of familial PD, though variants in LRRK2 and LRRK1 were reported to be associated with sporadic PD but not linked to familial PD. One plausible explanation is that LRRK2 mutant may act in a dominant negative manner, inhibiting the activity of the wild-type LRRK2. This notion is supported by structural studies showing that the GTPase domain forms a dimer, and that the R1441/C/G/H and I1371V mutations destabilize dimer formation and decrease GTPase activity (22), and is also consistent with the high resolution crystal structure of LRRK2’s WD40 domain (57). Earlier studies initially suggested that LRRK2 functions as a dimer (58-60), and that LRRK2 and LRRK1 could form a heterodimer (61). Furthermore, LRRK2 G2385R variant, a risk factor for PD (62-65), has been reported as a partial loss-of-function mutation (66, 67). Future studies are needed to differentiate possibilities between a toxic gain-of-function pathogenic mechanism versus a loss-of-function mechanism, in which loss-of-function mutations exert a dominant negative effect further impairing LRRK2 activity of the wild-type protein encoded by the LRRK2 wild-type allele, as the gain-of-function or loss-of-function pathogenic mechanism underlying LRRK2 mutations pivots opposing development of LRRK2-based PD therapies.

Finally, it was surprising to find similar accumulation of autophagic and autolysosomal vacuoles in SNpc neurons between LRRK cDKO and control mice. This suggests that non-DA neurons (e.g. microglia), in which LRRK expression is unaffected in DA neuron-specific LRRK cDKO mice but is absent in germline DKO mice, may contribute to this phenotypic difference. Development of microglia- and/or astrocyte-specific LRRK cDKO mice would permit further dissection of cell autonomous and non-cell autonomous roles of LRRK in DA neuron survival and determine whether glial LRRK plays a more important role in the regulation of the autophagy-lysosomal pathway and protein turnover. It would also be interesting to test whether reducing LRRK expression or function affects accumulation and aggregation of human mutant α-synuclein and tau in the aging brain, as both Lewy bodies and tauopathies are associated with PD patients carrying LRRK2 mutations (3).

Materials and methods

Key Resource Table

Mice

All animal use was approved by the IACUC committees of Harvard Medical School and Brigham and Women’s Hospital, and conformed to the USDA Animal Welfare Act, PHS Policy on Humane Care and Use of Laboratory Animals, the “ILAR Guide for the Care and Use of Laboratory Animals” and other applicable laws and regulations. Mice were housed in constant humidity- and temperature-controlled rooms and maintained on a 12 hr light/dark cycle and were given standard rodent chow and water. Mice of both sexes at multiple ages, from 2 months to 25 months, were used. Floxed LRRK1 (fLRRK1) and the resulting germline deleted LRRK1Δ alleles, and floxed LRRK2 (fLRRK2) and the resulting germline deleted LRRK2Δ alleles were generated and thoroughly validated at the genomic DNA, mRNA, and protein levels. DAT-IRES-Cre (The Jackson Laboratory, IMSR_JAX:006660), ACT-FLPe (IMSR_JAX:005703), CMV-Cre transgenic (IMSR_JAX:006054), and Rosa26-CAG-LSL-ZsGreen1 mice (IMSR_JAX:007906) used in the current study were previously characterized and reported (32-35). LRRK1/LRRK2 cDKO mice (fLRRK1/fLRRK1; fLRRK2/fLRRK2; DAT-IRES-Cre/+) and littermate controls (fLRRK1/fLRRK1; fLRRK2/fLRRK2) were obtained from multiple breeding cages of fLRRK1/fLRRK1; fLRRK2/fLRRK2 and cDKO mice. All mice were maintained on the C57BL6 and 129 hybrid genetic background (F1: IMSR_JAX:101043). All phenotypic analyses were performed in a genotype blind manner, as previously described (68).

Generation of targeted and floxed LRRK1 and LRRK2 alleles

The generation and validation of the LRRK1 and LRRK2 targeting vectors, the targeted, floxed, and deleted LRRK1 and LRRK2 alleles by Southern, Northern, RT-PCR, and the sequences of the floxed LRRK1 and LRRK2 alleles are included in Figure 1-figure supplement 1-11.

LRRK1

To generate the LRRK1 targeting vector, we first PCR amplified the left middle homologous region (2,079 bp) containing partial intron 26, exon 27, and partial intron 27 of LRRK1 from mouse BAC DNA (clone RP23-213J23, BACPAC Resources Center) using primers, P3 and P4. The PCR fragment was subcloned into the pGEM-T vector (A1360, Promega) to generate pLM1 (for details, see Figure 1-figure supplement 1). The right middle homologous region (3,403) containing LRRK1 genomic region from partial intron 27 to partial intron 29 was amplified by PCR using primers, P5 and P6, and then subcloned into the pGEM-T vector (pLM2). The left middle homologous region containing pGEM-T plasmid was then digested with NotI and KpnI (endogenous KpnI site in the intron 27) and subcloned into the NotI and KpnI sites of the right middle homologous region containing pGEM-T plasmid (pLM3). After that, the middle homologous region (5.5 kb), from partial intron 26 to partial intron 29, was released by NotI and SacII digestions and was blunted by Klenow, and then subcloned into the SmaI site of PGKneoF2L2DTAvector (#13445, addgene) to generate the middle homologous region-PGKneoF2L2DTA plasmid (pLM4).

The left homologous region (2,016 bp) containing partial intron 25, exon 26, and partial intron 27 of LRRK1 was PCR amplified from mouse BAC DNA (clone RP23-213J23, BACPAC Resources Center) using primers, P1 and P2. The PCR fragment was digested with SacII (introduced by P1) and NotI (introduced by P2 along with HindIII), and was subcloned into the SacII and NotI sites of the PGKneoF2L2DTA vector (pLM5). The right homologous region (3,131 bp) containing partial intron 29, exon 30, and partial intron 30 of LRRK1 was PCR amplified from mouse BAC DNA (clone RP23-213J23) using primers, P7 and P8. The PCR fragment was digested with SalI (introduced by P7) and HindIII (introduced by P8), and was subcloned into the SalI and HindIII sites of the PGKneoF2L2DTA vector (pLM6).

The left homologous region in the PGKneoF2L2DAT vector (pLM5) was digested with SacII and NotI, and subcloned into the SacII and NotI sites of the right homologous region containing pGKneoF2L2DTA plasmid (pLM6) to generate pLM7. Finally, the loxP-middle homologous region-PGK-Neo-loxP fragment was released from pLM4 by NotI and SalI digestions, and subcloned into the NotI and SalI sites of pLM7 to generate the final target vector (pLM8), which contains two loxP sites (intron 26 and intron 29). Upon Cre-mediated recombination, the endogenous LRRK1 genomic sequence flanked by the 5’ loxP site (1,288 bp upstream of LRRK1 exon 27) to the 3’ loxP site (1,023 bp downstream of exon 29) is excised. The PGK-neo cassette flanked by two FRT sites is under the control of the mouse phosphoglycerate kinase 1 (PGK) promoter and contains the bovine growth hormone polyA signal. To enhance the ratio of ES cells carrying homologous recombination events instead of random insertion of the targeting vector (69), the negative selection PGK-DTA cassette, which encodes diphtheria toxin A chain, was also included in the LRRK1 targeting vector.

The LRRK1 targeting vector was linearized by XhoI digestion, and then electroporated into MKV6.5 embryonic stem (ES) cells (a gift from R. Jaenisch), which were derived from B6/129 F1 mice (The Jackson Laboratory, IMSR_JAX:101043). G418 was applied to the culture at 150 μg/ml 24 h later, and after 6 days of G418 selection, the surviving ES clones (480) were picked and screened by Southern analysis using HindIII digestion of genomic DNA followed by hybridization with the 5’ external and 3’ external probes to confirm proper recombination events in the 5’ and 3’ homologous regions, respectively. 25 ES cell clones were confirmed to carry the proper homologous recombination events at the 5’ homologous region, giving rise to the 17.0 kb and the 4.8 kb bands, which represent the wild-type and the targeted allele, respectively (Figure 1D), and the 3’ homologous region, giving rise to the 17.0 kb and the 14.1 kb bands, which represent the wild-type and the targeted allele, respectively. We then expanded the selected 4 ES cell clones (3A11, 3D8, 3H1, and 3H6) and further verified by Southern using the 5’ and 3’ external, and neo probes. Two ES cell clones (3D8 and 3H6) were transfected with pCAGGS-flpE-puro (#20733, addgene) to delete the PGK-neo cassette by FLP recombinase. Three resulting ES cell clones (3D8C5, 3D8E5, and 3H6G7) were confirmed by Southern analysis using the 5’ and 3’ external probes and the neo probe to confirm the floxed LRRK1 allele by the deletion of the PGK-neo cassette.

LRRK2

To generate the LRRK2 targeting vector, the left homologous region (2,579 bp) containing the LRRK2 promoter region was PCR amplified from mouse BAC DNA (clone RP23-526A2, BACPAC Resources Center) using primers, P9 and P10, and subcloned into the pGEM-T vector to generate pLRRK2#1 (for details see Figure 1-figure supplement 4). The BamHI-loxP-NheI-SpeI fragment (56 bp), which was generated by annealing two complementary oligos, P39 and P40, was introduced to the BamHI and SpeI sites of pLRRK2#1 to generate pLRRK2#2. The middle homologous region, containing genomic sequences from the promoter region to partial intron 2, was PCR amplified from mouse BAC DNA (clone RP23-526A2) using primers, P11 and P12, digested with XbaI and NotI, and subcloned into the XbaI and NotI sites of the pGEM-T vector to generate pLRRK2#3, which was digested with XbaI and NotI, and subcloned into the XbaI and NotI sites of pLRRK2#2 to generate pLRRK2#4.

The right homologous region (3,503 bp), which contains partial intron 1, exon 2, and partial intron 2, was PCR amplified from mouse BAC DNA (clone RP23-526A2) using primers, P13 and P14, and the PCR fragment was subcloned into the EcoRV site of pBluescript II KS (+) vector (212207, Agilent) to generate pLRRK2#5. The right homologous region in pLRRK2#5 was digested with BamHI followed by Klenow, and then digested with ClaI. The released fragment was subcloned into the EcoRV and ClaI sites of the PGKneoF2L2DTA vector to generate pLRRK2#6. LFNT-tk/pBS plasmid (a gift from S. Tonegawa) was digested with SacII (followed by Klenow) and NotI (followed by SspI digestion to make it easier to isolate the NotI-SacII/KN fragment) to release the loxP-FRT-PKG-neo-loxP-FRT fragment (2,928 bp), and then subcloned into the XbaI (followed by Klenow) and NotI sites of pLRRK2#6 to generate pLRRK2#7. Finally, pLRRK2#4 containing the left-loxP-middle homologous regions was digested with EcoRV and NotI, and subcloned into AleI and NotI sites of pLRRK2#7 containing the loxP-FRT-PKG-neo-loxP-FRT-right homologous region to generate the final targeting vector, which contains the loxP sites in the promoter region (1,768 bp upstream of the transcription initiation site) and in LRRK2 intron 2 (878 bp downstream of exon 2). Upon Cre-mediated recombination, the floxed endogenous LRRK2 genomic sequences from the 5’ loxP site (1,768 bp upstream of LRRK2 transcription) to the 3’ loxP site in intron 2 (878 bp downstream of exon 2) are excised. The negative selection PGK-DTA cassette is also included in the LRRK2 targeting vector.

The LRRK2 targeting vector was linearized by AhdI digestion, and then electroporated into MKV6.5 ES cells. G418 was applied to the culture at 150 ug/ml 24 h later, and after 6 days of G418 selection, the surviving ES clones (480) were picked and screened by Southern analysis using NheI digestion of genomic DNA followed by hybridization with the 5’ external probe (753 bp, PCR amplified using P19 and P20). Southern analysis confirmed that two ES cell clones (N24 and N34) carry the proper recombination event in the 5’ homologous arm, giving rise to 11.6 kb and 3.6 kb bands, which represent the wild-type and targeted alleles, respectively (Figure 1F), followed by genomic PCR confirmation for the proper recombination in the right arm.

Three (3D8C5, 3A11, 3H6) and one (N24) ES clones for LRRK1 and LRRK2, respectively, were microinjected into C57BL/6 mouse blastocysts to generate chimera mice, which were bred with B6/129 F1 mice to produce heterozygous floxed LRRK1 and targeted LRRK2 mice. Floxed LRRK1 mice were confirmed by Southern analysis using the 5’ and 3’ external probes (Figure 1D). Targeted LRRK2 mice were confirmed by Southern (data shown in Figure 1-figure supplement 6; NheI digestion followed by hybridization with the 5’ external probe and SphI digestion followed by hybridization with the 3’ external probe). Targeted LRRK2 mice were then bred with Actin-FLP deleter mice (IMSR_JAX:005703) (33) to generate floxed LRRK2 mice, which were confirmed by Southern analysis using the 5’ and 3’ external probes following NheI digestion (Figure 1F). Heterozygous fLRRK1/+ and fLRRK2/+ mice were crossed with each other to obtain homozygous single floxed mice (fLRRK1/fLRRK1 and fLRRK2/fLRRK2) and double floxed mice (fLRRK1/fLRRK1; fLRRK2/fLRRK2).

Generation of deleted LRRK1/LRRK2 alleles and DA neuron-specific cDKO mice

In order to ensure that Cre-mediated deletion of the floxed LRRK1 and LRRK2 alleles results in null alleles, we crossed floxed LRRK1 and LRRK2 mice with germline deleter, CMV-Cre transgenic mice (B6.C-Tg(CMV-cre)1Cgn/J; IMSR_JAX:006054) (34), to generate LRRK1 deleted (Δ/Δ) mice (by removing exons 27-29) and LRRK2 deleted (Δ/Δ) mice (by removing the promoter region and exons 1-2) for further molecular characterization. To generate DA neuron-specific LRRK1/LRRK2 cDKO mice, we used DAT-Cre KI mice (B6.SJL-Slc6a3tm1.1(cre)Bkmn/J; IMSR_JAX:006660), which express Cre recombinase under the control of the endogenous DAT promoter (32). We crossed double floxed mice with DAT-Cre mice to generate LRRK cDKO mice (fLRRK1/fLRRK1; fLRRK2/fLRRK2; DAT-Cre/+). LRRK cDKO and littermate control mice used in the phenotypic analysis were obtained by crossing fLRRK1/fLRRK1; fLRRK2/fLRRK2; DAT-Cre/+ with fLRRK1/fLRRK1; fLRRK2/fLRRK2 mice. We only used cDKO and control mice that carry all floxed LRRK1 and LRRK2 alleles for phenotypic analysis.

Southern analysis

For the identification and validation of the targeted and floxed LRRK1 alleles, we used the 5’ external, 3’ external, and neo probes. Genomic DNA from ES cells or mouse tails was digested with HindIII. The 5’ external probe (377 bp), which is 839 bp upstream of the 5’ homologous region, was PCR amplified from mouse BAC DNA (clone RP23-213J23) using primers P15 and P16. The 3’ external probe (305 bp), which is 2,736 bp downstream of the 3’ homologous region, was PCR amplified from mouse BAC DNA (clone RP23-213J23) using primers P17 and P18. The neo probe (363 bp) was PCR amplified from pSoriano plasmid using primers P23 and P24. Following HindIII digestion, the presence of the 17.0 kb product using either the 5’ or 3’ external probe represents the wild-type allele, whereas the 4.8 kb (the 5’ external probe) and the 12.2 kb (the 3’ external probe) products represent the floxed LRRK1 allele. Genomic DNA digested by HindIII and hybridized with the neo probe further confirmed the wild-type and floxed alleles (no band) and the targeted LRRK1 allele (14.1 kb).

For the identification and validation of the targeted and floxed LRRK2 alleles, we used the 5’ and 3’ external probes as well as the neo probe. Genomic DNA from ES cells or mouse tails was digested with NheI or SphI followed by hybridization with the 5’ or 3’ external probe or the neo probe. The 5’ external probe (753 bp), which is 84 bp upstream of the 5’ homologous region, was PCR amplified from mouse BAC DNA (clone RP23-526A2) using primers P19 and P20. The 3’ external probe (622 bp), which is 38 bp downstream of the 3’ homologous region, was PCR amplified from mouse BAC DNA (clone RP23-526A2) using primers P21 and P22. Following NheI digestion, the presence of the 11.5 kb product using either the 5’ or 3’ external probe represents the wild-type allele, whereas the 3.6 kb (the 5’ external probe) and the 5.2 kb (the 3’ external probe) products represent the floxed LRRK2 allele.

PCR genotyping

Genomic PCR was performed to determine the presence of the deleted, the floxed, and/or the wild-type alleles. For LRRK1, the following primers were used: P25 (5’-ATTGGTCTTTGAAGAGACAGCATCTGG, forward primer, 392 nt downstream of exon 26), P26 (5’-TTTCCCTGAGGTGGAGAAGTGACTGG, reverse primer, 567 nt downstream of exon 26), and P27 (5’-TCACGTCGTCTAAGCCTCCT, reverse primer, 1,218 nt downstream of exon 29). The PCR products from P25 and P26 are 266 bp and 405 bp, which represent the wild-type and the fLRRK1 alleles, respectively, whereas the PCR product from P25 and P27 is 583 bp, which represents the deleted LRRK1 allele.

For LRRK2, the following primers were used: P28 (5’-CTTCCTCAGAAGTTAGGTAAACATTGAGTG, forward primer, 2,069 nt upstream of exon 1), P29 (5’-CTAAGTGACACCGTGTTTCCAAAGTC, reverse primer, 1,739 nt upstream of exon 1), and P30 (5’-GGAAAGTTTCACAATTGGAAAAATAAAAATATTTACTGCAGATA, reverse primer at 2,848 nt downstream of exon 2). The PCR products from P28 and P29 are 305 bp and 367 bp, representing the wild-type and the fLRRK2 alleles, respectively, whereas the PCR product from P28 and P30 is 587 bp, which represents the deleted LRRK2 allele.

For DAT-IRES-Cre, the following primers were used: JKM1823 (5’-TGGCTGTTGGTGTAAAGTGG, forward primer at exon 16 and 3’-UTR), JKM1824 (5’-GGACAGGGACATGGTTGACT, reverse primer at 3’-UTR), and JKM1825 (5’-CCAAAAGACGGCAATATGGT, reverse primer at IRES sequence). The PCR product from JKM1823 and JKM1824 is 264 bp, which represents the wild-type allele, whereas the PCR product from JKM1823 and JKM1825 is 152 bp, which represents the DAT-IRES-Cre KI allele.

Northern analysis

Total RNA was isolated from brains, kidneys, or lungs using TRI reagent (T9424, Millipore-Sigma) according to the manufacturer’s instruction. For the LRRK1 Northern analysis, polyA+ RNA was enriched from ∼500 μg total RNA using the Poly(A)Purist™ MAG Kit (AM1922, ThermoFisher) according to the manufacturer’s instruction. For the LRRK2, ∼10 μg of total RNA was used for Northern analysis. RNA was separated in formaldehyde agarose gel, and transferred into Amersham Hybond-nylon membrane (RPN303N, GE Healthcare). Probes were synthesized using Prime-It II random labeling kit (#300385, Agilent) and then used for membrane hybridization at 55°C overnight.

The cDNA probe specific for LRRK1 exons 2-3 (383 bp) was PCR amplified using primers P31 (5’-CAGGATGAGCGTGTGTCTGCAG) and P32 (5’-CCTTCTCCTGTGAGGATTCGCTCT). The cDNA probe specific for LRRK1 exons 27-29 (550 bp) was PCR amplified using primers P33 (5’-CTGGCCTACCTGCACAAGAA) and P34 (5’-CCTTCCCATCCCAGAACACC).

The cDNA probe specific for LRRK2 exons 1-5 (437 bp) was PCR amplified using primers P35 (5’-AGGAAGGCAAGCAGATCGAG) and P36 (5’-GGCTGAATATCTGTGCATGGC). The probe specific for GAPDH exons 5-7 (452 bp) was PCR amplified using primers P37 (5’-ACCACAGTCCATGCCATCAC) and P38 (5’-TCCACCACCCTGTTGCTGTA).

Hybridization was performed using α-32P-dCTP-labeled probes specific to each gene. Specific signals were detected by autoradiography with Hyperfilm (E3018, Amersham).

RT-PCR

Total RNA was isolated from brains, kidneys, or lungs using TRI reagent (T9424, Millipore-Sigma,) according to the manufacturer’s instruction. Approximately 1 μg of RNA was reverse-transcribed using Superscript III (18080093, Thermo Fisher Scientific) according to the manufacturer’s instructions. For RT-PCR analysis of LRRK1 transcripts in mice carrying the homozygous floxed or deleted alleles, we used primers P53 and P54 for exons 4-8 (714 bp), P57 and P58 for exons 11-17 (818 bp), P63 and P64 for exons 20-25 (922 bp), or P41 and P42 for exons 25-31 to confirm normal splicing of LRRK1 mRNA in fLRRK1/fLRRK1 mice and truncated LRRK1 transcripts lacking exons 27-29 in Δ/Δ mice (for details see Figure 1-figure supplement 9). For RT-PCR analysis of LRRK2 transcripts, we used primers P35 and P36 in exons 1-5 to confirm normal splicing of LRRK2 mRNA (437 bp) in fLRRK2/fLRRK2 mice and the absence of RT-PCR products in LRRK2 Δ/Δ mice (Figure 1-figure supplement 11). The identity of the PCR products was confirmed by sequencing.

Western analysis

Fresh tissues were collected and homogenized in an ice-cold stringent RIPA buffer (50 mM Tris-Cl (pH 7.6), 150 mM NaCl, 0.5 mM EDTA, 1% NP40, 0.5% sodium deoxycholate, 0.1% SDS, 1mM PMSF supplement with protease inhibitor cocktail (P8340, Sigma) and phosphatase inhibitor cocktail (P0044, Sigma)), followed by sonication. Homogenates were centrifuged at 14,000xg for 20 min at 4°C to separate supernatants (RIPA buffer-soluble fraction). An equal amount (10-40 μg per lane) of total proteins from each preparation were loaded and separated on NuPAGE gels (Invitrogen), and transferred to nitrocellulose membranes. The membranes were blocked in Intercept (TBS) Blocking Buffer (927-60001, Li-Cor) for 1 hr at room temperature, and incubated at 4°C overnight with specific primary antibodies. Primary antibodies used were rabbit anti-LRRK1 (ANR-101, Alomone Lab, RRID: AB_2756700), rabbit anti-LRRK2 (ab133474, abcam, RRID: AB_2713963), mouse anti-α-Vinculin (05-386, Millipore, RRID: AB_309711). Membranes were then incubated with dye-coupled secondary antibodies, goat anti-mouse IRdye680 (#925-68070, LI-COR, RRID: AB_2651128) or goat anti-rabbit IRdye800 (#925-32211, LI-COR, RRID: AB_2651127). Signals were quantified using the Odyssey Infrared Imaging System (LI-COR).

Histological analysis

Mice were anesthetized with ketamine (100 mg/kg) + xylazine (10 mg/kg) + acepromazine (3 mg/kg), and transcardially perfused with phosphate-buffered saline solution (PBS, pH 7.4) containing 0.25 g/L heparin (H3149, Sigma) and 5 g/L procaine (P9879, Sigma). Brains were post-fixed in 4% formaldehyde in PBS (pH 7.4) (15710, Electron Microscopy Sciences) at 4°C overnight and then processed for paraffin embedding following standard procedures. For frozen sections, post-fixed brains were immersed in a sucrose series solution (15% and 30% sucrose in PBS) at 4°C overnight for cryoprotection, and then brains were embedded in an OCT compound (4583, Sakura). Serial coronal sections (16 µm) of paraffinized brains or frozen brains were obtained using Leica RM2235 or Leica CM1860, respectively. Coronal sections containing the SNpc or LC were selected for immunohistochemical analysis.

Histological analyses were performed as described previously (30, 31, 68). Briefly, for DAB-derived TH-immunohistochemistry, coronal sections were deparaffinized, alcohol-dehydrated, and then subjected to permeabilization with a solution containing 0.1% Triton X-100 in TBS followed by antigen retrieval for 5 min in 10 mM sodium citrate buffer, pH 6.0. Endogenous peroxidase activity was quenched by incubating in 0.3% H2O2 in methanol. Sections were then blocked with a solution containing 5% normal goat serum (S-1000, Vector Laboratories) and 0.1% Triton X-100 in TBS for 1 hr at room temperature. After blocking, sections were incubated with the primary antibody, rabbit anti-TH (1:750, ab112, abcam, RRID: AB_297840) overnight at 4°C. Sections were washed three times in 0.1% Triton X-100 in TBS followed by 1 hr incubation with goat biotinylated anti-rabbit IgG secondary antibody (1:250, BA-1000, Vector Laboratories, RRID: AB_2313606) at room temperature and 30 min incubation with Vectastain Elite ABC reagent (PK-6100, Vector Laboratories) and then developed using chromogenic DAB substrate (SK-4100, Vector Laboratories).

For immunofluorescence staining of paraffin sections, coronal sections were deparaffinized, alcohol-dehydrated, and then subjected to permeabilization with a solution containing 0.1% Triton X-100 in PBS followed by antigen retrieval for 5 min in 10 mM sodium citrate buffer, pH 6.0 except those for cleaved-caspase3 immunostaining which were performed antigen retrieval for 10 min. Sections were then blocked with a solution containing 10% normal goat serum and 0.1% Triton X-100 in PBS for 1 hr at room temperature. After blocking, sections were incubated with primary antibodies overnight at 4°C. The primary antibodies used were mouse anti-TH (1:50, sc-25269, Santa Cruz, RRID: AB_628422), rabbit anti-NeuN (1:400, 12943S, Cell Signaling Technology, RRID: AB_2630395), rabbit anti-cleaved caspases-3 (1:150, 9661, Cell Signaling Technology, RRID: AB_2341188), or rabbit anti-Iba1 (1:500, 019-19,751, Wako, RRID: AB_839504). Sections were washed three times in 0.1% Triton X-100 in PBS followed by 1 hr incubation with fluorophore-conjugated secondary antibodies, goat anti-mouse IgG Alexa Fluor 488 (1:500, A11001, ThermoFisher, RRID: AB_2534069) and goat anti-rabbit IgG Alexa Fluor 555 (1:500, A32732, ThermoFisher, RRID: AB_2633281) at room temperature. Fluorescence images were taken (a stack of 2-3 confocal images spaced at 4 μm), projected with maximal intensity projection mode, and analyzed by FV1000 confocal microscope system (Olympus).

For immunofluorescence staining of cryopreserved sections, coronal brain sections were washed with PBS to rinse out OCT, and then blocked with a solution containing 5% normal goat serum and 0.1% Triton X-100 in TBS for 1 hr at room temperature. After blocking with 10% NGS, sections were incubated with primary antibodies, rabbit anti-GFP (1: 1000, ab290, abcam, RRID: AB_303395), and mouse anti-TH (1:50, sc-25269, Santa Cruz, RRID: AB_628422), overnight at 4°C. Sections were incubated with fluorophore-conjugated secondary antibodies, goat anti-rabbit IgG Alexa Fluor 488 (1:500, A11034, ThermoFisher, RRID: AB_2576217) and goat anti-mouse IgG Alexa Fluor 555 (1:250, A21424, ThermoFisher, RRID: AB_141780) for 1 hr at room temperature. Fluorescence images were taken and analyzed by FV1000 confocal microscope system (Olympus).

The number of GFP+ and GFP+/TH+ cells in the SNpc of DAT-IRES-Cre/+; Rosa26-CAG-LSL-ZsGreen1/+ reporter mice were quantified using 3 comparable coronal sections (16 µm in thickness, spaced 320 μm apart) per brain (n = 3 brains, one hemisphere). The percentage of GFP+ DA neurons in the SNpc was obtained by dividing the sum of GPF+/TH+ neurons by the sum of total TH+ neurons quantified.

Quantification of DA neurons in the SNpc

Quantification of TH+ DA neurons in the SNpc or LC was performed as previously described (30, 31, 68, 70). Briefly, TH+ neurons in the SNpc, which were marked based on morphological features as previously described (71), were quantified in every 10th serial coronal section (16 µm in thickness) throughout the SNpc (a total of 6-9 sections, spaced 160 µm apart). Total number of TH+ cells in the SNpc was calculated as follows: [total number of TH+ DA neurons quantified in all 6-9 sections] x 10 (every 10th section sampled) x 2 (both hemispheres). Total number of TH+ noradrenergic neurons in the LC (a total of 6-9 sections, spaced 80 µm apart) was calculated as follows: [total number of TH+ noradrenergic neurons quantified in all 6-9 sections] x 5 (every 5th section sampled) x 2 (both hemispheres).

DA neuron quantification was also performed independently by another investigator, also in a genotype blind manner, using stereological quantification of 25% of the SNpc area. The fractionator with 100 μm x 100 μm was set up in the SNpc, and TH+ DA neurons were counted using the optical dissector method with 50 μm x 50 μm sample box (25% of the total area). Total number of TH+ cells in the SNpc was calculated as follows: [total number of TH+ DA neurons quantified in sample boxes of all 6-9 sections] x 10 (every 10th section sampled) x 4 (1/4 area sampled: 50 x 50/100 x 100) x 2 (both hemispheres).

Quantification of NeuN+ neurons, apoptotic cells, Iba1+ microglia in the SNpc

NeuN+, active Caspase-3+, or Iba1+ cells in the SNpc, which was marked by TH immunoreactivity, were quantified using serial coronal sections (16 μm in thickness, every 10th section, a total of 6-9 sections per brain). Compared to DAB immunostaining followed by counting under the stereomicroscope, which captures TH-positive DA neurons in a single cell layer of the brain section, immunofluorescent staining picks up more cells (e.g. TH+ or NeuN+) in multiple layers of the brain section under the confocal microscope by the maximal intensity projection mode. Thus, more TH-positive DA neurons in the SNpc were captured and quantified under confocal microscopy compared to DAB staining. The total number of NeuN+ cells in the SNpc was calculated by multiplying the total number of NeuN+ cells in all sections counted] x 10 (every 10th section sampled) x 2 (both hemispheres). The total number of active Caspase-3+ or Iba1+ cells in the SNpc, which was marked by TH immunoreactivity, was calculated as follows: [total number of active Caspase-3+ or Iba1+ cells in all sections counted from one hemisphere] x 10 (every 10th section sampled).

Quantification of TH+ DA terminals in the striatum

For quantification of TH immunoreactivity in the striatum, we performed immunostaining using every 10th serial coronal sections (16 µm in thickness) throughout the entire striatum (a total of 12-15 sections, spaced 160 µm apart). The images of TH immunoreactivity in the striatum were captured under 2X objective lens (Olympus BX40, 8-bit RGB camera) using identical exposure time, sensitivity, brightness, contrast, and gamma (CellSens Entry Software) and then analyzed using the Fiji version of ImageJ, and the optical density was determined as previously described (31, 72, 73). The mean value of TH immunoreactivity in the striatum of control mice of each age group was set as 100%.

Quantitative EM analysis

The collection and quantification of the EM images were performed as described previously (30, 31). Mice were perfused with PBS containing 0.25 g/L heparin and 5 g/L procaine followed by a fixative solution containing 2.5% paraformaldehyde and 2.5% glutaraldehyde in 0.1 M sodium cacodylate buffer (pH 7.4) (#1549, Electron Microscopy Science). Brains were dissected and post-fixed overnight in a fixative solution at 4°C. The dissected tissues were trimmed to 1-2 mm3 cubes followed by osmication and uranyl acetate staining, dehydration in graded alcohol, and embedded in TABB 812 Resin (Marivac Ltd.) at the Harvard Medical School EM facility. 0.5 μm sections were stained with toluidine blue and viewed under the light stereomicroscope (Nikon Eclips E600) to find the SNpc area for EM viewing. Then adjacent sections were cut with 80 nm in thickness with the Leica Ultracut S microtome, picked up on formvar-carbon coated slot Copper grids, stained with 0.2% Lead Citrate, and viewed and imaged under the JEOL 1200x Electron Microscope. A minimum of 10 micrographs containing the entire cell body in the SNpc area were analyzed for each brain. The image was analyzed by using the Fiji version of ImageJ. The number of electron-dense autophagic and lysosomal vacuoles (>0.5 μm in diameter) in individual neuronal profiles was quantified. We previously calculated the diameter of electron-dense autophagic/lysosomal vacuoles by measuring the longest side manually (30, 31). In the current study, we used Feret’s Diameter (74) to accurately measure the longest distance between any of two points in the electron-dense autophagic/lysosomal vacuoles, resulting in a higher number of electron-dense vacuoles quantified. The number and the area of electron-dense vacuoles (>0.5 μm in diameter) in individual neuronal profiles was quantified, and the average number or area of vacuoles per mouse was calculated. Experiments were done in a genotype blind manner (after scarifying mice, the brain samples were coded and sent to the Harvard Medical School EM Core, where the images were captured, and the vacuoles were quantified by another independent experimenter).

Behavioral analysis

Naive LRRK cDKO mice and littermate controls at 10 and 22 months of age were used. Mice were acclimated in the behavior facility for a minimum of 7 days and were then individually handled daily for 5 consecutive days before testing. Mice were coded so the experimenter was unaware of their genotypes until the data analysis was complete. In the beam walk test, a Plexiglas beam of 100 cm in length (Plastic Zone), 20 mm or 10 mm in width, was raised 60 cm above a table, and safety bedding was placed under the beam to avoid any harm in case of falls. Mice were placed onto the starting point of the beam in bright light, and the time (in seconds) to reach their home cage on the other darker side of the beam (∼80 cm in distance) as well as the hindpaw slips (number of hindlimb errors) were recorded. Mice were trained three trials per day for 2 consecutive days to traverse the 20 mm beam (without the wire mesh) to their home cage. On the test day, mice were trained further with two additional trials on the 20 mm beam (without the wire mesh). Mice were then tested in two successive trials on the 20 mm beam (with the wire mesh) followed by two consecutive test trials on the 10 mm beam (with the wire mesh). All test trials were videotaped, and the travel time and the number of hindlimb errors were recorded. Between trials mice were placed in the home cage for 2 min to rest. Mice that fell off the beam during both trials or stalled on the beam for more than 120 sec during the test were excluded.

In the pole test, mice were placed on top of the pole (60 cm in height, 10 mm in diameter) with their head facing upward, and the base of the pole was placed in the home cage. Mice were trained three trials per day for 2 days to traverse the pole to the cage floor and were further trained two trials before testing on the test day. Mice were then tested for 2 trials, and the time to turn around (turning time) and the time to descend the pole (descending time) were recorded. Between trials, mice were placed in the home cage for 2 min to rest. Mice that stalled on the top of the pole for more than 120 sec were excluded.

Experimental design and statistical analysis

All experiments were performed in a genotype blind manner with the exception of molecular analysis (Southern, Northern, RT-PCR, and Western). All statistical analyses were performed using Prism 9 (GraphPad software) or Excel (Microsoft). All data are presented as the means ± SEM. The exact sample size of each experiment is indicated in the figure or the legend. The slight difference in sample size among various histological analysis (e.g. at 24M, 11 control and 8 cDKO brains analyzed for TH+ cells but 9 control and 8 cDKO analyzed for NeuN+, Caspase-3+, Iba1+ cells in the SNpc) is due to the paraffin blocks used for quantification (each paraffin block contains 7 brains of mixed genotypes; for details see Source Data files) or specific brain sections being damaged, thus, excluded in the analysis.

Statistical analyses were conducted using a two-tailed Student’s t-test for the comparison of a given variable in two genotypes or two-way ANOVA followed by Bonferroni’s post hoc comparisons for the comparison of more than two conditions. Statistical outliers were identified and excluded using the ROUT method with 1% the maximum desired false discovery rate developed by Prism, and the only statistical outliers of the current study were identified in the behavioral analysis and were marked in Figure 9 - Source Data. All statistical analysis was performed on data from ≥ 4 mouse brains per genotype per age group, and experiments were performed and repeated on different days, often by independent investigators. Statistical significance is indicated as *p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001, and values not significantly different are often not noted.

Data availability

All data generated and analyzed in this study are included in the manuscript and the supporting information. Source Data files are provided for key findings shown in Figures 1, 3, 4, 8, and 9.

Funding disclosure

This work was supported by a grant from the NIH (R37NS071251 to J.S.).

Supporting information

eLife Assessment

This current revision builds on observations in validated conditional double KO (cDKO) mice for LRRK1 and LRRK2 that will be useful for the field, given that LRRK2 is widely expressed in the brain and periphery, and many divergent phenotypes have been attributed previously to LRRK2 expression. The manuscript presents solid data demonstrating that it is the loss of LRRK1 and LRRK2 expression within the SNpc DA cells that is not well tolerated, as it was previously unclear from past work whether neurodegeneration in the LRRK double Knock Out (DKO) was cell autonomous or the result of loss of LRRK1/LRRK2 expression in other types of cells. Future studies may pursue the biochemical mechanisms underlying the reason for the apoptotic cells noted in this study, as here, the LRRK1/LRRK2 KO mice did not replicate the dramatic increase in the number of autophagic vacuoles previously noted in germline global LRRK1/LRRK2 KO mice.

Reviewer #1 (Public Review):

Summary:

This is an important work showing that loss of LRRK function causes late-onset dopaminergic neurodegeneration in a cell-autonomous manner. One of the LRRK members, LRRK2, is of significant translational importance as mutations in LRRK2 cause late-onset autosomal dominant Parkinson's disease (PD). While many in the field assume that LRRK2 mutant causes PD via increased LRRK2 activity (i.e., kinase activity), it is not a settled issue as not all disease-causing mutant LRRK2 exhibits increased activity. Further, while LRRK2 inhibitors are under clinical trials for PD, the consequence of chronic, long-term LRRK2 inhibition is unknown. Thus, studies evaluating the long-term impact of LRRK deficit have important translational implications. Moreover, because LRRK proteins, particularly LRRK2, are known to modulate immune response and intracellular membrane trafficking, the study's results and the reagents will be valuable for others interested in LRRK function.

Strengths:

This report describes a mouse model where LRRK1 and LRRK2 genes are conditionally deleted in dopaminergic neurons. Previously, this group showed that while loss of LRRK2 expression does not cause brain phenotype, loss of both LRRK1 and LRRK2 causes a later onset, progressive degeneration of catecholaminergic neurons, Dopaminergic (DAergic) neurons in substantia niga (SN) and Noradrenergic neurons in Locus Coeruleus (LC). However, because LRRK genes are widely expressed with some peripheral phenotypes, it was unknown if the neurodegeneration in LRRK double Knock Out (DKO) was cell autonomous. To rigorously test this question, the authors have generated a double conditional KO allele where both LRRK1 and LRRK2 genes were targeted to contain loxP sites. In my view, this was beyond what is normally required as most investigators might just combine one KO allele with another floxed allele. The authors provide a rigorous validation showing that the Driver (DAT-Cre) is expressed in the majority of DAergic neurons in SN and that LRRK levels are decreased selectively in the ventral midbrain. Using these mice, the authors show that the number of DA neurons is average at 15 but significantly decreased at 20 months of age. Moreover, the authors show that the number of apoptotic neurons is increased by ~2X in aged SN, demonstrating increased ongoing cell death, as well as an increase in activated microglia. The degeneration is limited to DA neurons as LC neurons are not lost as this population does not express DAT. Overall, the mouse genetics and experimental analysis were performed in a rigorous manner and the results were statistically sound and compelling.

Weakness: I only have a few minor comments. First, in PD and other degenerative conditions, axons and terminals loss occurs prior to cell bodies. It might be beneficial to show the status of DAergic markers in the striatum. Second, previous studies indicate that very little, if any, LRRK1 is expressed in SN DAergic neurons. This also seems to be the case with the Allen Brain Atlas profile. Thus, it is preferable that authors discuss the discrepancy as authors seem to imply significant LRRK1 expression in DA neurons.

Revision: I appreciate the authors revising the manuscript with additional data that have clarified my prior comments. They now show that TH levels in the striatum decrease with SNpc neurons. Further, while there is some disagreement regarding the expression LRRK1 in SNpc, the authors provide a convincing case that LRRK1 function is important in SNpc DA neurons.

Reviewer #2 (Public Review):

Summary: In this manuscript, Shen and collaborators described the generation of conditional double knockout (cDKO) mice lacking LRRK1 and LRRK2 selectively in DAT positive dopaminergic neurons. The Authors asked whether selective deletion of both LRRK isoforms could lead to a Parkinsonian phenotype, as previously reported by the same group in germline double LRRK1 and LRRK2 knockout mice (PMID: 29056298). Indeed, cDKO mice developed a late reduction of TH+ neurons in SNpc that partially correlated with the reduction of NeuN+ cells. This was associated with increased apoptotic cell and microglial cell numbers in SNpc. Unlike the constitutive DKO mice described earlier, however, cDKO mice did not replicate the dramatic increase in the number of autophagic vacuoles. The study supports the authors' hypothesis that loss of function rather than gain of function of LRRK2 leads to Parkinson's Disease.

Strengths: The study described for the first time a model where both the Parkinson's disease-associated gene LRRK2 and its homolog LRRK1 are deleted selectively in dopaminergic neurons, offering a new tool to understand the physiopathological role of LRRK2 and the compensating role of LRRK1 in modulating dopaminergic cell function.

Weaknesses: The model has no construct validity since loss of function mutations of LRRK2 are well tolerated in humans and do not lead to Parkinson's disease. The evidence of a Parkinsonian phenotype in these conditional knockout mice is limited and should be considered preliminary.

Reviewer #3 (Public Review):

Kang, Huang, and colleagues have provided new data to address concerns regarding confirmation of LRRK1 and LRRK2 deletion in their mouse model and the functional impact of the modest loss of TH+ neurons observed in the substantia nigra of their double KO mice. In the revised manuscript, the new data around the characterization of the germline-deleted LRRK1 and LRRK2 mice add confidence that LRRK1 and LRRK2 can be deleted using the genetic approach. They have also added new text to the discussion to try and address some of the comments and questions raised regarding how LRRK1/2 loss may impact cell survival and the implications of this work for PD-linked variants in LRRK2 and therapeutic approaches targeting LRRK2.

The new data provides additional support for the author's claims. I have provided below some suggestions for clarification/additions to the text that can be addressed without additional experiments.

1. The authors added additional text highlighting that more studies are warranted in mice where LRRK1/2 are deleted in other CNS cell types (microglia/astrocytes) to understand cell extrinsic drivers of the autophagy deficits observed in their previous work. It still remains unclear how loss of LRRK1/2 leads to increased apoptosis and gliosis in dopaminergic neurons in a cell-intrinsic manner, and, as suggested in the original review, it would be helpful to add some text to the discussion speculating on potential mechanisms by which this might occur.

2. Revisions have been made to the discussion to clarify their rationale around how variants in LRRK2 associated with PD may be loss-of-function to support the relevance of this mouse model to phenotypes observed in PD. However, as written, the argument that PD-linked variants are loss-of-function is based on the fact that the double KO mice have a mild loss of TH+ neurons while the transgenic mice overexpressing PD-linked LRRK2 variants often do not and that early characterization of kinase activity was done in vitro are relatively weak. Given that the majority of evidence generated by many labs in the field supports a gain-of-function mechanism, the discussion should be further tempered to better highlight the uncertainty around this (rather than strongly arguing for a loss-of-function effect). This could include the mention of increased Rab phosphorylation observed in cellular and animal models and opposing consequences on lysosomal function observed in cellular studies in KO and pathogenic variant expressing cells. Further, a reference to the Whiffen et al. 2020 paper mentioned by another reviewer should be included in the discussion for completeness.

Author response:

The following is the authors’ response to the previous reviews.

Public Reviews:

Reviewer #3 (Public Review):

Kang, Huang, and colleagues have provided new data to address concerns regarding confirmation of LRRK1 and LRRK2 deletion in their mouse model and the functional impact of the modest loss of TH+ neurons observed in the substantia nigra of their double KO mice. In the revised manuscript, the new data around the characterization of the germline-deleted LRRK1 and LRRK2 mice add confidence that LRRK1 and LRRK2 can be deleted using the genetic approach. They have also added new text to the discussion to try and address some of the comments and questions raised regarding how LRRK1/2 loss may impact cell survival and the implications of this work for PD-linked variants in LRRK2 and therapeutic approaches targeting LRRK2.

The new data provides additional support for the author's claims. I have provided below some suggestions for clarification/additions to the text that can be addressed without additional experiments.

(1) The authors added additional text highlighting that more studies are warranted in mice where LRRK1/2 are deleted in other CNS cell types (microglia/astrocytes) to understand cell extrinsic drivers of the autophagy deficits observed in their previous work. It still remains unclear how loss of LRRK1/2 leads to increased apoptosis and gliosis in dopaminergic neurons in a cell-intrinsic manner, and, as suggested in the original review, it would be helpful to add some text to the discussion speculating on potential mechanisms by which this might occur.

(2) Revisions have been made to the discussion to clarify their rationale around how variants in LRRK2 associated with PD may be loss-of-function to support the relevance of this mouse model to phenotypes observed in PD. However, as written, the argument that PD-linked variants are loss-offunction is based on the fact that the double KO mice have a mild loss of TH+ neurons while the transgenic mice overexpressing PD-linked LRRK2 variants often do not and that early characterization of kinase activity was done in vitro are relatively weak. Given that the majority of evidence generated by many labs in the field supports a gain-of-function mechanism, the discussion should be further tempered to better highlight the uncertainty around this (rather than strongly arguing for a loss-offunction effect). This could include the mention of increased Rab phosphorylation observed in cellular and animal models and opposing consequences on lysosomal function observed in cellular studies in KO and pathogenic variant expressing cells. Further, a reference to the Whiffen et al. 2020 paper mentioned by another reviewer should be included in the discussion for completeness.

We thank the reviewer for the comments. The discussion has been further revised and expanded to explain the cell extrinsic microglial response to pathophysiological changes in DA neurons of cDKO mice and propose future studies of single-cell RNA-sequencing to identify molecular changes within DA neurons of cDKO mice that may drive their apoptotic death during aging.

We also added paragraphs summarizing existing experimental evidence for the toxic gain-of-function mechanism (biochemical data of increased kinase activity but the lack of evidence for the elevated pRabs and the altered pLRRK2 driving dopaminergic neurodegeneration) and for the loss-of-function mechanism (genetic data of relevant physiological roles) as well as the relationships between LRRK1 and LRRK2 (functional homologues sharing functional domains and overlapping roles in dopaminergic neuron survival) and how dominantly inherited missense mutations can confer a loss of function mechanism (impairing its function in cis and inhibiting wild-type protein function in trans). We also provided a brief summary and discussion of the Whiffen et al. 2020 paper.

Version 3

Abstract

Abstract

Mutations in leucine-rich repeat kinase 2 (LRRK2) are the most common genetic cause of Parkinson’s disease (PD), which is the leading neurodegenerative movement disorder characterized by the progressive loss of dopaminergic (DA) neurons in the substantia nigra pars compacta (SNpc). However, whether LRRK2 mutations cause PD and degeneration of DA neurons via a toxic gain-of-function or a loss-of-function mechanism is unresolved and has pivotal implications for LRRK2-based PD therapies. In this study, we investigate whether LRRK2 and its functional homologue LRRK1 play an essential, intrinsic role in DA neuron survival through the development of DA neuron-specific LRRK conditional double knockout (cDKO) mice. We first generated and characterized floxed LRRK1 and LRRK2 mice and then confirmed that germline deletions of the floxed LRRK1 and LRRK2 alleles result in null mutations, as evidenced by the absence of LRRK1 and LRRK2 mRNA and protein in the respective homozygous deleted mutant mice. We further examined the specificity of Cre-mediated recombination driven by the dopamine transporter-Cre (DAT-Cre) knockin (KI) allele using a GFP reporter line and confirmed that DAT-Cre-mediated recombination is restricted to DA neurons in the SNpc. Crossing these validated floxed LRRK1 and LRRK2 mice with DAT-Cre KI mice, we then generated DA neuron-restricted LRRK cDKO mice and further showed that levels of LRRK1 and LRRK2 are reduced in dissected ventral midbrains of LRRK cDKO mice. While DA neuron-restricted LRRK cDKO mice of both sexes exhibit normal mortality and body weight, they develop age-dependent loss of DA neurons in the SNpc, as demonstrated by the progressive reduction of DA neurons in the SNpc of LRRK cDKO mice at the ages of 20 and 24 months but the unaffected number of DA neurons at the age of 15 months. Moreover, DA neurodegeneration is accompanied with increases of apoptosis and elevated microgliosis in the SNpc as well as decreases of DA terminals in the striatum, and is preceded by impaired motor coordination. Taken together, these findings provide the unequivocal evidence for the importance of LRRK in DA neurons and raise the possibility that LRRK2 mutations may impair its protection of DA neurons, leading to DA neurodegeneration in PD.

Introduction

Parkinson’s disease (PD) is the most common movement disorder and is characterized by the progressive loss of dopaminergic (DA) neurons in the substantia nigra pars compacta (SNpc). Dominantly inherited missense mutations in the leucine-rich repeat kinase 2 (LRRK2) gene are the most common cause of both familial and sporadic PD, highlighting the importance of LRRK2 in PD pathogenesis (1–16). LRRK2 is a large protein of 2,527 amino acid residues containing multiple functional domains including several leucine-rich repeats (LRRs), a GTPase-like domain of Ras-of-complex (Roc), a C-terminal of Roc (COR) domain, and a serine/threonine MAPKKK-like kinase domain. LRRK1, a homologue of LRRK2, belongs to the evolutionarily conserved Roco protein family and contains similar LRRs, Roc, COR, and kinase domains (17–19). LRRK proteins are broadly expressed with LRRK2 being most abundant in the kidney (20). While most PD mutations are found in LRRK2, rare variants in the Roc, COR, and kinase domains of LRRK1 have been reported and may be associated with PD (21). Furthermore, the Roc-COR domain of LRRK2 forms dimers and exhibits conventional Ras-like GTPase properties, and the R1441/C/G/H and I1371V mutations destabilize dimer formation and decrease GTPase activity (22, 23). Recent high-resolution cryoEM structural studies of full-length LRRK2 demonstrated its existence as dimers and pathogenic mutations such as R1441/C/G/H and Y1699I at the Roc-COR interface, whereas the G2910S mutant and wild-type LRRK2 share similar structure (24).

Previous genetic studies demonstrated that LRRK2 plays essential roles in the autophagy-lysosomal pathway (25–28). Consistent with high levels of LRRK2 expression in kidneys, LRRK2-/- mice develop age-dependent phenotypes in the kidney, including autophagy-lysosomal impairments and increases of α-synuclein, apoptosis, and inflammatory responses (25, 26, 29). It has also been reported that LRRK2 is important for maintaining lung homeostasis, and LRRK2 deficiency results in impaired autophagy in alveolar type II epithelial cells (28). It was proposed that the lack of brain phenotypes in LRRK2-/- mice might be due to the presence of LRRK1 that compensates functionally in the absence of LRRK2 (25, 26). Indeed, LRRK double knockout (DKO) mice develop age-dependent, progressive loss of DA neurons in the SNpc, beginning at 14 months of age (30). However, LRRK DKO mice also exhibit lower body weight and increased mortality, raising the possibility that DA neurodegeneration in LRRK DKO mice may be secondary to poor health.

In this study, we investigate whether LRRK2 and its functional homologue LRRK1 play an essential, intrinsic role in DA neuron survival through the development of DA neuron-specific LRRK conditional DKO (cDKO) mice. We first generated and confirmed floxed LRRK1 and LRRK2 mice, and then crossed them with the CMV-Cre deleter to create germline deletions of the floxed LRRK1 and LRRK2 alleles, followed by Northern and Western analyses to confirm the absence of LRRK1 and LRRK2 mRNA and protein in the respective homozygous deleted mutant mice. We also crossed a GFP reporter mouse with DAT-Cre KI mice and confirmed that Cre-mediated recombination occurs in most, if not all, DA neurons in the SNpc and is restricted to DA neurons. We then crossed these thoroughly validated floxed LRRK1 and LRRK2 mice with DAT-Cre KI mice to generate DA neuron-restricted LRRK cDKO mice and further confirmed the reduction of LRRK1 and LRRK2 in dissected ventral midbrains of LRRK cDKO mice. While DA neuron-restricted LRRK cDKO mice of both sexes exhibit normal mortality and body weight, they develop age-dependent loss of DA neurons in the SNpc, as demonstrated by the progressive reduction of TH+ DA neurons or NeuN+ neurons in the SNpc of cDKO mice at the ages of 20 and 24 months. Moreover, DA neurodegeneration is accompanied with increases of apoptosis and elevated microgliosis in the SNpc as well as decreases of DA terminals in the striatum, and is preceded by impaired motor coordination. Interestingly, quantitative electron microscopy (EM) analysis showed a similar number of electron-dense vacuoles in SNpc neurons of LRRK cDKO mice relative to controls, in contrast to age-dependent increases of vacuoles in SNpc neurons of germline LRRK DKO mice (30, 31). These findings provide the unequivocal evidence for the importance of LRRK in DA neurons and raise the possibility that LRRK2 mutations may impair this crucial physiological function, leading to the loss of DA neurons in Parkinson’s disease.

Results

Generation and molecular characterization of the floxed and deleted LRRK1 and LRRK2 alleles

LRRK2 and its homologue LRRK1 share several functional (LRRs, GTPase Roc, COR, and kinase) domains (Figure 1A). To investigate the intrinsic role of LRRK in DA neurons, we generated floxed LRRK1 (fLRRK1) and floxed LRRK2 (fLRRK2) mice, which permit deletions of LRRK1 and LRRK2 selectively in DA neurons by DAT-Cre, in which Cre recombinase is expressed under the control of the endogenous DAT promoter (32). We introduced two loxP sites in introns 26 and 29 of LRRK1 through homologous recombination and site-specific recombination by FLP recombinase to remove the positive selection PGK-neo cassette, which is flanked by two FRT sites (Figure 1B, C; Figure 1-figure supplement 1, 2). The embryonic stem (ES) cells carrying the targeted allele or the floxed allele were identified and validated by Southern analysis using the 5’ and 3’ external probes as well as the neo probe (Figure 1D; Figure 1-figure supplement 3). The validated ES cells carrying the floxed LRRK1 allele were injected into mouse blastocysts to generate floxed LRRK1 mice, which were further confirmed by Southern using the 5’ and 3’ external probes (Figure 1D). In the presence of Cre recombinase, the floxed LRRK1 genomic region containing part of intron 26 (1,288 bp), exons 27-29, which encode the kinase domain, and part of intron 29 (1,023 bp) is deleted (Figure 1C), and the removal of exons 27-29 (625 bp) results in a frameshift of downstream codons.

Generation and characterization of floxed and deleted LRRK1 and LRRK2 alleles

(A) Schematic illustrations of human LRRK1 and LRRK2 proteins showing similar functional domains. LRRK12015 protein is derived from exons 2-34 (Ensembl Genome Database: ENSG00000154237). LRRK22527 protein is derived from exons 1-51 (ENSG00000188906). LRR: Leucine rich repeats; Roc: Ras-of-complex; COR: C-terminal of Roc; KIN: Kinase domain.

(B) Schematic illustrations of the gene structures of mouse LRRK1 and LRRK2. The boxes in blue are exons that encode the LRRK1 and LRRK2 proteins, and the gray boxes represent the 5’ and 3’ UTRs. The exons are not drawn in scale. The start codon ATG is in exon 2 of LRRK1 and exon 1 of LRRK2. The exons 27-29 of LRRK1 and the promoter/exons 1-2 of LRRK2 are flanked with loxP sites (green arrowheads).

(C) Targeting strategy for the generation of the targeted, floxed, and deleted LRRK1 alleles. The red boxes represent the targeted exons 27-29, and the blue boxes represent the untargeted exons. The locations and sizes of the 5’ and 3’ external probes are shown. The targeting vector contains the 5’ and 3’ homologous regions (marked by dashed lines) and the middle region (from intron 26 to intron 29), which includes a loxP site (green arrowhead) in intron 26 (1,288 bp upstream of exon 27) and the PGK-neo selection cassette flanked by two FRT (FLP recognition target) sequences (gray circles) followed by another loxP site in intron 29 (1,023 bp downstream of exon 29). A negative selection cassette encoding diphtheria toxin fragment A (DTA) is also included in the targeting vector to reduce ES cells bearing randomly inserted targeting vectors. ES cells carrying the correctly targeted LRRK1 allele were transfected with pCAG-FLP to remove the PGK-neo cassette and generate the floxed LRRK1 allele. Floxed LRRK1 mice were bred with CMV-Cre transgenic mice to generate germline deleted LRRK1 mice. Detailed strategy for generating targeting vector and DNA sequence of floxed LRRK1 allele can be found in Figure 1-figure supplement 1 and 2, respectively.

(D) Southern analysis of the targeted and floxed LRRK1 alleles. Genomic DNA from ES cells or mouse tails was digested with HindIII and hybridized with the 5’ or 3’ external probe. For the 5’ probe, the resulting 17.0 kb and 4.8 kb bands represent the wild-type (WT) and the targeted (T) or floxed (F) alleles, respectively. For the 3’ probe, the resulting 17.0 kb and 12.2 kb bands represent the WT and the floxed alleles, respectively. Detailed Southern strategy can be found in Figure 1-figure supplement 3.

(E) Targeting strategy for the generation of the targeted, floxed, and deleted LRRK2 alleles. The red boxes represent LRRK2 exons 1 and 2, and the start codon ATG resides in exon 1. The locations and sizes of the 5’ and 3’ external probes are shown. The targeting vector contains the 5’ and 3’ homologous regions (marked by dashed lines) and the middle region (from the promoter to intron 2), which includes a loxP site (green arrowhead) upstream (1,768 bp) of the transcription initiation site and the PGK-neo selection cassette flanked by two FRT sequences (gray circles) and two loxP sites (green arrowheads) in intron 2 (878 bp downstream of exon 2). A negative selection cassette encoding DTA is also included in the targeting vector. Mice carrying the correctly targeted LRRK2 allele were crossed with Actin-FLP deleter mice to generate floxed LRRK2 mice, which were bred with CMV-Cre transgenic mice to generate germline deleted LRRK2 mice. Detailed strategy for generating targeting vector and DNA sequence of floxed LRRK2 allele can be found in Figure 1-figure supplement 4 and 5, respectively.

(F) Southern analysis of the targeted and floxed LRRK2 alleles. Genomic DNA from ES cells or mouse tails was digested with NheI and hybridized with the 5’ or 3’ external probe. For the 5’ probe, the resulting 11.5 kb and 3.6 kb bands represent the wild-type (WT) and the targeted (T) or floxed (F) alleles, respectively. For the 3’ probe, the resulting 11.5 kb and 5.2 kb bands represent the WT and the floxed alleles, respectively. Detailed Southern strategy can be found in Figure 1-figure supplement 6.

(G) Northern analysis of poly(A)+ RNA prepared from the lung of mice carrying homozygous germline deleted (11/11) LRRK1 alleles derived from the floxed LRRK1 alleles using the cDNA probe of exons 2-3 (left) and exons 27-29 (right). Using the upstream exons 2-3 probe, the LRRK1 transcripts in wild-type mice are the expected size of ∼7.4 kb, whereas the detected LRRK1 transcripts in LRRK1 11/11 mice are truncated, consistent with the deletion of exons 27-29 (625 bp), which results in a frameshift, and are expressed at lower levels, likely due to nonsense mediated degradation of the truncated LRRK1 mRNA. Using a probe specific for exons 27-29, there is no LRRK1 transcript in LRRK1 11/11 mice, as expected. Both blots were hybridized with a GAPDH probe as loading controls. Detailed Northern strategy and full-size blots are included in Figure 1-figure supplement 7 and 8, respectively. Extensive RT-PCR analysis of LRRK1 transcripts in 11/11 mice are shown in Figure 1-figure supplement 9.

(H) Northern analysis of total RNA prepared from the neocortex of mice carrying homozygous germline LRRK2 deleted (11/11) alleles using the cDNA probe of exons 1-5 shows the absence of LRRK2 transcripts. The blot was hybridized with a GAPDH probe as a loading control. The full-size blot is included in Figure 1-figure supplement 10. RT-PCR analysis of LRRK2 transcripts in 11/11 mice are shown in Figure 1-figure supplement 11

(I) Left: Western analysis of wild-type (+/+) and homozygous LRRK1 11/11 brains show the absence of LRRK1 protein. Right: Western analysis of the neocortex of wild-type (+/+) and homozygous LRRK2 11/11 mice shows the absence of LRRK2 protein. Vinculin was used as a loading control. Full-size blots can be found in Figure 1-Source Data 1.

The LRRK2 targeting vector contains the 5’ homologous region, a loxP site 1,768 bp upstream of the transcription initiation site, exons 1-2, the PGK-neo cassette flanked by two loxP sites and two FRT sequences, and the 3’ homologous region (Figure 1B, E; Figure 1-figure supplement 4, 5). The ES cell clones carrying the targeted allele were identified and validated by Southern analysis using the 5’ external probe (Figure 1F; Figure 1-figure supplement 6) and genomic PCR followed by sequencing. Mice carrying the LRRK2 targeted allele were further verified by Southern analysis using the 5’ and 3’ external probes (Figure 1-figure supplement 6) as well as the neo probe and were then crossed with the Actin-FLP mice (33) to remove the PGK-neo cassette flanked by two FRT sites to generate floxed LRRK2 mice (Figure 1E). The resulting floxed LRRK2 mice were confirmed by Southern analysis using the 5’ and 3’ external probes (Figure 1F). In the presence of Cre recombinase, the floxed LRRK2 region containing the promoter and exons 1-2 are deleted, likely resulting in a null allele, as we previously targeted a very similar region (∼2.5 kb upstream of the transcription initiation site and exons 1-2) to generate germline deletion of LRRK2 (25).

Dopaminergic neurons in the SNpc are a small neuronal population embedded in the ventral midbrain, making it difficult to confirm whether DA neuron-specific deletions of the floxed LRRK1 and LRRK2 regions result in null alleles. We previously generated three independent LRRK1 knockout (KO) mice, and only one KO line (Line 2) represents a LRRK1 null allele (30), whereas deletion of LRRK2 promoter region and exons 1-2 resulted in a LRRK2 null allele (25). We therefore generated germline deleted LRRK1 and LRRK2 mice from floxed LRRK1 and LRRK2 mice, respectively, by crossing them to germline deleter CMV-Cre transgenic mice (34).

We then performed Northern analysis of LRRK1 using both an upstream probe specific for exons 2-3 and a downstream probe specific for exons 27-29 (Figure 1G; Figure 1-figure supplement 7, 8). Because of the low expression level of LRRK1 mRNA and the relative abundance of LRRK1 mRNA in the lung (20), we enriched polyA+ RNA from the lung of the mice carrying homozygous germline deleted (11/11) LRRK1 alleles derived from the floxed LRRK1 alleles. Using the exons 2-3 probe, the LRRK1 transcripts in wild-type mice are of the expected size of ∼7.4 kb, whereas the detected LRRK1 transcripts in LRRK1 11/11 mice are smaller and expressed at lower levels, consistent with the deletion of exons 27-29 (625 bp), resulting in a frameshift of downstream codons and the likely degradation of the truncated LRRK1 mRNA (Figure 1G; Figure 1-figure supplement 8). Using a probe specific for exons 27-29, there is no LRRK1 transcript in LRRK1 11/11 mice (Figure 1G; Figure 1-figure supplement 8). Extensive RT-PCR analysis of total RNA isolated from the kidney, brain, and lung of LRRK1 11/11 mice using an exon 32-specific primer for RT and exon-specific primer sets for PCR (e.g. exons 4-8, 11-17, 20-25, and 25-31), followed by sequencing confirmation of the PCR products, indicated normal LRRK1 splicing in fLRRK1/fLRRK1 mice and the lack of LRRK1 exons 27-29 in LRRK1 11/11 mice (Figure 1-figure supplement 9).

Similarly, Northern analysis of LRRK2 using a probe specific for exons 1-5 and RT-PCR followed by sequencing confirmed the absence of LRRK2 mRNA in LRRK2 11/11 brains and normal LRRK2 transcripts in fLRRK2/fLRRK2 brains (Figure 1H; Figure 1-figure supplement 10, 11). Furthermore, Western analysis confirmed the absence of LRRK1 and LRRK2 proteins in the brain of LRRK1 11/11 and LRRK2 11/11 mice, respectively (Figure 1I). Taken together, our Northern, RT-PCR followed by sequencing, and Western analyses demonstrated that deletion of the floxed LRRK1 and LRRK2 alleles results in null mutations. Thus, floxed LRRK1 and LRRK2 alleles can be used to generate DA neuron-specific LRRK cDKO mice.

Generation and molecular characterization of DA neuron-specific LRRK cDKO mice

To generate DA neuron-specific LRRK cDKO mice, we used DAT-Cre KI mice, which express Cre recombinase under the control of the endogenous DAT promoter (32). To confirm if Cre-mediated recombination occurs broadly and specifically in DA neurons of the SNpc, we crossed DAT-Cre mice with the Rosa26-CAG-LSL-ZsGreen1 reporter mouse (35). Upon Cre expression, Cre recombinase removes the floxed "stop" cassette, resulting in the expression of EGFP. We found that Cre-mediated recombination (GFP+) occurs in TH+ cells in the SNpc (Figure 2A). Quantification of GFP+ and/or TH+ cells in the SNpc showed that 99% of TH+ DA neurons are also GFP+, demonstrating that Cre-mediated recombination takes place in essentially all DA neurons in the SNpc (Figure 2B).

Generation and characterization of DA neuron-specific LRRK cDKO mice

(A) Immunostaining of GFP and/or TH in the SNpc of DAT-Cre/+; Rosa26-CAG-LSL-ZsGreen1/+ mice at 2 months of age. Cre recombinase is expressed under the control of the DAT endogenous promoter and removes the floxed "stop" cassette, resulting in the expression of EGFP under the control of the ubiquitous CAG promoter.

(B) Quantification of GFP+/TH+ and TH+ cells shows that 99% of TH+ DA neurons (722 ± 46 TH+ cells) in the SNpc are also GFP+ (713 ± 46 cells), indicating that DAT-Cre mediated recombination occurs in essentially all TH+ DA neurons. N = 3 mice, 3 comparable sections per hemisphere, 320 μm apart.

(C) Similar body weight between LRRK cDKO mice and littermate controls at all ages examined (F1,68 = 0.001310, p = 0.9712; 2M, 20M : p > 0.9999; 15M: p = 0.7857, 25M: p = 0.8084, two-way ANOVA with Bonferroni’s post hoc multiple comparisons).

(D) Similar brain weight between LRRK cDKO and control mice (F1,68 = 3.603, p = 0.0619; 2M: p = 0.3893; 15M: p > 0.9999; 20M: p > 0.9999; 25M: p = 0.3223, two-way ANOVA with Bonferroni’s post hoc multiple comparisons).

(E) Western analysis of LRRK1 and LRRK2 proteins in the dissected cerebral cortex (CTX) and ventral midbrain (vMB) of LRRK cDKO and littermate controls at 2 months of age.

(F) Quantification shows significant decreases of LRRK1 and LRRK2 in the dissected ventral midbrain of LRRK cDKO mice (LRRK1, p = 0.0432; LRRK2, p = 0.0162, Student’s t-test), compared to controls, but not in the dissected cortex of cDKO mice (LRRK1: p = 0.7648; LRRK2: p = 0.2325).

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. *p < 0.05. Scale bar: 100 μm.

Having confirmed fLRRK1 and fLRRK2 mice as well as DAT-Cre mice, we then bred them together to generate LRRK cDKO mice (fLRRK1/fLRRK1; fLRRK2/fLRRK2; DAT-Cre/+), which were further bred with fLRRK1/fLRRK1; fLRRK2/fLRRK2 mice to generate cDKO and littermate controls (fLRRK1/fLRRK1; fLRRK2/fLRRK2). It was previously reported that germline LRRK DKO mice failed to gain weight as they aged (30). However, DA neuron-specific LRRK cDKO and littermate control mice have similar body and brain weights at the ages of 2-24 months (Figure 2C, D). Western analysis showed a significant reduction of LRRK1 and LRRK2 proteins in the dissected ventral midbrain but not in the cerebral cortex of DA neuron-specific LRRK cDKO mice at 2-3 months of age, relative to littermate controls (Figure 2E, F), further validating these DA neuron-specific LRRK cDKO mice.

Age-dependent loss of TH+ DA neurons in the SNpc of LRRK cDKO mice

To determine whether the inactivation of LRRK selectively in DA neurons of the SNpc affects their survival, we performed TH immunostaining and quantified TH+ DA neurons in the SNpc of LRRK cDKO mice and littermate controls using stereological methods. The morphology of TH+ DA neurons in LRRK cDKO mice at the ages of 15, 20, and 24 months appears normal (Figure 3A). Quantification of TH+ neurons in the SNpc using serial sections emcompassing the entire SNpc revealed that the number of DA neurons in the SNpc at the age of 15 months is similar between cDKO mice (10,000 ± 141) and littermate controls (10,077 ± 310; F1,46 = 16.59, p = 0.0002, two-way ANOVA with Bonferroni’s post hoc multiple comparisons, p > 0.9999; Figure 3B). However, at the age of 20 months, the number of TH+ neurons in the SNpc of cDKO mice (8,948 ± 273) is significantly reduced, compared to controls (10,244 ± 220; p = 0.0041), and is further decreased at the age of 24 months (Control: 9,675 ± 232, cDKO: 8,188 ± 452; p = 0.0010; Figure 3B). Similar genotypic differences were observed in an independent quantification by another investigator, also conducted in a genotype blind manner, using the fractionator and optical dissector to randomly sample 25% area of the SNpc (Figure 3-figure supplement 1).

Age-dependent loss of DA neurons in the SNpc of LRRK cDKO mice

(A) TH immunostaining shows TH+ DA neurons in the SNpc of LRRK cDKO and littermate controls at the age of 15, 20, and 24 months. Higher power views of the boxed areas show grossly normal DA neuron morphology in LRRK cDKO mice.

(B) Quantification of TH+ DA neurons in the SNpc reveals similar numbers of DA neurons in LRRK cDKO mice (10,000 ± 141) and littermate controls (10,077 ± 310, p > 0.9999) at 15 months of age. At 20 months of age, the number of DA neurons in the SNpc of LRRK cDKO mice (8,948 ± 273) is significantly reduced, compared to control mice (10,244 ± 220, F1,46 = 16.59, p = 0.0002; p = 0.0041, two-way ANOVA with Bonferroni’s post hoc multiple comparisons). By 24 months of age, the reduction of DA neurons in the SNpc of LRRK cDKO mice (8,188 ± 452) relative to controls (9,675 ± 232, p = 0.0010) is greater, compared to 20 months of age. Raw quantification data are included in Figure 3-Source Data 1.

(C) Immunohistological analysis of TH and NeuN shows TH+ DA neurons (green) and NeuN+ neurons (red) in the SNpc of LRRK cDKO mice and controls at 24 months of age.

(D) Quantification of NeuN+ cells in the SNpc shows that the number of NeuN+ neurons in LRRK cDKO mice (17,923 ± 813) is significantly lower than that in control mice (21,907 ± 469, p = 0.0006, Student’s t-test), indicating loss of neurons in the SNpc of LRRK cDKO mice. All TH+ cells are NeuN+. The number of TH+/NeuN+ cells in the SNpc of LRRK cDKO mice (10,500 ± 644) is also lower, compared to control mice (14,102 ± 310, p = 0.0001). There is no significant difference in the number of NeuN+/TH-neurons between littermate controls (7,804 ± 249) and cDKO mice (7,423 ± 344, p = 0.3747).

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. **p < 0.01, ***p < 0.001. Scale bar: 100 μm.

We further performed TH and NeuN double immunostaining of LRRK cDKO and control mice at 24 months of age (Figure 3C). Quantification of NeuN+ and TH+/NeuN+ cells in the SNpc using serial sections encompassing the entire SNpc showed that the number of NeuN+ neurons is also significantly reduced in the SNpc of cDKO mice (17,923 ± 813), compared to controls (21,907 ± 469, p = 0.0006, Student’s t-test; Figure 3D). The number of TH+/NeuN+ cells is also lower in the SNpc of LRRK cDKO mice (10,500 ± 644), compared to control mice (14,102 ± 310, p = 0.0001; Figure 3D). These data indicate that the reduction in TH+ cells is not due to decreases of TH expression in DA neurons, but rather a result of the loss of DA neuron cell bodies in the SNpc of LRRK cDKO mice.

We further evaluated apoptosis in the SNpc of LRRK cDKO and littermate controls at the age of 24 months using an antibody specific for active Caspase-3 to label apoptotic cells. We observed increases in apoptotic DA neurons, labeled by active Caspase-3+ and TH+ immunoreactivity, in the SNpc of LRRK cDKO mice (Figure 4A). Quantification of active Caspase-3+/TH+ apoptotic DA neurons shows a significant increase in the SNpc of LRRK cDKO mice (323 ± 38), compared to controls (157 ± 8, p = 0.0004, Student’s t-test; Figure 4B). These results further support that LRRK plays an intrinsic role in the survival of DA neurons in the SNpc during aging.

Increases in apoptotic DA neurons in the SNpc of LRRK cDKO mice

(A) Representative images of TH and active Caspase-3 immunostaining show TH+ DA neurons (green) and active Caspase-3+ apoptotic cells (red) in the SNpc of LRRK cDKO and control mice at the age of 24 months.

(B) Quantification of active Caspase-3+/TH+ cells shows significant increases of apoptotic DA neurons in the SNpc of LRRK cDKO mice (323 ± 38) at 24 months of age, relative to controls (157 ± 8, p = 0.0004, Student’s t-test). Raw quantification data are included in Figure 4-Source Data 1.

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. ***p < 0.001. Scale bar: 100 μm.

Age-dependent loss of TH+ DA terminals in the striatum of LRRK cDKO mice

To determine whether loss of DA neurons in the SNpc is accompanied with loss of DA terminals in the striatum, we performed TH immunostaining and quantified TH immunoreactivity in the striatum of LRRK cDKO and littermate control mice at the ages of 15 and 24 months (Figure 5A). Quantitative analysis showed normal levels of TH immunoreactivity in the striatum of cDKO mice at 15 months of age but reduced levels of TH immunoreactivity in the striatum of cDKO mice at 24 months of age (-19%, p = 0.0215, Student’s t-test), suggesting an age-dependent loss of TH+ dopaminergic terminals in the striatum (Figure 5B).

Age-dependent loss of TH+ DA terminals in the striatum of LRRK cDKO mice

(A) Representative TH immunostaining images in the striatum of LRRK cDKO mice and littermate controls at the ages of 15 and 24 months.

(B) Quantification of TH immunoreactivity in the striatum of LRRK cDKO and control mice shows similar TH immunoreactivity at the age of 15 months (p = 0.8766, Student’s t-test), but there is a significant decrease in TH immunoreactivity in the striatum of LRRK cDKO mice at 24 months of age (-19%, p = 0.0215), compared to controls.

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. *p < 0.05. Scale bar: 1 mm.

Unaffected TH+ noradrenergic neurons in the LC of LRRK cDKO mice

Previously, we reported the reduction of noradrenergic neurons in the locus coeruleus (LC) of LRRK DKO mice (30). We therefore quantified TH+ noradrenergic neurons in the LC of LRRK cDKO mice at the age of 24 months. The number of TH+ cells in the LC is similar between cDKO and littermate controls (Control: 3,418 ± 86, cDKO: 3,350 ± 99, p = 0.6110, Student’s t-test; Figure 6A, B). Further examination of the GFP reporter line crossed with DAT-Cre showed the lack of Cre-mediated recombination in the LC (Figure 6C), further support the cell intrinsic nature of DA neuron loss in the SNpc of aged LRRK cDKO mice.

Normal number of TH+ noradrenergic neurons in the LC of LRRK cDKO mice

(A) Representative images of TH+ noradrenergic neurons in the locus coeruleus (LC) of LRRK cDKO mice and littermate controls at 24 months of age.

(B) Quantification of TH+ cells shows similar number of TH+ noradrenergic neurons in the LC of LRRK cDKO mice (3,350 ± 99) and controls (3,418 ± 86, p = 0.6110, Student’s t-test).

(C) Top: Immunostaining of TH and GFP in the LC of DAT-Cre/+; Rosa26-CAG-LSL-ZsGreen1/+ mice at 2 months of age. There is no GFP+ (green) cell in the LC, indicating that DAT-Cre is not expressed in the LC. Bottom: Higher power views of the boxed areas.

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. Scale bar: 100 μm.

Quantitative EM analysis of the SNpc in LRRK cDKO mice

We previously reported age-dependent increases of electron-dense autophagic and autolysosomal vacuoles as well as the presence of large lipofuscin granules in the surviving SNpc neurons of LRRK DKO mice beginning at 10 months of age (30, 31). To determine whether selective inactivation of LRRK in DA neurons similarly results in the accumulation of electron-dense vacuoles in the SNpc, we performed EM analysis in the SNpc of LRRK cDKO mice and littermate controls at the age of 25 months (Figure 7). We observed various electron-dense double membrane autophagosomes and single membrane autolysosomes as well as lipofuscin granules composed of lipid-containing residues of lysosomal digestion in the SNpc of LRRK cDKO and littermate control mice (Figure 7C-F). Interestingly, the number of electron-dense vacuoles in the SNpc is similar between LRRK cDKO mice and littermate controls at the age of 25 months (Control: 6.72 ± 0.43, cDKO: 6.99 ± 0.52, p = 0.6839, Student’s t-test; Figure 7G). We also found no significant difference in the area of electron-dense vacuoles in the SNpc between LRRK cDKO and littermate controls (Control: 4.43 ± 0.44 μm2; cDKO: 4.60 ± 0.49 μm2, p = 0.8048; Figure 7G). The difference in accumulation of electron-dense vacuoles in the SNpc between germline DKO mice and DA neuron-restricted cDKO suggests that LRRK in non-DA neurons, possibly microglia, may play a more prominent role in the regulation of the autophagy-lysosomal pathway.

Unchanged number of electron-dense vacuoles in the SNpc of LRRK cDKO mice

(A, B) Representative EM images showing electron-dense vacuoles (arrowheads) in SNpc neurons of cDKO mice and littermate controls at the age of 25 months.

(C-F) Higher power views showing various electron-dense vacuoles, autolysosomes (single arrows), autophagosomes (double arrows), and lipid-containing vacuoles (asterisks) in SNpc neurons of littermate control (C, D) and cDKO (E, F) mice.

(G) Left: The average number of electron-dense vacuoles (>0.5 μm in diameter) in the SNpc neuronal profiles per mouse is not significantly different between LRRK cDKO mice and littermate controls at the age of 25 months (Control: 6.72 ± 0.43; cDKO: 6.99 ± 0.52, p = 0.6839, Student’s t-test). Right: The total area of electron-dense vacuoles (>0.5 μm in diameter) in the SNpc neuronal profiles per mouse is similar between LRRK cDKO and littermate controls (Control: 4.43 ± 0.44 μm2; cDKO: 4.60 ± 0.49 μm2, p = 0.8048).

The value in parentheses indicates the number of mice (left) and neuron profiles (right) used in the quantification. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. Scale bar: 1 μm.

Enhanced microgliosis in the SNpc of LRRK cDKO mice

To determine whether selective inactivation of LRRK in DA neurons results in elevated microgliosis in the SNpc of LRRK cDKO mice. We performed immunohistochemical analysis of Iba1, which labels microglia, and TH, which marks DA neurons and processes, thus showing the boundary of the SNpc (Figure 8A-C). We found that the number of Iba1+ immunoreactive microglia is significantly increased in the SNpc of LRRK cDKO mice at 15 months of age (2541 ± 193), compared to controls (1737 ± 83; F1,45 = 102.6, p < 0.0001, two-way ANOVA with Bonferroni’s post hoc multiple comparisons, p = 0.0017; Figure 8A, D). The number of Iba1+ immunoreactive microglia in the SNpc of LRRK cDKO is further increased compared to controls at the age of 20 (Control: 2426 ± 68, cDKO: 3639 ± 127, p < 0.0001; Figure 8B, D) and 25 months (Control: 2640 ± 187, cDKO: 4089 ± 100, p < 0.0001; Figure 8C, D). These results show that despite the selective inactivation of LRRK in DA neurons of LRRK cDKO mice, microgliosis accompanies DA neuronal loss in the SNpc.

Elevated microgliosis in the SNpc of LRRK cDKO mice

(A-C) Representative images of Iba1+ microglia (red, marked by yellow arrowheads) and TH+ dopaminergic neurons (green) in the SNpc of LRRK DKO mice and controls at 15 (A), 20 (B), and 24 (C) months of age.

(D) Quantification of Iba1+ microglia shows significant increases in the number of Iba1+ microglia in the SNpc of LRRK cDKO mice, compared to control mice at the ages of 15 months (Control: 1737 ± 83, cDKO: 2541 ± 193; F1,45 = 102.6, p < 0.0001, p = 0.0017, two-way ANOVA with Bonferroni’s post hoc multiple comparisons), 20 months (Control: 2426 ± 68, cDKO: 3639 ± 127, p < 0.0001), and 24 months (Control: 2640 ± 187, cDKO: 4089 ± 100, p < 0.0001). Raw quantification data are included in Figure 8-Source Data 1.

The number in the column indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. **p < 0.01, ****p < 0.0001. Scale bar: 100 μm.

Impaired motor coordination of LRRK cDKO mice

To determine whether LRRK cDKO mice show motor deficits, we performed behavioral analysis of LRRK cDKO and littermate controls at the ages of 10 and 22 months using two versions of the elevated beam walk test with varying width of the beam (Figure 9). LRRK cDKO mice at 10 months of age displayed significantly more hindlimb slips/errors (4.4 ± 0.5) and longer traversal time (7.3 ± 0.3) in the 10 mm beam walk test, relative to littermate controls, which exhibited fewer slips (2.0 ± 0.3, p = 0.0005, Student’s t-test) and shorter traversal time (5.8 ± 0.4, p = 0.0075; Figure 9A). In the 20 mm beam walk, which is less challenging than the narrower beam walk, both LRRK cDKO (1.5 ± 0.2) and littermate controls (1.0 ± 0.2, p = 0.0733) at 10 months of age performed well with few hindlimb slips and with no difference between LRRK cDKO and control mice. In addition, the traversal time is similar between LRRK cDKO (5.2 ± 0.3) and control littermates (5.1 ± 0.4, p = 0.9796; Figure 9A). These results show that LRRK cDKO mice at 10 months of age already exhibit deficits in motor coordination. However, in the pole test LRRK cDKO and control mice showed similar turning time (Control: 1.2 ± 0.1; cDKO: 1.5 ± 0.1; p = 0.1219) and descending time (Control: 4.7 ± 0.1; cDKO: 4.7 ± 0.2; p = 0.8620; Figure 9B). The LRRK cDKO and control mice at 10 months of age exhibit similar body weight (Figure 9C).

Impairment of motor coordination in LRRK cDKO mice

(A) In the 10 mm beam walk test, compared to control mice, LRRK cDKO mice at 10 months of age exhibit markedly more hindlimb slips (Control: 2.0 ± 0.3; cDKO: 4.4 ± 0.5; p = 0.0005, Student’s t-test) and longer traversal time (Control: 5.8 ± 0.4; cDKO: 7.3 ± 0.3; p = 0.0075). In the less challenging 20 mm beam walk test, there is no significant difference in the number of hindlimb slips (Control: 1.0 ± 0.2; cDKO: 1.5 ± 0.2; p = 0.0733) and traversal time (Control: 5.1 ± 0.4; cDKO: 5.2 ± 0.3; p = 0.9796) between LRRK cDKO and control mice.

(B) In the pole test, LRRK cDKO and control mice at 10 months of age display similar turning time (Control: 1.2 ± 0.1; cDKO: 1.5 ± 0.1; p = 0.1219) and descending time (Control: 4.7 ± 0.1; cDKO: 4.7 ± 0.2; p = 0.8620).

(C) LRRK cDKO (31.1 ± 1.3) and control (32.0 ± 1.5; p = 0.6410) mice at 10 months of age show similar body weight.

(D) LRRK cDKO mice and control mice at 22 months of age in the 10 mm beam walk test show similar hindlimb slips (Control: 12.0 ± 3.0; cDKO: 13.8 ± 3.3; p = 0.7022) and traversal time (Control: 15.3 ± 2.0; cDKO: 16.2 ± 2.8; p = 0.8139). In the 20 mm team walk test, there is also no difference in hindlimb slips (Control: 4.8 ± 1.6; cDKO: 4.1 ± 1.0; p = 0.7142) and traversal time (Control: 8.7 ± 1.6; cDKO: 6.4 ± 0.7; p = 0.2223) between LRRK cDKO and control mice.

(E) In the pole test, LRRK cDKO mice at 22 months of age exhibit similar turning time (Control: 1.7 ± 0.2; cDKO: 2.8 ± 0.7; p = 0.1184) and descending time (Control: 6.6 ± 0.7; cDKO: 5.4 ± 0.3; p = 0.1413) compared to control mice.

(F) LRRK cDKO mice (34.8 ± 1.5) have similar body weight as control mice (39.6 ± 2.4; p = 0.1194).

The number in parentheses indicates the number of mice used in the study. Red-filled and open circles represent data obtained from individual male and female mice, respectively. All data are expressed as mean ± SEM. **p < 0.01, ***p < 0.001. Raw behavior data are included in Figure 9-Source Data 1.

We then performed the beam walk test in naïve LRRK cDKO and control mice at 22 months of age (Figure 9D). The aged LRRK cDKO and control mice performed similarly in hindlimb errors (p = 0.7022) and traversal time (p = 0.8139) in 10 mm beam walk test. The hindlimb slips (Control: 12.0 ± 3.0; cDKO: 13.8 ± 3.3) and traversal time (Control: 15.3 ± 2.0; cDKO: 16.2 ± 2.8) of both genotypic groups were much higher, compared to the respective younger mice (p <0.0001; Figure 9D). Moreover, higher percentage of the aged mice (2-3 out of 8-9 per genotypic group) failed the test, compared to the younger mice (0-2 mice out of 18-20 per genotypic group, Figure 9A, D). These results show that aged mice of both genotypes perform poorly in the challenging narrow beam walk test, suggesting that it is not optimal to reveal subtle differences in motor coordination between the genotypic groups. The performance of LRRK cDKO and control mice at 22 months of age in the 20 mm beam walk test was also not significantly different with similar hindlimb errors (p = 0.7142) and traversal time (p = 0.2223; Figure 9D). In the pole test, LRRK cDKO mice and littermate controls also displayed similar turning time (p = 0.1184) and descending time to their home cage (p = 0.1413; Figure 9E) as well as body weight (Figure 9F).

Discussion

LRRK2 mutations are linked to familial PD and are also associated with sporadic PD with the G2019S mutation being the most common but exhibiting lower penetrance and higher age of onset (e.g. 74% at age 79) (36). Despite the importance of LRRK2 in PD pathogenesis, whether LRRK2 mutations cause DA neurodegeneration via a loss- or gain-of-function mechanism remains unresolved, even though the distinction between these two pathogenic mechanisms is critical for directing LRRK2-based PD therapeutic development. Neither LRRK2-deficient mice nor LRRK2 R1441C and G2019S KI mice develop dopaminergic neurodegeneration (25, 37, 38). It was proposed that the lack of brain phenotypes in LRRK2-/- mice might be due to the presence of LRRK1 (25, 26), which is broadly expressed in the brain including the midbrain (https://www.proteinatlas.org/ENSG00000154237-LRRK1/brain). Indeed, germline deletions of LRRK2 and LRRK1 results in age-dependent loss of DA neurons and increases of apoptosis and microgliosis in the SNpc of LRRK DKO mice (30, 31). However, the pleiotropic roles of LRRK1 and LRRK2 in the kidney (25, 26, 29), lung (28), and bone (39, 40), and the observed earlier mortality and lower body weight in LRRK DKO mice raised the possibility that DA neurodegeneration in LRRK DKO mice may be due to poor health.

To address this question, we generated floxed LRRK1 and LRRK2 mice (Figure 1, Figure1-figure supplement 1-10) and DA neuron-specific LRRK cDKO mice using the DAT-Cre KI allele (32) to delete floxed LRRK1 and LRRK2 regions selectively in DA neurons (Figure 2). Using an EGFP reporter, we found that DAT-Cre drives Cre-mediated recombination in almost all DA neurons in the SNpc. Unlike germline DKO mice (30), DA neuron-specific LRRK cDKO mice exhibit normal body weight and mortality during mouse lifespan. Importantly, LRRK cDKO mice develop age-dependent, progressive DA neurodegeneration, as evidenced by the normal number of DA neurons at the age of 15 months and the progressive reduction of DA neurons in the SNpc at the ages of 20 and 24 months (Figure 3), accompanied with increases of apoptotic DA neurons in the SNpc (Figure 4) and decreases of DA terminals in the striatum (Figure 5), whereas the number of noradrenergic neurons in the LC is unchanged, consistent with the lack of Cre-mediated recombination in the LC by DAT-Cre (Figure 6). Moreover, LRRK cDKO mice exhibit impaired performance in a narrow beam walk, a behavioral paradigm sensitive to altered motor coordination (Figure 9). Furthermore, microgliosis is elevated in the SNpc of LRRK cDKO mice, even though LRRK expression in microglia is unaffected in these mice, suggesting a cell extrinsic microglial response to pathophygiological changes in DA neurons (Figure 8). The molecular mechanism by which LRRK supports cell autonomous DA neuron survival is unknown. Single-cell RNA sequencing of dissected ventral midbrains may permit identification of genes and pathways that are selectively altered in DA neurons of LRRK cDKO mice before and after the onset of DA neurodegeneration.

Interestingly, quantitative EM analysis showed similar numbers of electron-dense vacuoles in the SNpc of LRRK cDKO and control mice at 25 months of age (Figure 7), in contrast to age-dependent increases of vacuoles in the SNpc of LRRK DKO mice (30, 31), suggesting non-DA neurons (e.g. microglia), in which LRRK expression is unaffected in DA neuron-specific LRRK cDKO mice but is absent in germline DKO mice, may contribute to this phenotypic difference. Development of microglia- and/or astrocyte-specific LRRK cDKO mice would permit further dissection of cell autonomous and non-cell autonomous roles of LRRK in DA neuron survival and determine whether glial LRRK plays a more important role in the regulation of the autophagy-lysosomal pathway and protein turnover. It would also be interesting to test whether reducing LRRK expression or function affects accumulation and aggregation of human mutant α-synuclein and tau in the aging brain, as both Lewy bodies and tauopathies are associated with PD patients carrying LRRK2 mutations (3).

The toxic gain-of-function pathogenic mechanism was initially proposed based on in vitro kinase assays using recombinant mutant LRRK2, which showed that R1441C and G2019S resulted in increases of autophosphorylation and phosphorylation of a generic substrate (41). Subsequent biochemical studies further reported that LRRK2 mutations (e.g. R1441C/G, G2019S) enhanced kinase activity, leading to elevated levels of pSer1292-LRRK2 (42) as well as pT73-Rab10 and pS106-Rab12 (43). While these biochemical changes are excellent biomarkers of LRRK2 mutations, there is no experimental evidence showing that increased phosphorylation of LRRK2 substrates drives DA neurodegeneration.

The loss-of-function pathogenic mechanism was prompted by mouse knockout findings showing that germline inactivation of LRRK results in age-dependent loss of DA neurons in the SNpc and DA terminals in the striatum (30, 31). The current study demonstrated a cell intrinsic role of LRRK in support of DA neuron survival in the SNpc of the aging brain, providing an unequovacal genetic evidence of an essential, cell autonomous requirement of LRRK in DA neurons. While most transgenic mice overexpressing mutant LRRK2 did not produce neurodegeneration or DA neuron loss (44–52), two studies reported that overexpression of LRRK2 G2019S but not R1441C resulted in DA neurodegeneration (53, 54). These transgenic mouse results are intriguing, as R1441C is a more potent causative mutaion (highly penetrant, in a critical amino acid residue harboring 3 distinct PD mutations), whereas G2019S is known to be a weaker mutation with lower penetrance and older age of disease onset.

It remains possible that the relevant physiological role of LRRK in protection of DA neuron survival during aging may be irrelevant to how LRRK2 mutations cause DA neurodegeneration and PD. Two reports of genomic database analysis suggested that loss-of-function LRRK2 and LRRK1 variants are not associated with PD (55, 56), which were sometimes taken as conclusive evidence for LRRK2 mutations not being loss of function One study was a predictive analysis using several available sequencing databases (141,456 individuals sequenced in the Genome Aggregation Database, 49,960 exome-sequenced individuals from the UK Biobank, and more than 4 million participants in the 23andMe genotyped dataset), and the authors cautiously stated that heterozygous predicted loss-of-function variants are not strongly associated with PD (55). Another study analyzed next-generation sequencing data from 11,095 PD patients and 12,615 controls, and found that LRRK1 loss-of-function variants were identified in 0.205% PD cases and 0.139% of controls, whereas LRRK2 loss-of-function variants were found in 0.117% of PD cases and 0.087% of controls (56). In contrast to linkage studies of large pedigrees with mutations segregating completely with the disease but being absent among control populations, these genomic database analyses are considerably weaker with the results suggestive but highly inconclusive. For example, while the occurrence of these loss-of-function variants among PD patients is not statistically different from those among control populations, a larger sample size may change this outcome. Furthermore, such sequencing data tend to be from highly heterogeneous populations, contributing to further complexity. Moreover, it is quite possible that haploinsufficency or heterozygous complete loss of function mutations are not sufficient to cause the disease; rather a dominant negative mechanism coupled with loss of function missense mutations may be at play, reducing overall protein function further, especially if dimers are the functional entities. Therefore, these studies are interesting but inconclusive as to whether loss-of-function mutations in LRRK2 or LRRK1 cause or do not cause PD.

Genetically, it is well-known that missense mutations may gain a toxic function or may cause a partial loss of function in cis and a dominant negative inhibition of wild-type protein in trans, further reducing function. Missense mutations in the PSEN genes linked to familial Alzheimer’s disease are good examples of such a loss-of-function coupled with a dominant negative mechanism (57–63). While PD-linked LRRK2 mutations are mostly dominantly inherited missense mutations, though homozygous R1441H carriers have been reported (12), variants in LRRK1 have also been reported with inconclusive pathogenicity (21, 64). The notion that mutant LRRK2 may act in a dominant negative manner to inhibit the activity of wild-type LRRK2 is supported by structural and functional studies of LRRK2 as homodimers or heterodimers with LRRK1(22, 24, 65–69). Furthermore, LRRK2 G2385R variant, a risk factor of PD (70–73), has been reported as a partial loss-of-function mutation (74, 75). Future studies are needed to reconcile the biochemical findings of elevated kinase activity by LRRK2 mutations and the genetic findings of an essential physiological requirement of LRRK in DA neuron survival, and to determine whether LRRK2 mutations impair this relevant physiological function in protecting DA neurons while increasing kinase activity.

Materials and methods

Key Resource Table

Mice

All animal use was approved by the IACUC committees of Harvard Medical School and Brigham and Women’s Hospital, and conformed to the USDA Animal Welfare Act, PHS Policy on Humane Care and Use of Laboratory Animals, the “ILAR Guide for the Care and Use of Laboratory Animals” and other applicable laws and regulations. Mice were housed in constant humidity- and temperature-controlled rooms and maintained on a 12 hr light/dark cycle and were given standard rodent chow and water. Mice of both sexes at multiple ages, from 2 months to 25 months, were used. Floxed LRRK1 (fLRRK1) and the resulting germline deleted LRRK1Λ1 alleles, and floxed LRRK2 (fLRRK2) and the resulting germline deleted LRRK2Λ1 alleles were generated and thoroughly validated at the genomic DNA, mRNA, and protein levels. DAT-IRES-Cre (The Jackson Laboratory, IMSR_JAX:006660), ACT-FLPe (IMSR_JAX:005703), CMV-Cre transgenic (IMSR_JAX:006054), and Rosa26-CAG-LSL-ZsGreen1 mice (IMSR_JAX:007906) used in the current study were previously characterized and reported (32–35). LRRK1/LRRK2 cDKO mice (fLRRK1/fLRRK1; fLRRK2/fLRRK2; DAT-IRES-Cre/+) and littermate controls (fLRRK1/fLRRK1; fLRRK2/fLRRK2) were obtained from multiple breeding cages of fLRRK1/fLRRK1; fLRRK2/fLRRK2 and cDKO mice. All mice were maintained on the C57BL6 and 129 hybrid genetic background (F1: IMSR_JAX:101043). All phenotypic analyses were performed in a genotype blind manner, as previously described (76).

Generation of targeted and floxed LRRK1 and LRRK2 alleles

The generation and validation of the LRRK1 and LRRK2 targeting vectors, the targeted, floxed, and deleted LRRK1 and LRRK2 alleles by Southern, Northern, RT-PCR, and the sequences of the floxed LRRK1 and LRRK2 alleles are included in Figure 1-figure supplement 1-11.

LRRK1

To generate the LRRK1 targeting vector, we first PCR amplified the left middle homologous region (2,079 bp) containing partial intron 26, exon 27, and partial intron 27 of LRRK1 from mouse BAC DNA (clone RP23-213J23, BACPAC Resources Center) using primers, P3 and P4. The PCR fragment was subcloned into the pGEM-T vector (A1360, Promega) to generate pLM1 (for details, see Figure 1-figure supplement 1). The right middle homologous region (3,403) containing LRRK1 genomic region from partial intron 27 to partial intron 29 was amplified by PCR using primers, P5 and P6, and then subcloned into the pGEM-T vector (pLM2). The left middle homologous region containing pGEM-T plasmid was then digested with NotI and KpnI (endogenous KpnI site in the intron 27) and subcloned into the NotI and KpnI sites of the right middle homologous region containing pGEM-T plasmid (pLM3). After that, the middle homologous region (5.5 kb), from partial intron 26 to partial intron 29, was released by NotI and SacII digestions and was blunted by Klenow, and then subcloned into the SmaI site of PGKneoF2L2DTA vector (#13445, addgene) to generate the middle homologous region-PGKneoF2L2DTA plasmid (pLM4).

The left homologous region (2,016 bp) containing partial intron 25, exon 26, and partial intron 27 of LRRK1 was PCR amplified from mouse BAC DNA (clone RP23-213J23, BACPAC Resources Center) using primers, P1 and P2. The PCR fragment was digested with SacII (introduced by P1) and NotI (introduced by P2 along with HindIII), and was subcloned into the SacII and NotI sites of the PGKneoF2L2DTA vector (pLM5). The right homologous region (3,131 bp) containing partial intron 29, exon 30, and partial intron 30 of LRRK1 was PCR amplified from mouse BAC DNA (clone RP23-213J23) using primers, P7 and P8. The PCR fragment was digested with SalI (introduced by P7) and HindIII (introduced by P8), and was subcloned into the SalI and HindIII sites of the PGKneoF2L2DTA vector (pLM6).

The left homologous region in the PGKneoF2L2DAT vector (pLM5) was digested with SacII and NotI, and subcloned into the SacII and NotI sites of the right homologous region containing pGKneoF2L2DTA plasmid (pLM6) to generate pLM7. Finally, the loxP-middle homologous region-PGK-Neo-loxP fragment was released from pLM4 by NotI and SalI digestions, and subcloned into the NotI and SalI sites of pLM7 to generate the final target vector (pLM8), which contains two loxP sites (intron 26 and intron 29). Upon Cre-mediated recombination, the endogenous LRRK1 genomic sequence flanked by the 5’ loxP site (1,288 bp upstream of LRRK1 exon 27) to the 3’ loxP site (1,023 bp downstream of exon 29) is excised. The PGK-neo cassette flanked by two FRT sites is under the control of the mouse phosphoglycerate kinase 1 (PGK) promoter and contains the bovine growth hormone polyA signal. To enhance the ratio of ES cells carrying homologous recombination events instead of random insertion of the targeting vector (77), the negative selection PGK-DTA cassette, which encodes diphtheria toxin A chain, was also included in the LRRK1 targeting vector.

The LRRK1 targeting vector was linearized by XhoI digestion, and then electroporated into MKV6.5 embryonic stem (ES) cells (a gift from R. Jaenisch), which were derived from B6/129 F1 mice (The Jackson Laboratory, IMSR_JAX:101043). G418 was applied to the culture at 150 μg/ml 24 h later, and after 6 days of G418 selection, the surviving ES clones (480) were picked and screened by Southern analysis using HindIII digestion of genomic DNA followed by hybridization with the 5’ external and 3’ external probes to confirm proper recombination events in the 5’ and 3’ homologous regions, respectively. 25 ES cell clones were confirmed to carry the proper homologous recombination events at the 5’ homologous region, giving rise to the 17.0 kb and the 4.8 kb bands, which represent the wild-type and the targeted allele, respectively (Figure 1D), and the 3’ homologous region, giving rise to the 17.0 kb and the 14.1 kb bands, which represent the wild-type and the targeted allele, respectively. We then expanded the selected 4 ES cell clones (3A11, 3D8, 3H1, and 3H6) and further verified by Southern using the 5’ and 3’ external, and neo probes. Two ES cell clones (3D8 and 3H6) were transfected with pCAGGS-flpE-puro (#20733, addgene) to delete the PGK-neo cassette by FLP recombinase. Three resulting ES cell clones (3D8C5, 3D8E5, and 3H6G7) were confirmed by Southern analysis using the 5’ and 3’ external probes and the neo probe to confirm the floxed LRRK1 allele by the deletion of the PGK-neo cassette.

LRRK2

To generate the LRRK2 targeting vector, the left homologous region (2,579 bp) containing the LRRK2 promoter region was PCR amplified from mouse BAC DNA (clone RP23-526A2, BACPAC Resources Center) using primers, P9 and P10, and subcloned into the pGEM-T vector to generate pLRRK2#1 (for details see Figure 1-figure supplement 4). The BamHI-loxP-NheI-SpeI fragment (56 bp), which was generated by annealing two complementary oligos, P39 and P40, was introduced to the BamHI and SpeI sites of pLRRK2#1 to generate pLRRK2#2. The middle homologous region, containing genomic sequences from the promoter region to partial intron 2, was PCR amplified from mouse BAC DNA (clone RP23-526A2) using primers, P11 and P12, digested with XbaI and NotI, and subcloned into the XbaI and NotI sites of the pGEM-T vector to generate pLRRK2#3, which was digested with XbaI and NotI, and subcloned into the XbaI and NotI sites of pLRRK2#2 to generate pLRRK2#4.

The right homologous region (3,503 bp), which contains partial intron 1, exon 2, and partial intron 2, was PCR amplified from mouse BAC DNA (clone RP23-526A2) using primers, P13 and P14, and the PCR fragment was subcloned into the EcoRV site of pBluescript II KS (+) vector (212207, Agilent) to generate pLRRK2#5. The right homologous region in pLRRK2#5 was digested with BamHI followed by Klenow, and then digested with ClaI. The released fragment was subcloned into the EcoRV and ClaI sites of the PGKneoF2L2DTA vector to generate pLRRK2#6. LFNT-tk/pBS plasmid (a gift from S. Tonegawa) was digested with SacII (followed by Klenow) and NotI (followed by SspI digestion to make it easier to isolate the NotI-SacII/KN fragment) to release the loxP-FRT-PKG-neo-loxP-FRT fragment (2,928 bp), and then subcloned into the XbaI (followed by Klenow) and NotI sites of pLRRK2#6 to generate pLRRK2#7. Finally, pLRRK2#4 containing the left-loxP-middle homologous regions was digested with EcoRV and NotI, and subcloned into AleI and NotI sites of pLRRK2#7 containing the loxP-FRT-PKG-neo-loxP-FRT-right homologous region to generate the final targeting vector, which contains the loxP sites in the promoter region (1,768 bp upstream of the transcription initiation site) and in LRRK2 intron 2 (878 bp downstream of exon 2). Upon Cre-mediated recombination, the floxed endogenous LRRK2 genomic sequences from the 5’ loxP site (1,768 bp upstream of LRRK2 transcription) to the 3’ loxP site in intron 2 (878 bp downstream of exon 2) are excised. The negative selection PGK-DTA cassette is also included in the LRRK2 targeting vector.

The LRRK2 targeting vector was linearized by AhdI digestion, and then electroporated into MKV6.5 ES cells. G418 was applied to the culture at 150 ug/ml 24 h later, and after 6 days of G418 selection, the surviving ES clones (480) were picked and screened by Southern analysis using NheI digestion of genomic DNA followed by hybridization with the 5’ external probe (753 bp, PCR amplified using P19 and P20). Southern analysis confirmed that two ES cell clones (N24 and N34) carry the proper recombination event in the 5’ homologous arm, giving rise to 11.6 kb and 3.6 kb bands, which represent the wild-type and targeted alleles, respectively (Figure 1F), followed by genomic PCR confirmation for the proper recombination in the right arm.

Three (3D8C5, 3A11, 3H6) and one (N24) ES clones for LRRK1 and LRRK2, respectively, were microinjected into C57BL/6 mouse blastocysts to generate chimera mice, which were bred with B6/129 F1 mice to produce heterozygous floxed LRRK1 and targeted LRRK2 mice. Floxed LRRK1 mice were confirmed by Southern analysis using the 5’ and 3’ external probes (Figure 1D). Targeted LRRK2 mice were confirmed by Southern (data shown in Figure 1-figure supplement 6; NheI digestion followed by hybridization with the 5’ external probe and SphI digestion followed by hybridization with the 3’ external probe). Targeted LRRK2 mice were then bred with Actin-FLP deleter mice (IMSR_JAX:005703) (33) to generate floxed LRRK2 mice, which were confirmed by Southern analysis using the 5’ and 3’ external probes following NheI digestion (Figure 1F). Heterozygous fLRRK1/+ and fLRRK2/+ mice were crossed with each other to obtain homozygous single floxed mice (fLRRK1/fLRRK1 and fLRRK2/fLRRK2) and double floxed mice (fLRRK1/fLRRK1; fLRRK2/fLRRK2).

Generation of deleted LRRK1/LRRK2 alleles and DA neuron-specific cDKO mice

In order to ensure that Cre-mediated deletion of the floxed LRRK1 and LRRK2 alleles results in null alleles, we crossed floxed LRRK1 and LRRK2 mice with germline deleter, CMV-Cre transgenic mice (B6.C-Tg(CMV-cre)1Cgn/J; IMSR_JAX:006054) (34), to generate LRRK1 deleted (11/11) mice (by removing exons 27-29) and LRRK2 deleted (11/11) mice (by removing the promoter region and exons 1-2) for further molecular characterization. To generate DA neuron-specific LRRK1/LRRK2 cDKO mice, we used DAT-Cre KI mice (B6.SJL-Slc6a3tm1.1(cre)Bkmn/J; IMSR_JAX:006660), which express Cre recombinase under the control of the endogenous DAT promoter (32). We crossed double floxed mice with DAT-Cre mice to generate LRRK cDKO mice (fLRRK1/fLRRK1; fLRRK2/fLRRK2; DAT-Cre/+). LRRK cDKO and littermate control mice used in the phenotypic analysis were obtained by crossing fLRRK1/fLRRK1; fLRRK2/fLRRK2; DAT-Cre/+ with fLRRK1/fLRRK1; fLRRK2/fLRRK2 mice. We only used cDKO and control mice that carry all floxed LRRK1 and LRRK2 alleles for phenotypic analysis.

Southern analysis

For the identification and validation of the targeted and floxed LRRK1 alleles, we used the 5’ external, 3’ external, and neo probes. Genomic DNA from ES cells or mouse tails was digested with HindIII. The 5’ external probe (377 bp), which is 839 bp upstream of the 5’ homologous region, was PCR amplified from mouse BAC DNA (clone RP23-213J23) using primers P15 and P16. The 3’ external probe (305 bp), which is 2,736 bp downstream of the 3’ homologous region, was PCR amplified from mouse BAC DNA (clone RP23-213J23) using primers P17 and P18. The neo probe (363 bp) was PCR amplified from pSoriano plasmid using primers P23 and P24. Following HindIII digestion, the presence of the 17.0 kb product using either the 5’ or 3’ external probe represents the wild-type allele, whereas the 4.8 kb (the 5’ external probe) and the 12.2 kb (the 3’ external probe) products represent the floxed LRRK1 allele. Genomic DNA digested by HindIII and hybridized with the neo probe further confirmed the wild-type and floxed alleles (no band) and the targeted LRRK1 allele (14.1 kb).

For the identification and validation of the targeted and floxed LRRK2 alleles, we used the 5’ and 3’ external probes as well as the neo probe. Genomic DNA from ES cells or mouse tails was digested with NheI or SphI followed by hybridization with the 5’ or 3’ external probe or the neo probe. The 5’ external probe (753 bp), which is 84 bp upstream of the 5’ homologous region, was PCR amplified from mouse BAC DNA (clone RP23-526A2) using primers P19 and P20. The 3’ external probe (622 bp), which is 38 bp downstream of the 3’ homologous region, was PCR amplified from mouse BAC DNA (clone RP23-526A2) using primers P21 and P22. Following NheI digestion, the presence of the 11.5 kb product using either the 5’ or 3’ external probe represents the wild-type allele, whereas the 3.6 kb (the 5’ external probe) and the 5.2 kb (the 3’ external probe) products represent the floxed LRRK2 allele.

PCR genotyping

Genomic PCR was performed to determine the presence of the deleted, the floxed, and/or the wild-type alleles. For LRRK1, the following primers were used: P25 (5’-ATTGGTCTTTGAAGAGACAGCATCTGG, forward primer, 392 nt downstream of exon 26), P26 (5’-TTTCCCTGAGGTGGAGAAGTGACTGG, reverse primer, 567 nt downstream of exon 26), and P27 (5’-TCACGTCGTCTAAGCCTCCT, reverse primer, 1,218 nt downstream of exon 29). The PCR products from P25 and P26 are 266 bp and 405 bp, which represent the wild-type and the fLRRK1 alleles, respectively, whereas the PCR product from P25 and P27 is 583 bp, which represents the deleted LRRK1 allele.

For LRRK2, the following primers were used: P28 (5’-CTTCCTCAGAAGTTAGGTAAACATTGAGTG, forward primer, 2,069 nt upstream of exon 1), P29 (5’-CTAAGTGACACCGTGTTTCCAAAGTC, reverse primer, 1,739 nt upstream of exon 1), and P30 (5’-GGAAAGTTTCACAATTGGAAAAATAAAAATATTTACTGCAGATA, reverse primer at 2,848 nt downstream of exon 2). The PCR products from P28 and P29 are 305 bp and 367 bp, representing the wild-type and the fLRRK2 alleles, respectively, whereas the PCR product from P28 and P30 is 587 bp, which represents the deleted LRRK2 allele.

For DAT-IRES-Cre, the following primers were used: JKM1823 (5’-TGGCTGTTGGTGTAAAGTGG, forward primer at exon 16 and 3’-UTR), JKM1824 (5’-GGACAGGGACATGGTTGACT, reverse primer at 3’-UTR), and JKM1825 (5’-CCAAAAGACGGCAATATGGT, reverse primer at IRES sequence). The PCR product from JKM1823 and JKM1824 is 264 bp, which represents the wild-type allele, whereas the PCR product from JKM1823 and JKM1825 is 152 bp, which represents the DAT-IRES-Cre KI allele.

Northern analysis

Total RNA was isolated from brains, kidneys, or lungs using TRI reagent (T9424, Millipore-Sigma) according to the manufacturer’s instruction. For the LRRK1 Northern analysis, polyA+ RNA was enriched from ∼500 μg total RNA using the Poly(A)Purist™ MAG Kit (AM1922, ThermoFisher) according to the manufacturer’s instruction. For the LRRK2, ∼10 μg of total RNA was used for Northern analysis. RNA was separated in formaldehyde agarose gel, and transferred into Amersham Hybond-nylon membrane (RPN303N, GE Healthcare). Probes were synthesized using Prime-It II random labeling kit (#300385, Agilent) and then used for membrane hybridization at 55°C overnight.

The cDNA probe specific for LRRK1 exons 2-3 (383 bp) was PCR amplified using primers P31 (5’-CAGGATGAGCGTGTGTCTGCAG) and P32 (5’-CCTTCTCCTGTGAGGATTCGCTCT). The cDNA probe specific for LRRK1 exons 27-29 (550 bp) was PCR amplified using primers P33 (5’-CTGGCCTACCTGCACAAGAA) and P34 (5’-CCTTCCCATCCCAGAACACC).

The cDNA probe specific for LRRK2 exons 1-5 (437 bp) was PCR amplified using primers P35 (5’-AGGAAGGCAAGCAGATCGAG) and P36 (5’-GGCTGAATATCTGTGCATGGC). The probe specific for GAPDH exons 5-7 (452 bp) was PCR amplified using primers P37 (5’-ACCACAGTCCATGCCATCAC) and P38 (5’-TCCACCACCCTGTTGCTGTA).

Hybridization was performed using α-32P-dCTP-labeled probes specific to each gene. Specific signals were detected by autoradiography with Hyperfilm (E3018, Amersham).

RT-PCR

Total RNA was isolated from brains, kidneys, or lungs using TRI reagent (T9424, Millipore-Sigma,) according to the manufacturer’s instruction. Approximately 1 μg of RNA was reverse-transcribed using Superscript III (18080093, Thermo Fisher Scientific) according to the manufacturer’s instructions. For RT-PCR analysis of LRRK1 transcripts in mice carrying the homozygous floxed or deleted alleles, we used primers P53 and P54 for exons 4-8 (714 bp), P57 and P58 for exons 11-17 (818 bp), P63 and P64 for exons 20-25 (922 bp), or P41 and P42 for exons 25-31 to confirm normal splicing of LRRK1 mRNA in fLRRK1/fLRRK1 mice and truncated LRRK1 transcripts lacking exons 27-29 in 11/11 mice (for details see Figure 1-figure supplement 9). For RT-PCR analysis of LRRK2 transcripts, we used primers P35 and P36 in exons 1-5 to confirm normal splicing of LRRK2 mRNA (437 bp) in fLRRK2/fLRRK2 mice and the absence of RT-PCR products in LRRK2 11/11 mice (Figure 1-figure supplement 11). The identity of the PCR products was confirmed by sequencing.

Western analysis

Fresh tissues were collected and homogenized in an ice-cold stringent RIPA buffer (50 mM Tris-Cl (pH 7.6), 150 mM NaCl, 0.5 mM EDTA, 1% NP40, 0.5% sodium deoxycholate, 0.1% SDS, 1mM PMSF supplement with protease inhibitor cocktail (P8340, Sigma) and phosphatase inhibitor cocktail (P0044, Sigma)), followed by sonication. Homogenates were centrifuged at 14,000xg for 20 min at 4°C to separate supernatants (RIPA buffer-soluble fraction). An equal amount (10-40 μg per lane) of total proteins from each preparation were loaded and separated on NuPAGE gels (Invitrogen), and transferred to nitrocellulose membranes. The membranes were blocked in Intercept (TBS) Blocking Buffer (927-60001, Li-Cor) for 1 hr at room temperature, and incubated at 4°C overnight with specific primary antibodies. Primary antibodies used were rabbit anti-LRRK1 (ANR-101, Alomone Lab, RRID: AB_2756700), rabbit anti-LRRK2 (ab133474, abcam, RRID: AB_2713963), mouse anti-α-Vinculin (05-386, Millipore, RRID: AB_309711). Membranes were then incubated with dye-coupled secondary antibodies, goat anti-mouse IRdye680 (#925-68070, LI-COR, RRID: AB_2651128) or goat anti-rabbit IRdye800 (#925-32211, LI-COR, RRID: AB_2651127). Signals were quantified using the Odyssey Infrared Imaging System (LI-COR).

Histological analysis

Mice were anesthetized with ketamine (100 mg/kg) + xylazine (10 mg/kg) + acepromazine (3 mg/kg), and transcardially perfused with phosphate-buffered saline solution (PBS, pH 7.4) containing 0.25 g/L heparin (H3149, Sigma) and 5 g/L procaine (P9879, Sigma). Brains were post-fixed in 4% formaldehyde in PBS (pH 7.4) (15710, Electron Microscopy Sciences) at 4°C overnight and then processed for paraffin embedding following standard procedures. For frozen sections, post-fixed brains were immersed in a sucrose series solution (15% and 30% sucrose in PBS) at 4°C overnight for cryoprotection, and then brains were embedded in an OCT compound (4583, Sakura). Serial coronal sections (16 µm) of paraffinized brains or frozen brains were obtained using Leica RM2235 or Leica CM1860, respectively. Coronal sections containing the SNpc or LC were selected for immunohistochemical analysis.

Histological analyses were performed as described previously (30, 31, 76). Briefly, for DAB-derived TH-immunohistochemistry, coronal sections were deparaffinized, alcohol-dehydrated, and then subjected to permeabilization with a solution containing 0.1% Triton X-100 in TBS followed by antigen retrieval for 5 min in 10 mM sodium citrate buffer, pH 6.0. Endogenous peroxidase activity was quenched by incubating in 0.3% H2O2 in methanol. Sections were then blocked with a solution containing 5% normal goat serum (S-1000, Vector Laboratories) and 0.1% Triton X-100 in TBS for 1 hr at room temperature. After blocking, sections were incubated with the primary antibody, rabbit anti-TH (1:750, ab112, abcam, RRID: AB_297840) overnight at 4°C. Sections were washed three times in 0.1% Triton X-100 in TBS followed by 1 hr incubation with goat biotinylated anti-rabbit IgG secondary antibody (1:250, BA-1000, Vector Laboratories, RRID: AB_2313606) at room temperature and 30 min incubation with Vectastain Elite ABC reagent (PK-6100, Vector Laboratories) and then developed using chromogenic DAB substrate (SK-4100, Vector Laboratories).

For immunofluorescence staining of paraffin sections, coronal sections were deparaffinized, alcohol-dehydrated, and then subjected to permeabilization with a solution containing 0.1% Triton X-100 in PBS followed by antigen retrieval for 5 min in 10 mM sodium citrate buffer, pH 6.0 except those for cleaved-caspase3 immunostaining which were performed antigen retrieval for 10 min. Sections were then blocked with a solution containing 10% normal goat serum and 0.1% Triton X-100 in PBS for 1 hr at room temperature. After blocking, sections were incubated with primary antibodies overnight at 4°C. The primary antibodies used were mouse anti-TH (1:50, sc-25269, Santa Cruz, RRID: AB_628422), rabbit anti-NeuN (1:400, 12943S, Cell Signaling Technology, RRID: AB_2630395), rabbit anti-cleaved caspases-3 (1:150, 9661, Cell Signaling Technology, RRID: AB_2341188), or rabbit anti-Iba1 (1:500, 019-19,751, Wako, RRID: AB_839504). Sections were washed three times in 0.1% Triton X-100 in PBS followed by 1 hr incubation with fluorophore-conjugated secondary antibodies, goat anti-mouse IgG Alexa Fluor 488 (1:500, A11001, ThermoFisher, RRID: AB_2534069) and goat anti-rabbit IgG Alexa Fluor 555 (1:500, A32732, ThermoFisher, RRID: AB_2633281) at room temperature. Fluorescence images were taken (a stack of 2-3 confocal images spaced at 4 μm), projected with maximal intensity projection mode, and analyzed by FV1000 confocal microscope system (Olympus).

For immunofluorescence staining of cryopreserved sections, coronal brain sections were washed with PBS to rinse out OCT, and then blocked with a solution containing 5% normal goat serum and 0.1% Triton X-100 in TBS for 1 hr at room temperature. After blocking with 10% NGS, sections were incubated with primary antibodies, rabbit anti-GFP (1: 1000, ab290, abcam, RRID: AB_303395), and mouse anti-TH (1:50, sc-25269, Santa Cruz, RRID: AB_628422), overnight at 4°C. Sections were incubated with fluorophore-conjugated secondary antibodies, goat anti-rabbit IgG Alexa Fluor 488 (1:500, A11034, ThermoFisher, RRID: AB_2576217) and goat anti-mouse IgG Alexa Fluor 555 (1:250, A21424, ThermoFisher, RRID: AB_141780) for 1 hr at room temperature. Fluorescence images were taken and analyzed by FV1000 confocal microscope system (Olympus).

The number of GFP+ and GFP+/TH+ cells in the SNpc of DAT-IRES-Cre/+; Rosa26-CAG-LSL-ZsGreen1/+ reporter mice were quantified using 3 comparable coronal sections (16 µm in thickness, spaced 320 μm apart) per brain (n = 3 brains, one hemisphere). The percentage of GFP+ DA neurons in the SNpc was obtained by dividing the sum of GPF+/TH+ neurons by the sum of total TH+ neurons quantified.

Quantification of DA neurons in the SNpc

Quantification of TH+ DA neurons in the SNpc or LC was performed as previously described (30, 31, 76, 78). Briefly, TH+ neurons in the SNpc, which were marked based on morphological features as previously described (79), were quantified in every 10th serial coronal section (16 µm in thickness) throughout the SNpc (a total of 6-9 sections, spaced 160 µm apart). Total number of TH+ cells in the SNpc was calculated as follows: [total number of TH+ DA neurons quantified in all 6-9 sections] x 10 (every 10th section sampled) x 2 (both hemispheres). Total number of TH+ noradrenergic neurons in the LC (a total of 6-9 sections, spaced 80 µm apart) was calculated as follows: [total number of TH+ noradrenergic neurons quantified in all 6-9 sections] x 5 (every 5th section sampled) x 2 (both hemispheres).

DA neuron quantification was also performed independently by another investigator, also in a genotype blind manner, using stereological quantification of 25% of the SNpc area. The fractionator with 100 μm x 100 μm was set up in the SNpc, and TH+ DA neurons were counted using the optical dissector method with 50 μm x 50 μm sample box (25% of the total area). Total number of TH+ cells in the SNpc was calculated as follows: [total number of TH+ DA neurons quantified in sample boxes of all 6-9 sections] x 10 (every 10th section sampled) x 4 (1/4 area sampled: 50 x 50/100 x 100) x 2 (both hemispheres).

Quantification of NeuN+ neurons, apoptotic cells, Iba1+ microglia in the SNpc

NeuN+, active Caspase-3+, or Iba1+ cells in the SNpc, which was marked by TH immunoreactivity, were quantified using serial coronal sections (16 μm in thickness, every 10th section, a total of 6-9 sections per brain). Compared to DAB immunostaining followed by counting under the stereomicroscope, which captures TH-positive DA neurons in a single cell layer of the brain section, immunofluorescent staining picks up more cells (e.g. TH+ or NeuN+) in multiple layers of the brain section under the confocal microscope by the maximal intensity projection mode. Thus, more TH-positive DA neurons in the SNpc were captured and quantified under confocal microscopy compared to DAB staining. The total number of NeuN+ cells in the SNpc was calculated by multiplying the total number of NeuN+ cells in all sections counted] x 10 (every 10th section sampled) x 2 (both hemispheres). The total number of active Caspase-3+ or Iba1+ cells in the SNpc, which was marked by TH immunoreactivity, was calculated as follows: [total number of active Caspase-3+ or Iba1+ cells in all sections counted from one hemisphere] x 10 (every 10th section sampled).

Quantification of TH+ DA terminals in the striatum

For quantification of TH immunoreactivity in the striatum, we performed immunostaining using every 10th serial coronal sections (16 µm in thickness) throughout the entire striatum (a total of 12-15 sections, spaced 160 µm apart). The images of TH immunoreactivity in the striatum were captured under 2X objective lens (Olympus BX40, 8-bit RGB camera) using identical exposure time, sensitivity, brightness, contrast, and gamma (CellSens Entry Software) and then analyzed using the Fiji version of ImageJ, and the optical density was determined as previously described (31, 80, 81). The mean value of TH immunoreactivity in the striatum of control mice of each age group was set as 100%.

Quantitative EM analysis

The collection and quantification of the EM images were performed as described previously (30, 31). Mice were perfused with PBS containing 0.25 g/L heparin and 5 g/L procaine followed by a fixative solution containing 2.5% paraformaldehyde and 2.5% glutaraldehyde in 0.1 M sodium cacodylate buffer (pH 7.4) (#1549, Electron Microscopy Science). Brains were dissected and post-fixed overnight in a fixative solution at 4°C. The dissected tissues were trimmed to 1-2 mm3 cubes followed by osmication and uranyl acetate staining, dehydration in graded alcohol, and embedded in TABB 812 Resin (Marivac Ltd.) at the Harvard Medical School EM facility. 0.5 μm sections were stained with toluidine blue and viewed under the light stereomicroscope (Nikon Eclips E600) to find the SNpc area for EM viewing. Then adjacent sections were cut with 80 nm in thickness with the Leica Ultracut S microtome, picked up on formvar-carbon coated slot Copper grids, stained with 0.2% Lead Citrate, and viewed and imaged under the JEOL 1200x Electron Microscope. A minimum of 10 micrographs containing the entire cell body in the SNpc area were analyzed for each brain. The image was analyzed by using the Fiji version of ImageJ. The number of electron-dense autophagic and lysosomal vacuoles (>0.5 μm in diameter) in individual neuronal profiles was quantified. We previously calculated the diameter of electron-dense autophagic/lysosomal vacuoles by measuring the longest side manually (30, 31). In the current study, we used Feret’s Diameter (82) to accurately measure the longest distance between any of two points in the electron-dense autophagic/lysosomal vacuoles, resulting in a higher number of electron-dense vacuoles quantified. The number and the area of electron-dense vacuoles (>0.5 μm in diameter) in individual neuronal profiles was quantified, and the average number or area of vacuoles per mouse was calculated. Experiments were done in a genotype blind manner (after scarifying mice, the brain samples were coded and sent to the Harvard Medical School EM Core, where the images were captured, and the vacuoles were quantified by another independent experimenter).

Behavioral analysis

Naive LRRK cDKO mice and littermate controls at 10 and 22 months of age were used. Mice were acclimated in the behavior facility for a minimum of 7 days and were then individually handled daily for 5 consecutive days before testing. Mice were coded so the experimenter was unaware of their genotypes until the data analysis was complete. In the beam walk test, a Plexiglas beam of 100 cm in length (Plastic Zone), 20 mm or 10 mm in width, was raised 60 cm above a table, and safety bedding was placed under the beam to avoid any harm in case of falls. Mice were placed onto the starting point of the beam in bright light, and the time (in seconds) to reach their home cage on the other darker side of the beam (∼80 cm in distance) as well as the hindpaw slips (number of hindlimb errors) were recorded. Mice were trained three trials per day for 2 consecutive days to traverse the 20 mm beam (without the wire mesh) to their home cage. On the test day, mice were trained further with two additional trials on the 20 mm beam (without the wire mesh). Mice were then tested in two successive trials on the 20 mm beam (with the wire mesh) followed by two consecutive test trials on the 10 mm beam (with the wire mesh). All test trials were videotaped, and the travel time and the number of hindlimb errors were recorded. Between trials mice were placed in the home cage for 2 min to rest. Mice that fell off the beam during both trials or stalled on the beam for more than 120 sec during the test were excluded.

In the pole test, mice were placed on top of the pole (60 cm in height, 10 mm in diameter) with their head facing upward, and the base of the pole was placed in the home cage. Mice were trained three trials per day for 2 days to traverse the pole to the cage floor and were further trained two trials before testing on the test day. Mice were then tested for 2 trials, and the time to turn around (turning time) and the time to descend the pole (descending time) were recorded. Between trials, mice were placed in the home cage for 2 min to rest. Mice that stalled on the top of the pole for more than 120 sec were excluded.

Experimental design and statistical analysis

All experiments were performed in a genotype blind manner with the exception of molecular analysis (Southern, Northern, RT-PCR, and Western). All statistical analyses were performed using Prism 9 (GraphPad software) or Excel (Microsoft). All data are presented as the means ± SEM. The exact sample size of each experiment is indicated in the figure or the legend. The slight difference in sample size among various histological analysis (e.g. at 24M, 11 control and 8 cDKO brains analyzed for TH+ cells but 9 control and 8 cDKO analyzed for NeuN+, Caspase-3+, Iba1+ cells in the SNpc) is due to the paraffin blocks used for quantification (each paraffin block contains 7 brains of mixed genotypes; for details see Source Data files) or specific brain sections being damaged, thus, excluded in the analysis.

Statistical analyses were conducted using an unpaired two-tailed Student’s t-test for the comparison of a given variable in two genotypes or two-way ANOVA followed by Bonferroni’s post hoc comparisons for the comparison of more than two conditions. Statistical outliers were identified and excluded using the ROUT method with 1% the maximum desired false discovery rate developed by Prism, and the only statistical outliers of the current study were identified in the behavioral analysis and were marked in Figure 9 - Source Data. All statistical analysis was performed on data from ≥ 4 mouse brains per genotype per age group, and experiments were performed and repeated on different days, often by independent investigators. Statistical significance is indicated as *p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001, and values not significantly different are often not noted.

Supporting information